---
title: "Newton–Raphson Method — Translating Math into Code"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

# Mathematical Intuition

We want to find a root of $f(x)=0$.\
We use the tangent line at $x_n$ to estimate a better guess:

$x_{n+1} = x_n - \frac{f(x_n)}{f'(x_n)}$. 

Repeat until ( $|x_{n+1} - x_n|$ ) is small.

::: {.callout-note title="In a nutshell: why the derivative?"}
The tangent line at our current guess points towards the root and its slope (the derivative) tells us how big a step to take to get there — a steep slope means a small, careful step, a shallow slope means a larger leap.
:::

```{r}
#| echo: false
#| warning: false
#| message: false

# Define function and derivative
f  <- function(x) x^3 - 2*x - 5
df <- function(x) 3*x^2 - 2

# Newton–Raphson iterations
newton <- function(x0, n) {
  x <- numeric(n + 1)
  x[1] <- x0
  for (i in 1:n) {
    x[i + 1] <- x[i] - f(x[i]) / df(x[i])
  }
  x
}

# Initial guess and iterations
x0 <- 2
iters <- newton(x0, 5)

# Plot function
curve(f, from = 1, to = 3, lwd = 2,
      ylab = "f(x)", xlab = "x",
      main = "Newton–Raphson Method")

abline(h = 0, col = "gray")

# Plot Newton steps
for (i in 1:5) {
  xN <- iters[i]
  yN <- f(xN)
  slope <- df(xN)

  # Tangent line
  abline(a = yN - slope * xN, b = slope,
         col = "steelblue", lty = 2)

  # Vertical drop to x-axis
  segments(xN, yN, iters[i + 1], 0,
           col = "firebrick")

  # Points
  points(xN, yN, pch = 16)
}

# Final root estimate
points(iters[6], 0, pch = 19, col = "firebrick")

```

------------------------------------------------------------------------

# Math → Algorithm → Code Mapping

| Math concept | R translation             |
|--------------|---------------------------|
| update rule  | `xNew <- x - f(x)/df(x)`  |
| derivative   | function `df(x)`          |
| convergence  | `abs(xNew - x) < tol`     |

------------------------------------------------------------------------

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart TD
  Start([Start: choose x0, tol, max_iter])
  Compute["x_next with the algorithm"]
  Check{Is the absolute difference between <br/> the next guess x_next and <br/> the current guess x less than the tolerance}
  Accept([Return x_next])
  Update["x <- x_next"]

  Start --> Compute --> Check
  Check -- Yes --> Accept
  Check -- No --> Update --> Compute
```

------------------------------------------------------------------------

# Implementation in R

```{r}
newton <- function(f, df, x0, tol = 1e-10, maxIter = 100) {
  x <- x0
  for (i in 1:maxIter) {
    fx <- f(x)
    dfx <- df(x)
    if (dfx == 0) stop("Derivative is zero; choose a new x0")
    xNew <- x - fx / dfx
    if (abs(xNew - x) < tol) return(xNew)
    x <- xNew
  }
  stop("Did not converge")
}

# Examples
f1 <- function(x) x^2 - 2
df1 <- function(x) 2*x
newton(f1, df1, x0 = 1)
```

------------------------------------------------------------------------

# Reflection

The Newton–Raphson method shows how an **iterative analytic idea** (tangent approximation) translates into a **computational fixed-point loop**.

::: {.callout-tip title="Going further 🚀"}
Writing your own solver is great for understanding, but R already has robust, well-tested tools. `uniroot()` finds a root within a bracketing interval, and `optim()` minimises (or, by negating, maximises) functions of one or more variables.

```{r}
###| Built-in root finder over an interval
uniroot(function(x) x^3 - 2*x - 5, interval = c(1, 3))$root
```

You also do not have to supply the derivative by hand. Base R's `D()` differentiates simple expressions **symbolically**, while the `numDeriv` package estimates gradients **numerically** for any function:

```{r}
###| Symbolic derivative with base D()
D(expression(x^3 - 2*x - 5), "x")
```

```{r}
#| eval: false
###| Numerical gradient (numDeriv package)
library(numDeriv)
grad(function(x) x^3 - 2*x - 5, x = 2)
```
:::