---
title: "Euclidean Algorithm — Translating Math into Code"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

# Mathematical Intuition

Recall from Sets and Numbers the Euclidean algorithm where we want to find the greatest common divisor (GCD) of two integers $(a, b)$.

::: {.callout-note title="In a nutshell: what is a GCD?"}
The greatest common divisor is simply the **largest whole number that divides both** of your numbers exactly (with no remainder). For example, $\gcd(12, 8) = 4$, because $4$ divides both $12$ and $8$, and nothing bigger does. Keep that plain picture in mind as you read the formal definition below.
:::

The GCD is defined as: $$
\gcd(a,b) =
\begin{cases}
|a|, & \text{if } b = 0, \\
\gcd(b, a \bmod b), & \text{otherwise.}
\end{cases}
$$

The algorithmic idea: repeatedly replace the pair $(a, b)$ with $(b, a \mod b)$ until the second number becomes $0$. Let us now look at a simplified flowchart of what we want our computer program to do.

**Question for the reader: try to draw the flowchart on your own!**

------------------------------------------------------------------------

# Math → Algorithm → Code Mapping

| Math operation    | R equivalent     | Meaning     |
|-------------------|------------------|-------------|
| $(a \mod b)$      | `a %% b`         | remainder   |
| recursive call    | `gcd(b, a %% b)` | repeat step |
| stop if ($b = 0$) | `if (b == 0)`    | base case   |

------------------------------------------------------------------------

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart TD
  A([Start: a, b])
  B{Is b equal to 0?}
  C[Set a to b<br/>Set b to a mod b]
  D([Return absolute value of a])

  A --> B
  B -- No --> C --> B
  B -- Yes --> D

```

------------------------------------------------------------------------

# Implementation in R

```{r}
# Define the GCD function
gcdIterative <- function(a, b) {
  a <- abs(as.integer(a))
  b <- abs(as.integer(b))
  while (b != 0) {
    r <- a %% b
    a <- b
    b <- r
  }
  return(a)
}

# Set up the recursive part of the function
gcdRecursive <- function(a, b) {
  a <- abs(as.integer(a))
  b <- abs(as.integer(b))
  if (b == 0) return(a)
  gcdRecursive(b, a %% b)
}
```

------------------------------------------------------------------------

# Example usage

```{r}
gcdIterative(384, 24)
```

```{r}
gcdIterative(1071, 462)
```

In this example, we can see how we can **break down** algebraic reasoning (these are the skills proofs of theorems and lemmas give you!) into their **basic building blocks** and then express them as a **loop** or **recursive function** as you would do on paper! We have successfully turned a symbolic rule into a computational process! Now it is your turn, try to implement any algorithm of your choosing and see if you can get it to work!

::: {.callout-tip title="Going further 🚀"}
The **extended** Euclidean algorithm returns not just the GCD but also integers $s, t$ (the *Bézout coefficients*) satisfying $\gcd(a, b) = s \cdot a + t \cdot b$ — useful for modular inverses in cryptography. We can also extend the GCD to a whole vector by **folding** the two-argument function across it with `Reduce()`.

```{r}
###| Extended Euclidean algorithm: returns gcd and Bezout coefficients s, t
extendedGcd <- function(a, b) {
  if (b == 0) return(c(g = a, s = 1, t = 0))
  prev <- extendedGcd(b, a %% b)
  c(g = prev["g"], s = prev["t"], t = prev["s"] - (a %/% b) * prev["t"])
}
extendedGcd(240, 46)

###| GCD of a whole vector by folding the pairwise function
Reduce(gcdIterative, c(48, 36, 60, 24))
```

In practice you rarely need to write your own: the `numbers` package provides a ready-made `numbers::GCD()`, so reach for that in production code.
:::