---
title: "Monte Carlo Approximation of π — Translating Math into Code"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
editor: visual
---

# Mathematical Intuition

We throw random points $(x,y)$ uniformly in the square $[-1,1]^2$.\
The fraction inside the unit circle $x^2 + y^2 \le 1$ equals $\pi/4$.

*Why $\pi/4$?* The square $[-1,1]^2$ has side $2$, so its area is $2 \times 2 = 4$. The unit circle has radius $1$, so its area is $\pi r^2 = \pi$. A uniformly scattered point lands in the circle with probability equal to the ratio of those areas, $\pi/4$. Rearranging gives our estimator:

$$
\frac{\text{points inside circle}}{\text{total points}} \approx \frac{\pi}{4} \quad\Longrightarrow\quad \pi \approx 4 \times \frac{\text{points inside circle}}{\text{total points}} .
$$

------------------------------------------------------------------------

# Math → Algorithm → Code Mapping

| Math idea         | R translation      |
|-------------------|--------------------|
| sample $(x, y)$   | `runif(N, -1, 1)`  |
| indicator $(I_i)$ | `(x^2 + y^2 <= 1)` |
| estimate          | `4 * mean(I)`      |

------------------------------------------------------------------------

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart TD
  A([Start: choose N])
  B[Sample N points in the unit circle]
  C{Inside unit circle?}
  D[Count proportion inside]
  E[pi_hat = 4 x proportion]
  F([Return pi_hat])

  A --> B --> C --> D --> E --> F
```

------------------------------------------------------------------------

# Implementation in R

```{r}
set.seed(0)

estimatePi <- function(N) {
  x <- runif(N, -1, 1)
  y <- runif(N, -1, 1)
  inCirc <- (x^2 + y^2) <= 1
  4 * mean(inCirc)
}

estimatePi(100)

estimatePi(10000)

estimatePi(1000000)
```

------------------------------------------------------------------------

# Visualisation

```{r}
set.seed(1)
N <- 10000
x <- runif(N, -1, 1)
y <- runif(N, -1, 1)
inCirc <- (x^2 + y^2) <= 1

plot(x[inCirc], y[inCirc], col='skyblue', pch=16, cex=0.5,
     xlab='x', ylab='y', asp=1,
     main='Monte Carlo π Approximation',
     xlim=c(-1,1), ylim=c(-1,1))
points(x[!inCirc], y[!inCirc], col='gray', pch=16, cex=0.5)
symbols(0, 0, circles=1, inches=FALSE, add=TRUE, lwd=2)
```

------------------------------------------------------------------------

::: {.callout-tip title="Going further 🚀"}
How fast does the estimate actually converge? We can *watch* it by plotting the running estimate as each new point arrives — reusing the `inCirc` vector from the plot above:

```{r}
#| code-fold: true
running <- cumsum(inCirc) / seq_len(N)        # running fraction inside the circle
plot(4 * running, type = "l", col = "steelblue", log = "x",
     xlab = "number of points (log scale)", ylab = expression(hat(pi)),
     main = "The estimate settles down as N grows")
abline(h = pi, col = "firebrick", lwd = 2, lty = 2)
```

The wild swings early on calm into a tight line around the true $\pi$ (red). The error shrinks like $1/\sqrt{N}$ — so to get one more decimal place of accuracy you need about **100×** more points. That slow-but-sure convergence is the signature of every Monte Carlo method. Try `uniroot()`/`integrate()` for R's exact tools, or estimate a harder integral the same way for a real challenge.
:::

# Reflection

Monte Carlo estimation translates a **probabilistic argument** into a **simulation-based computation**.