---
title: "Markov Chains"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

# Mathematical Intuition

A **Markov chain** is a model for a system that hops between a set of **states** over time, where the rule for where it goes next depends *only on where it is now* — not on the whole history of how it got there. That "no memory of the past" property is the defining feature.

Picture the weather, simplified to three states — **Sunny**, **Cloudy**, **Rainy**. If we assume tomorrow's weather depends only on today's (not on the whole week), the weather is a Markov chain. All we need to describe it is, for each state, the probabilities of moving to each other state.

::: {.callout-note title="In a nutshell: the Markov property"}
"The future depends on the present, not the past." If you know the system is in state $X$ *right now*, knowing how it arrived there tells you nothing extra about where it goes next. This memorylessness is what makes Markov chains so easy to simulate — at every step you only need to look at the current state.
:::

We collect those move-probabilities into a **transition matrix** $P$, where the entry $P_{ij}$ is the probability of going **from** state $i$ **to** state $j$. Every row is a probability distribution, so **each row sums to 1**. For our weather:

$$
P = \begin{array}{c|ccc}
 & \text{Sun} & \text{Cloud} & \text{Rain} \\\hline
\text{Sun}   & 0.6 & 0.3 & 0.1 \\
\text{Cloud} & 0.3 & 0.4 & 0.3 \\
\text{Rain}  & 0.2 & 0.4 & 0.4 \\
\end{array}
$$

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart LR
  S["☀️ Sunny"] -->|0.6| S
  S -->|0.3| C["☁️ Cloudy"]
  S -->|0.1| R["🌧️ Rainy"]
  C -->|0.3| S
  C -->|0.4| C
  C -->|0.3| R
  R -->|0.2| S
  R -->|0.4| C
  R -->|0.4| R
```

# Math → Algorithm → Code Mapping

| Math idea | R translation |
|---|---|
| states $\{1,\dots,k\}$ | `states <- c("Sunny", "Cloudy", "Rainy")` |
| transition matrix $P$ | a `matrix` with rows summing to 1 |
| one step: draw next state from row $P_{i\cdot}$ | `sample(states, 1, prob = P[current, ])` |
| run for $T$ steps | a `for` loop updating the current state |
| long-run behaviour | `table(chain) / T`, or the eigenvector of $P^\top$ |

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart TD
  A([Start in some state])
  B[Look up the current state's row of P]
  C[Draw the next state with those probabilities]
  D[Record it; it becomes the current state]
  E{Done T steps?}
  A --> B --> C --> D --> E
  E -->|No| B
  E -->|Yes| F([Return the chain])
```

# Implementation in R

First we encode the chain and **simulate** a long run of days. At each step we simply sample the next state using the current state's row of $P$ — a direct translation of the Markov rule.

```{r}
states <- c("Sunny", "Cloudy", "Rainy")

P <- matrix(c(0.6, 0.3, 0.1,
              0.3, 0.4, 0.3,
              0.2, 0.4, 0.4),
            nrow = 3, byrow = TRUE,
            dimnames = list(states, states))

set.seed(1)
nDays <- 10000
chain <- character(nDays)
chain[1] <- "Sunny"                                  # start sunny

for (t in 2:nDays) {
  chain[t] <- sample(states, size = 1, prob = P[chain[t - 1], ])
}

head(chain, 15)                                      # the first fortnight of simulated weather
```

## The stationary distribution

Run the chain long enough and the *fraction of days* spent in each state settles down to a fixed set of proportions — the **stationary distribution** $\pi$. It answers "in the long run, what fraction of days are sunny?", and remarkably it does *not* depend on where we started. The simplest way to find it is to count.

```{r}
table(chain) / nDays                                 # long-run fraction of days in each state
```

The stationary distribution is also the solution of $\pi P = \pi$ — the distribution that the chain leaves unchanged. Mathematically that makes $\pi$ the (left) **eigenvector of $P$ with eigenvalue 1**, which we can compute exactly rather than by simulation.

```{r}
eig <- eigen(t(P))                                   # work with t(P): left eigenvectors of P
stat <- Re(eig$vectors[, which.max(Re(eig$values))]) # the eigenvector for eigenvalue 1
stat <- stat / sum(stat)                             # normalise to a probability distribution
names(stat) <- states
round(stat, 4)
```

A third route makes the convergence visible: multiply $P$ by itself many times. Every row of $P^{n}$ converges to the *same* stationary distribution — whatever today's weather, the forecast far enough ahead is the same.

```{r}
Pn <- P
for (i in 1:50) Pn <- Pn %*% P                       # P raised to the 51st power
round(Pn, 4)                                          # every row is now (almost) identical
```

All three methods — simulation, eigenvector, and matrix powers — agree on the same long-run answer. That agreement is the whole theory of Markov chains in miniature.

# Reflection

A Markov chain turns a simple, local rule ("given today, here is the chance of each tomorrow") into rich global behaviour (the long-run climate). This is a recurring theme in computational statistics: **complex distributions built from simple, repeated random steps.**

::: {.callout-tip title="Going further 🚀"}
This page is secretly the engine room of modern Bayesian statistics. **Markov chain Monte Carlo (MCMC)** — the Metropolis sampler on the *Bayesian computation* page — runs a cleverly designed Markov chain whose **stationary distribution is the posterior you want to sample from**. Everything here transfers: the chain wanders, forgets its start (that is the "burn-in"), and settles into $\pi$. Other places the same idea appears: **Google's PageRank** is the stationary distribution of a random surfer clicking links (a Markov chain on billions of web pages); **absorbing** chains model processes that eventually stop (gambler's ruin, board games); and **continuous-time** chains model queues and populations. In R, the `markovchain` package provides ready-made tools, and `expm::%^%` computes matrix powers directly. Try changing a row of `P` and watch the stationary distribution shift.
:::
