---
title: "Deep Learning: Neural Networks"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

**Deep learning** is the part of machine learning behind the headlines — image recognition, speech, self-driving cars, and the large language models that power tools like ChatGPT. At its core it is built from **neural networks**: flexible models, loosely inspired by the brain, that stack many simple computing units into something able to learn extraordinarily complex patterns. This page demystifies them, fits a small one in R, and shows the modern frameworks in both R and Python.

::: {.callout-note title="In a nutshell: what makes it 'deep'"}
A neural network is just a very flexible function fitted to data. "**Deep**" means it has **many layers** stacked between the input and the output. Each layer learns to transform the data a little, and stacking many layers lets the network build up from simple features (edges in an image) to complex ones (faces) — without you ever specifying those features by hand.
:::

# The building block: a neuron

Everything starts with a single **neuron**. It takes some inputs $x_1, \dots, x_p$, forms a **weighted sum** plus a bias, and passes the result through a non-linear **activation function** $g$:
$$
\text{output} = g\!\left(w_1 x_1 + w_2 x_2 + \dots + w_p x_p + b\right).
$$
This should look familiar — the part inside the brackets is exactly a **linear predictor**, the same $X\beta$ from linear models and GLMs. A single neuron with a sigmoid activation *is* logistic regression. The power comes from the activation and from wiring many neurons together.

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart LR
  X1["x₁"] --> S(("Σ wᵢxᵢ + b"))
  X2["x₂"] --> S
  X3["x₃"] --> S
  S --> A["activation g(·)"]
  A --> O["output"]
```

The **activation function** is what lets a network bend and curve. Common choices: **ReLU** ($\max(0, z)$, the modern default for hidden layers), **sigmoid** (squashes to $[0,1]$, for probabilities), and **softmax** (turns a vector into class probabilities, for the output of a classifier).

# From neurons to a network

Stack neurons into **layers**, and connect the layers in sequence: an **input layer** (one node per feature), one or more **hidden layers**, and an **output layer** (one node per thing to predict). Information flows left to right — the *forward pass* — each layer feeding the next.

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart LR
  subgraph In["Input layer"]
    I1["x₁"]
    I2["x₂"]
  end
  subgraph Hid["Hidden layer(s)"]
    H1["○"]
    H2["○"]
    H3["○"]
  end
  subgraph Out["Output layer"]
    O1["ŷ"]
  end
  I1 --> H1 & H2 & H3
  I2 --> H1 & H2 & H3
  H1 & H2 & H3 --> O1
```

A network with even one hidden layer is a **universal approximator** — given enough neurons it can represent essentially any function. Adding *depth* (more layers) lets it do so far more efficiently, learning a hierarchy of features.

# How a network learns

A freshly built network has random weights and predicts nonsense. Learning means **tuning the weights to reduce error**, and it is the same optimisation idea as everywhere else in the course — just at scale.

1. **Loss function** — a number measuring how wrong the predictions are (squared error for regression, cross-entropy for classification). This is the negative log-likelihood in disguise.
2. **Gradient descent** — nudge every weight a little in the direction that *decreases* the loss. The step size is the **learning rate**.
3. **Backpropagation** — the chain rule applied cleverly to compute the loss's gradient with respect to *every* weight at once, even across many layers.
4. **Epochs** — repeat over the whole dataset many times until the loss stops improving.

::: {.callout-note title="In a nutshell: it's just 'guess, measure, adjust'"}
The network **guesses**, a loss function **measures** how far off it is, and gradient descent **adjusts** the weights a touch to do better next time — millions of times over. Backpropagation is simply the bookkeeping that works out which weights to nudge and in which direction. No magic, just calculus and repetition.
:::

# A small neural network in R

R ships with the `nnet` package (a *recommended* package, so no install needed), which fits a classic single-hidden-layer network. Let us classify the `iris` species — the same task as the supervised-learning page, now with a neural net.

```{r}
#| warning: false
#| message: false

library(nnet)
set.seed(1)

# Train/test split
trainIdx <- sample(nrow(iris), 0.7 * nrow(iris))
trainSet <- iris[trainIdx, ]
testSet  <- iris[-trainIdx, ]

# A network with 5 hidden units; 'decay' is weight regularisation (curbs overfitting)
net <- nnet(Species ~ ., data = trainSet,
            size = 5, decay = 0.01, maxit = 200, trace = FALSE)

# Accuracy on the held-out test set
pred <- predict(net, testSet, type = "class")
mean(pred == testSet$Species)
```

```{r}
table(predicted = pred, actual = testSet$Species)
```

That single hidden layer already classifies the held-out flowers almost perfectly. `nnet` is tiny by modern standards, but the principles — hidden units, an activation, weight decay to fight overfitting, iterative training — are exactly those of the giant networks below.

# The modern frameworks (R and Python)

Real deep learning uses frameworks built for **many layers and GPU acceleration**. The two dominant engines are **TensorFlow/Keras** and **PyTorch**; both have first-class interfaces in R *and* Python. The same small network looks like this:

```{r}
#| eval: false

# R, using keras3 (a friendly interface to TensorFlow)
library(keras3)

model <- keras_model_sequential() |>
  layer_dense(units = 8, activation = "relu", input_shape = 4) |>
  layer_dense(units = 3, activation = "softmax")        # 3 species

model |> compile(optimizer = "adam",
                 loss = "sparse_categorical_crossentropy",
                 metrics = "accuracy")

model |> fit(xTrain, yTrain, epochs = 50, validation_split = 0.2)
```

```{python}
#| eval: false

# Python, using Keras (TensorFlow) — the same network
from tensorflow import keras

model = keras.Sequential([
    keras.layers.Dense(8, activation="relu", input_shape=(4,)),
    keras.layers.Dense(3, activation="softmax"),
])

model.compile(optimizer="adam",
              loss="sparse_categorical_crossentropy",
              metrics=["accuracy"])

model.fit(X_train, y_train, epochs=50, validation_split=0.2)
```

Read across: a `Dense` ("fully connected") layer, an activation, a loss, an optimiser, and `fit` over epochs — identical concepts in both languages.

::: {.callout-warning title="Deep learning is not always the answer"}
Neural networks are **data-hungry** and easy to over-engineer. For the **tabular data** most statisticians meet (rows and columns in a spreadsheet), a well-tuned **gradient-boosting** model (`xgboost`, `lightgbm`) or even a humble GLM usually matches or beats a neural net, trains in seconds, and is easier to interpret. Reach for deep learning when the data is **large and unstructured** — images, audio, text — where its feature-learning genuinely shines.
:::

# Where deep learning goes from here

The feedforward network above is the foundation; specialised architectures handle specialised data:

- **Convolutional neural networks (CNNs)** — exploit spatial structure for **images** (medical scans, object detection).
- **Recurrent networks and Transformers** — handle **sequences**; the **Transformer** is the architecture behind modern **large language models** like ChatGPT.
- **Transfer learning** — start from a giant pre-trained network and fine-tune it on your small dataset, getting state-of-the-art results without state-of-the-art data.

::: {.callout-tip title="Going further 🚀"}
To go hands-on: in **R**, `torch` (native, no Python needed) and `keras3` are the two routes; in **Python**, `PyTorch` and `keras`/`tensorflow` dominate, with `fastai` and `pytorch-lightning` as high-level wrappers and **Hugging Face `transformers`** for pre-trained language models. A **GPU** (free on Google Colab or Kaggle) turns hours of training into minutes. And before you build anything bespoke, remember the tabular-data lesson above: benchmark against `xgboost` first.
:::

# Summary

| Concept | What it is |
|---|---|
| **Neuron** | A weighted sum of inputs through an activation — a linear predictor with a twist |
| **Layer / depth** | Neurons stacked; many hidden layers = "deep" |
| **Activation** | ReLU (hidden), sigmoid/softmax (output) — adds non-linearity |
| **Loss** | How wrong the predictions are (cross-entropy, squared error) |
| **Gradient descent + backprop** | Nudge weights to cut the loss, repeatedly |
| **Frameworks** | `nnet` (small), `torch`/`keras` (R and Python) for real depth |
| **When to use** | Large, unstructured data (images, text); not usually tabular |

Deep learning is the modelling you already know — a linear predictor, a loss, and gradient-based fitting — scaled up with many layers and a great deal of data. It is the natural frontier beyond this course, and the concepts here are the foundation everything else is built on.
