---
title: "Regularisation and Cross-Validation"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

In the Modelling section we fitted linear models and worried about which variables to keep. This page picks up exactly where that left off, but from a machine-learning angle: instead of adding or removing whole variables by hand, we let the data *automatically* shrink the coefficients of unhelpful variables. This is called **regularisation**, and we use **cross-validation** to decide how much shrinking to do.

We will work entirely with the built-in `mtcars` dataset, predicting fuel economy `mpg` from the other 10 predictors. Many of these predictors (weight, displacement, number of cylinders, horsepower...) measure closely related aspects of "how big and powerful is the engine", so they are heavily correlated. That correlation is precisely what makes ordinary least squares struggle and regularisation shine.

We show every idea twice: once in **R** (which you can run) and once in **Python** (shown for comparison, not executed here). The two snippets always solve the same task.

# The problem: overfitting and the bias–variance trade-off

Recall the *principle of parsimony* from the Linear modelling section: given two models that behave similarly, prefer the simpler one. There is a deep statistical reason for this. With 10 predictors and only 32 rows, ordinary least squares (OLS) has so much freedom that it starts fitting the random **noise** in our particular sample rather than the true underlying signal. The model looks fantastic on the data it was trained on, then performs poorly on new cars it has never seen. That gap is **overfitting**.

The formal way to think about this is the **bias–variance trade-off**. The expected prediction error of a model can be decomposed as

$$
\mathbb{E}\big[(y - \hat{f}(x))^2\big] = \underbrace{\big(\text{Bias}[\hat{f}(x)]\big)^2}_{\text{too simple}} + \underbrace{\text{Var}[\hat{f}(x)]}_{\text{too flexible}} + \underbrace{\sigma^2}_{\text{irreducible noise}}.
$$

A model that is too simple has high **bias** (it systematically misses the pattern). A model that is too flexible has high **variance** (it lurches around as the training data changes). Regularisation deliberately adds a little bias in exchange for a large reduction in variance, which lowers the *total* error.

::: {.callout-note title="In a nutshell: what overfitting really means"}
Imagine revising for an exam by **memorising the answers to last year's paper** word-for-word. You will score 100% on last year's paper, but you have learnt nothing transferable, so you do badly on this year's. Overfitting is the model equivalent: it memorises the quirks of the training data instead of learning the general rule. We want a model that *generalises* to new cars, not one that recites the 32 cars it has already seen. Adding more predictors than the data can support is one of the easiest ways to overfit.
:::

Let us first fit a plain OLS model so we have a baseline. This connects directly to the GLM and AIC-based model selection you saw in Modelling: AIC penalises complexity to avoid exactly this trap, and regularisation is a smooth, automatic version of the same idea.

```{r}
#| label: ols-baseline
set.seed(2026)

# Plain ordinary least squares with all 10 predictors
olsModel <- lm(mpg ~ ., data = mtcars)
summary(olsModel)$coefficients

# Coefficient of determination on the TRAINING data (optimistic!)
summary(olsModel)$r.squared
```

Notice how high the training $R^2$ is, yet several coefficients have large standard errors and non-significant $p$-values. That instability is the symptom of correlated predictors and too little data — a textbook case for regularisation.

```{python}
#| eval: false
import statsmodels.api as sm
import statsmodels.formula.api as smf

# mtcars ships with statsmodels' R datasets; or load your own X, y.
mtcars = sm.datasets.get_rdataset("mtcars").data

olsModel = smf.ols("mpg ~ cyl + disp + hp + drat + wt + qsec + vs + am + gear + carb",
                   data=mtcars).fit()
print(olsModel.params)
print(olsModel.rsquared)   # optimistic training R^2
```

# Ridge regression (the L2 penalty)

**Ridge regression** keeps all the predictors but discourages large coefficients. Instead of minimising just the residual sum of squares, it minimises

$$
\sum_{i=1}^{n}\big(y_i - x_i^\top \beta\big)^2 \;+\; \lambda \sum_{j=1}^{p} \beta_j^2,
$$

where the second term is the **L2 penalty** (the squared length of the coefficient vector). The tuning parameter $\lambda \ge 0$ controls the strength of shrinkage: at $\lambda = 0$ we recover OLS, and as $\lambda \to \infty$ all coefficients are pulled toward zero. Ridge never sets a coefficient *exactly* to zero, but it tames the wild, correlated coefficients that cause high variance.

Ridge is available in the **MASS** package, which ships with R (it is a "recommended" package), so the chunk below runs.

```{r}
#| label: ridge-massridge
library(MASS)
set.seed(2026)

# Try a grid of lambda values
lambdaGrid <- seq(0, 50, length.out = 200)
ridgeFit <- lm.ridge(mpg ~ ., data = mtcars, lambda = lambdaGrid)

# The generalised cross-validation (GCV) criterion suggests a good lambda
bestLambda <- lambdaGrid[which.min(ridgeFit$GCV)]
bestLambda

# Coefficients at the chosen lambda (note they are shrunk toward zero)
coef(lm.ridge(mpg ~ ., data = mtcars, lambda = bestLambda))
```

```{r}
#| label: ridge-pathplot
# Visualise how each coefficient shrinks as lambda grows
matplot(lambdaGrid, t(ridgeFit$coef), type = "l", lty = 1,
        xlab = expression(lambda), ylab = "Standardised coefficient",
        main = "Ridge coefficient paths for mtcars")
abline(h = 0, col = "grey60", lty = 2)
abline(v = bestLambda, col = "#31BAE9", lwd = 2)
```

The matching Python code uses `scikit-learn`'s `Ridge`. Scikit-learn calls the penalty strength `alpha` rather than `lambda`, and it is good practice to standardise the predictors first (ridge is not scale-invariant).

```{python}
#| eval: false
import numpy as np
import statsmodels.api as sm
from sklearn.linear_model import Ridge
from sklearn.preprocessing import StandardScaler

mtcars = sm.datasets.get_rdataset("mtcars").data
X = mtcars.drop(columns="mpg").values
y = mtcars["mpg"].values

Xstd = StandardScaler().fit_transform(X)

ridge = Ridge(alpha=10.0)          # alpha is sklearn's name for lambda
ridge.fit(Xstd, y)
print(ridge.coef_)                 # shrunk coefficients
```

# Lasso (the L1 penalty)

The **lasso** (least absolute shrinkage and selection operator) swaps the squared penalty for the **L1 penalty** — the sum of the *absolute* values of the coefficients:

$$
\sum_{i=1}^{n}\big(y_i - x_i^\top \beta\big)^2 \;+\; \lambda \sum_{j=1}^{p} |\beta_j|.
$$

This small change has a big consequence: as $\lambda$ grows, the lasso pushes some coefficients **exactly to zero**, effectively dropping those predictors from the model. So the lasso does shrinkage *and* automatic variable selection at the same time — a data-driven version of the parsimony principle. For correlated predictors like `disp`, `wt` and `hp`, the lasso will often keep one and discard the others.

The most popular lasso implementation in R is **glmnet**. It is not a recommended package (you would install it with `install.packages("glmnet")`), so we mark the chunk as not evaluated, but this is exactly what you would run.

```{r}
#| eval: false
#| label: lasso-glmnet
library(glmnet)
set.seed(2026)

# glmnet wants a numeric predictor matrix x and response vector y
x <- model.matrix(mpg ~ ., data = mtcars)[, -1]   # drop the intercept column
y <- mtcars$mpg

# alpha = 1 selects the LASSO (alpha = 0 would be ridge)
lassoFit <- glmnet(x, y, alpha = 1)

# Coefficients at a particular lambda; some are forced to exactly zero
coef(lassoFit, s = 1.0)

# Plot the coefficient paths against the L1 norm
plot(lassoFit, xvar = "lambda", label = TRUE)
```

The Python equivalent is `scikit-learn`'s `Lasso`. Again `alpha` plays the role of $\lambda$, and standardising first is recommended.

```{python}
#| eval: false
import statsmodels.api as sm
from sklearn.linear_model import Lasso
from sklearn.preprocessing import StandardScaler

mtcars = sm.datasets.get_rdataset("mtcars").data
X = mtcars.drop(columns="mpg").values
y = mtcars["mpg"].values
Xstd = StandardScaler().fit_transform(X)

lasso = Lasso(alpha=0.5)
lasso.fit(Xstd, y)
print(lasso.coef_)                  # several entries will be exactly 0.0
```

# Cross-validation: choosing the penalty strength

We have been waving our hands at "pick a good $\lambda$". How do we actually do it? We cannot use the training error, because that always prefers $\lambda = 0$ (OLS), which overfits. We need to estimate how the model performs on **unseen** data, and the standard tool for that is **k-fold cross-validation (CV)**.

The recipe is:

1. Split the data into $k$ roughly equal parts ("folds"), say $k = 10$.
2. Hold out fold 1, train the model on the other $k-1$ folds, and measure the prediction error on the held-out fold 1.
3. Repeat, holding out fold 2, then fold 3, and so on, until every fold has been the test set exactly once.
4. Average the $k$ test errors. This average is an honest estimate of out-of-sample error.

We run this whole procedure for each candidate $\lambda$ and pick the $\lambda$ with the lowest cross-validated error (a common refinement is the "one-standard-error rule", which picks the simplest model within one standard error of the best).

::: {.callout-note title="In a nutshell: what cross-validation buys you"}
Cross-validation lets you **test the model on data it has not seen, without needing a separate fresh dataset**. By rotating which slice is held out, every observation gets used for both training and testing — just never at the same time. The pay-off is an honest estimate of how well the model will generalise, which is exactly what we need to choose $\lambda$ without fooling ourselves with optimistic training-set numbers.
:::

In R, `glmnet` provides `cv.glmnet`, which runs the entire k-fold search for you. As with the lasso chunk above, glmnet is not a recommended package, so this is shown but not run.

```{r}
#| eval: false
#| label: cv-glmnet
library(glmnet)
set.seed(2026)

x <- model.matrix(mpg ~ ., data = mtcars)[, -1]
y <- mtcars$mpg

# 10-fold cross-validated lasso
cvFit <- cv.glmnet(x, y, alpha = 1, nfolds = 10)

# The lambda with the lowest CV error, and the parsimonious "1-SE" choice
cvFit$lambda.min
cvFit$lambda.1se

# Plot CV error against log(lambda)
plot(cvFit)

# Coefficients at the cross-validated best lambda
coef(cvFit, s = "lambda.min")
```

The Python equivalent is `LassoCV`, which bakes the cross-validation directly into the fit. Alternatively `cross_val_score` cross-validates any estimator you hand it.

```{python}
#| eval: false
import statsmodels.api as sm
from sklearn.linear_model import LassoCV
from sklearn.preprocessing import StandardScaler

mtcars = sm.datasets.get_rdataset("mtcars").data
X = mtcars.drop(columns="mpg").values
y = mtcars["mpg"].values
Xstd = StandardScaler().fit_transform(X)

# LassoCV searches a grid of alphas with 10-fold CV
lassoCv = LassoCV(cv=10, random_state=2026).fit(Xstd, y)
print("Best alpha:", lassoCv.alpha_)
print(lassoCv.coef_)

# Or cross-validate any model explicitly:
# from sklearn.model_selection import cross_val_score
# scores = cross_val_score(lassoCv, Xstd, y, cv=10, scoring="r2")
```

# Going further

We have met the two classic penalties. In practice you rarely stop there, and the tooling is richer than the individual function calls above.

::: {.callout-tip title="Going further 🚀"}
**Elastic net** blends the L1 and L2 penalties, minimising
$$
\sum_i (y_i - x_i^\top\beta)^2 + \lambda\Big[\alpha\textstyle\sum_j |\beta_j| + (1-\alpha)\textstyle\sum_j \beta_j^2\Big].
$$
It enjoys the variable selection of the lasso while handling groups of correlated predictors more gracefully (the lasso tends to pick one and drop the rest arbitrarily). In `glmnet` you simply set `alpha` between 0 and 1; in scikit-learn you use `ElasticNet`.

Once you have several models and tuning parameters to juggle, you will want a **unified workflow** that handles preprocessing, resampling and tuning consistently:

- **R:** [`tidymodels`](https://www.tidymodels.org/) (the modern, tidyverse-friendly framework) or the older [`caret`](https://topepo.github.io/caret/) package wrap fitting, cross-validation and grid search behind one tidy interface.
- **Python:** combine a [`Pipeline`](https://scikit-learn.org/stable/modules/compose.html) (so scaling is refitted inside every CV fold, avoiding data leakage) with [`GridSearchCV`](https://scikit-learn.org/stable/modules/grid_search.html) to search over `alpha` and the L1/L2 mix automatically.

These tools turn the manual steps on this page into a few clean, reproducible lines.
:::