---
title: "Bayesian Forecasting: Time Series"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

Everything so far has estimated a parameter from data already in hand. **Forecasting** points the same machinery at the future: given a series observed up to today, what happens next? This is where the Bayesian approach really shines. A classical forecast hands you a point prediction with an interval bolted on; a Bayesian forecast hands you an entire **predictive distribution** for every future time point, built by propagating *two* sources of uncertainty at once — uncertainty about the model's parameters, and the inherent randomness of future shocks. It is the **posterior predictive check** from the worked-examples page, simply run *forwards in time*.

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart LR
  D["Observed series<br/>y₁ … yₙ"] --> M["Fit a time-series model<br/>(MCMC → posterior for θ)"]
  M --> S["For each posterior draw,<br/>simulate the future forward"]
  S --> F["Posterior predictive forecast<br/>a full distribution per horizon<br/>→ median + credible band"]
```

# A simple time-series model: the AR(1)

The workhorse starting model is the **first-order autoregression**, AR(1): each value is a fraction of the previous one plus a fresh random shock.
$$
y_t = c + \phi\, y_{t-1} + \epsilon_t, \qquad \epsilon_t \sim \mathcal N(0, \sigma^2).
$$
Three parameters: an intercept $c$, the **autoregressive coefficient** $\phi$ (how strongly today depends on yesterday), and the shock size $\sigma$. For the series to be *stationary* — to revert toward a stable level rather than wander off — we need $|\phi| < 1$, which we encode as a flat prior on $(-1, 1)$.

::: {.callout-note title="In a nutshell: what AR(1) says"}
"Tomorrow looks like a shrunk-down version of today, plus a random nudge." If $\phi = 0.8$, each year keeps 80% of this year's deviation from the long-run average and forgets the rest. The closer $\phi$ is to 1, the longer the memory and the slower the series drifts back to its mean.
:::

We use R's built-in `LakeHuron` — the annual level of Lake Huron from 1875 to 1972 — a classic, gently autocorrelated series.

```{r}
#| fig-align: center

y <- as.numeric(LakeHuron)
n <- length(y)

plot(1875:1972, y, type = "l", lwd = 2, col = "steelblue",
     xlab = "Year", ylab = "Lake level (ft)",
     main = "Lake Huron annual level, 1875–1972")
```

# Fitting it the Bayesian way (MCMC)

There is no conjugate shortcut here, so we reuse the **Metropolis sampler** from the *Bayesian computation* page. The likelihood is the AR(1) model written as a product of normals — each $y_t$ is normal around $c + \phi y_{t-1}$ — and we sample $\sigma$ on the log scale (keeping it positive) with the usual Jacobian correction.

```{r}
#| warning: false
#| message: false

yPrev <- y[-n]    # y_{t-1}  (t = 2 … n)
yCurr <- y[-1]    # y_t

# log posterior for par = (c, phi, log sigma)
logPost <- function(par) {
  cc <- par[1]; phi <- par[2]; sigma <- exp(par[3])
  if (abs(phi) >= 1) return(-Inf)                    # stationarity: flat prior on (-1, 1)
  ll <- sum(dnorm(yCurr, mean = cc + phi * yPrev, sd = sigma, log = TRUE))
  lp <- dnorm(cc, 0, 100, log = TRUE) +              # weak priors
        dnorm(sigma, 0, 10, log = TRUE)
  ll + lp + par[3]                                   # + Jacobian for log sigma
}

set.seed(1)

# Seed the sampler at the ordinary least-squares AR(1) fit (a sensible starting point)
ols   <- lm(yCurr ~ yPrev)
start <- c(coef(ols)[[1]], coef(ols)[[2]], log(summary(ols)$sigma))

nIter <- 20000
step  <- c(6, 0.05, 0.10)                            # random-walk step sizes per parameter
chain <- matrix(NA, nIter, 3, dimnames = list(NULL, c("c", "phi", "logSigma")))
cur   <- start; lpCur <- logPost(cur); nAcc <- 0

for (i in seq_len(nIter)) {
  prop   <- cur + rnorm(3, 0, step)
  lpProp <- logPost(prop)
  if (log(runif(1)) < lpProp - lpCur) {
    cur <- prop; lpCur <- lpProp; nAcc <- nAcc + 1
  }
  chain[i, ] <- cur
}

nAcc / nIter      # acceptance rate (aim for roughly 0.2–0.5)
```

Drop the burn-in and read off the posterior for each parameter:

```{r}
draws <- chain[-(1:4000), ]
post  <- cbind(draws[, c("c", "phi")], sigma = exp(draws[, "logSigma"]))

round(t(apply(post, 2, function(z)
  c(mean = mean(z), `2.5%` = quantile(z, 0.025), `97.5%` = quantile(z, 0.975)))), 3)
```

The posterior for $\phi$ sits well inside $(-1,1)$ and clearly above 0 — strong, persistent year-to-year memory — with a credible interval that quantifies how sure we are. We have not a single fitted model but a *whole population* of plausible AR(1) models, weighted by the posterior. That is precisely what makes the forecast honest.

# Forecasting: simulate the future forward

Here is the Bayesian forecasting recipe in full. For **each** posterior draw $(c, \phi, \sigma)$, start from the last observed value and roll the AR(1) equation forward, drawing a fresh shock $\epsilon_t$ at every step. Each draw produces one possible *future path*; together, thousands of paths trace out the **posterior predictive distribution** at every horizon.

```{r}
set.seed(2)

h     <- 25                       # forecast 25 years ahead
nDraw <- nrow(draws)
paths <- matrix(NA, nDraw, h)     # one simulated future per posterior draw

for (d in seq_len(nDraw)) {
  cc <- draws[d, "c"]; phi <- draws[d, "phi"]; sigma <- exp(draws[d, "logSigma"])
  prev <- y[n]                                    # start from the last observation
  for (t in seq_len(h)) {
    prev <- cc + phi * prev + rnorm(1, 0, sigma)  # AR(1) step with a new random shock
    paths[d, t] <- prev
  }
}

# Summarise the cloud of paths into a median and a 95% predictive band per year
bands <- apply(paths, 2, quantile, c(0.025, 0.5, 0.975))
```

```{r}
#| fig-align: center

obsYears <- 1875:1972
fYears   <- 1973:(1972 + h)

plot(obsYears, y, type = "l", lwd = 2, col = "steelblue",
     xlim = c(1875, max(fYears)), ylim = range(y, bands),
     xlab = "Year", ylab = "Lake level (ft)",
     main = "Bayesian AR(1) forecast with 95% predictive band")
polygon(c(fYears, rev(fYears)), c(bands[1, ], rev(bands[3, ])),
        col = adjustcolor("firebrick", 0.2), border = NA)
lines(fYears, bands[2, ], lwd = 2, col = "firebrick")
abline(v = 1972, lty = 3, col = "grey50")
legend("bottomleft", bty = "n", lwd = 2, col = c("steelblue", "firebrick"),
       legend = c("observed", "forecast median + 95% band"))
```

The forecast shows the signature shape of an autoregressive prediction: the median **decays back toward the series' long-run mean**, while the band **fans out** as we look further ahead and uncertainty compounds. Crucially, that band already includes *both* kinds of uncertainty — we did not know the parameters exactly (different draws), and the future is random anyway (the shocks) — without a single extra formula.

# The Bayesian pay-off: ask any question of the future

Because a forecast is a full distribution of simulated futures, **any** question becomes a simple count over the `paths` matrix — no new theory, just `mean()`. What is the probability the lake drops below 578 ft a decade from now?

```{r}
mean(paths[, 10] < 578)      # posterior predictive P(level < 578 ft in year 10)
```

That single number — a genuine probability about the future — is the kind of statement a frequentist forecast cannot make, and it falls straight out of the simulated paths. You could just as easily ask for the chance of a record low, or the probability of two consecutive years above a threshold, by querying the same matrix.

::: {.callout-tip title="Going further 🚀"}
Real forecasting rarely stops at AR(1). The same Bayesian recipe — posterior via MCMC, then simulate forward — scales up through proper probabilistic-programming tools: **`brms`** fits autoregressive and many other time-series models with `ar()` terms in one formula; **`bsts`** (Bayesian structural time series) decomposes a series into trend, seasonality and regression components; and full **state-space / dynamic linear models** live in **`dlm`** and **`KFAS`**. For the classical (non-Bayesian) comparison, fit the same model with `arima(LakeHuron, order = c(1,0,0))` and `forecast::forecast()` and contrast the prediction intervals with the credible band above.
:::

# Summary

| Step | What we did |
|---|---|
| Model | AR(1): $y_t = c + \phi y_{t-1} + \epsilon_t$, stationary for $\lvert\phi\rvert<1$ |
| Fit | Metropolis MCMC → joint posterior for $(c, \phi, \sigma)$ |
| Forecast | For each draw, simulate the future forward → posterior predictive paths |
| Summarise | Median + 95% credible band per horizon (the fan) |
| Pay-off | Any probability about the future is a `mean()` over simulated paths |

Bayesian forecasting is the section's whole philosophy aimed at tomorrow: keep the *full* posterior, push it through the model's forward equation, and read decisions off the resulting distribution of futures — uncertainty about parameters and uncertainty about fate, handled in one honest sweep.