---
title: "Bayesian Testing and Model Comparison"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

The frequentist section turned estimation around to ask *"is $\theta = \theta_0$ believable?"* and answered with p-values and the Wald/score/likelihood-ratio trinity. Bayesian inference asks the same question but answers it from the posterior — and, crucially, can do something frequentism cannot: assign a **probability to a hypothesis directly**. This page covers the three Bayesian ways to weigh a hypothesis: reading **posterior probabilities** straight off the posterior, the **region of practical equivalence (ROPE)**, and the **Bayes factor** for formal model comparison. As before, no new machinery — it is all the posterior, pointed at a decision.

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart LR
  PO["Posterior<br/>p(θ | data)"] --> Q["Is θ = θ₀ / θ > θ₀ believable?"]
  Q --> A["Posterior probability<br/>P(θ in region | data)"]
  Q --> B["ROPE<br/>is θ₀ practically<br/>equivalent to the truth?"]
  Q --> C["Bayes factor<br/>compare H₀ vs H₁<br/>by marginal likelihood"]
```

# The Bayesian advantage: probabilities of hypotheses

Recall the central frequentist caution: a p-value is $P(\text{data}\mid H_0)$, **not** $P(H_0 \mid \text{data})$ — the latter is forbidden because $\theta$ is not random. The Bayesian has no such restriction. Because $\theta$ carries a posterior distribution, the probability of *any* statement about $\theta$ is just an area under that posterior. Want the probability the coin favours heads? Integrate the posterior above $0.5$:

$$
P(\theta > 0.5 \mid \text{data}) = \int_{0.5}^{1} p(\theta \mid \text{data})\, d\theta .
$$

This is the quantity people *think* a p-value gives them, and here it genuinely does.

```{r}
#| warning: false
#| message: false

set.seed(1)
trueTheta <- 0.7
n <- 200
k <- sum(rbinom(n, 1, trueTheta))

aPri <- 2; bPri <- 2
aPost <- aPri + k
bPost <- bPri + n - k

# Direct posterior probabilities — areas under the Beta posterior
pHeads <- 1 - pbeta(0.5, aPost, bPost)   # P(theta > 0.5 | data)

c(P_theta_gt_0.5 = pHeads)
```

The posterior probability that the coin favours heads is essentially 1 — a direct, interpretable answer with no appeal to hypothetical repetitions.

# Testing a point null: three approaches

Suppose, as on the frequentist page, we want to test whether the coin is fair, $H_0: \theta = 0.5$. There are three Bayesian routes, of increasing formality.

## 1. The credible interval (test–interval duality)

The quickest check reuses the duality from the frequentist section. **If $\theta_0$ falls outside the 95% credible interval, it is implausible** at that level. This is the Bayesian mirror of "reject $H_0$ if $\theta_0$ lies outside the confidence interval" — except the interval now carries a genuine posterior-probability interpretation.

```{r}
cred95 <- qbeta(c(0.025, 0.975), aPost, bPost)
cred95                        # is 0.5 inside?
```

The value $0.5$ sits far below the lower endpoint, so a fair coin is not a plausible description of this data.

## 2. The ROPE: practical, not exact, equivalence

A point null like "$\theta$ is *exactly* 0.5" is often a fiction — no real coin is exactly fair to infinite decimal places. The **region of practical equivalence (ROPE)** replaces the knife-edge with an interval of values "close enough to fair to not care," say $[0.45, 0.55]$, and asks how much posterior mass falls in it.

```{r}
rope <- c(0.45, 0.55)
pRope <- pbeta(rope[2], aPost, bPost) - pbeta(rope[1], aPost, bPost)
pRope
```

Almost none of the posterior lies in the ROPE, so we **reject practical equivalence to fair**: not only is the coin probably not *exactly* fair, it is probably not even *close*. (The opposite verdict — most of the posterior inside the ROPE — would let us *accept* the null as practically true, something a frequentist test can never do, only "fail to reject".)

## 3. The Bayes factor: formal model comparison

The fullest answer compares two models head to head by how well each *predicted the data before seeing it* — the **marginal likelihood** (the evidence $p(\text{data}\mid H)$ from Bayes' theorem) of each hypothesis. Their ratio is the **Bayes factor**:

$$
\boxed{\,\text{BF}_{10} = \frac{p(\text{data} \mid H_1)}{p(\text{data} \mid H_0)}\,}
$$

It updates the prior odds on the hypotheses into posterior odds:
$$
\underbrace{\frac{P(H_1\mid\text{data})}{P(H_0\mid\text{data})}}_{\text{posterior odds}} = \underbrace{\frac{P(H_1)}{P(H_0)}}_{\text{prior odds}} \times \underbrace{\text{BF}_{10}}_{\text{Bayes factor}} .
$$

::: {.callout-note title="In a nutshell: what \"odds\" means"}
**Odds** are just a ratio of probabilities: the odds on $H_1$ are $P(H_1)/P(H_0)$. So "odds of 3" means $H_1$ is three times as probable as $H_0$. The equation above says the Bayes factor *multiplies* your odds: a **Bayes factor of 10 turns your prior odds into ten times larger posterior odds** — if you started at even (1:1), you end at 10:1 in favour of $H_1$.
:::

$\text{BF}_{10} = 10$ means the data are ten times better predicted by $H_1$ than $H_0$. Here $H_0:\theta=0.5$ is a point, and $H_1$ spreads $\theta$ over its $\text{Beta}(a,b)$ prior. Both marginal likelihoods are available in closed form for the coin:

$$
p(\text{data}\mid H_0) \propto 0.5^{n}, \qquad p(\text{data}\mid H_1) \propto \frac{B(a+k,\,b+n-k)}{B(a,b)},
$$
where $B$ is the Beta function (the binomial coefficient cancels in the ratio).

::: {.callout-note title="In a nutshell: you can take these two formulas on trust"}
You do not need to derive these to use them. $H_0$ fixes $\theta = 0.5$, so it just plugs that one value into the likelihood, giving $0.5^{n}$. $H_1$ instead **averages** the likelihood over every $\theta$ weighted by the prior — $p(\text{data}\mid H_1)=\int \theta^{k}(1-\theta)^{n-k}\,p(\theta)\,d\theta$ — and because $p(\theta)$ is a Beta, that integral is a standard **Beta integral** with the tidy closed form above. Trust the result on a first read; the only idea that matters is "$H_1$ spreads its bet across all $\theta$, $H_0$ stakes everything on one."
:::

```{r}
# log marginal likelihoods (the binomial coefficient cancels, so omit it)
logMlH1 <- lbeta(aPri + k, bPri + n - k) - lbeta(aPri, bPri)
logMlH0 <- n * log(0.5)

bf10 <- exp(logMlH1 - logMlH0)
bf10
```

The Bayes factor is enormous — the data are overwhelmingly better explained by "the bias is unknown" than by "the bias is exactly $0.5$," so we decisively favour $H_1$. This is the Bayesian echo of the frequentist trinity all rejecting the fair-coin null on the very same dataset.

::: {.callout-note title="Reading a Bayes factor — the Jeffreys scale"}
A rough conventional calibration of $\text{BF}_{10}$:

| $\text{BF}_{10}$ | Evidence for $H_1$ |
|---|---|
| 1–3 | barely worth mentioning |
| 3–10 | substantial |
| 10–30 | strong |
| 30–100 | very strong |
| > 100 | decisive |

Unlike a p-value, the Bayes factor can also support the *null*: $\text{BF}_{10} < 1$ (i.e. $\text{BF}_{01} > 1$) is evidence *for* $H_0$, so Bayesian model comparison can confirm a null, not merely fail to reject it.
:::

## Cross-check: the Savage–Dickey density ratio

For a point null nested inside $H_1$, there is an elegant shortcut that should agree with the marginal-likelihood calculation — the **Savage–Dickey density ratio**: the Bayes factor in favour of $H_0$ is simply the posterior density at $\theta_0$ divided by the prior density at $\theta_0$.

$$
\text{BF}_{01} = \frac{p(\theta_0 \mid \text{data},\, H_1)}{p(\theta_0 \mid H_1)} .
$$

```{r}
bf01SD <- dbeta(0.5, aPost, bPost) / dbeta(0.5, aPri, bPri)
bf10SD <- 1 / bf01SD

c(BF10_marginal_lik = bf10,
  bf10SD = bf10SD)
```

The two routes to $\text{BF}_{10}$ agree, confirming the calculation independently — the same "verify by a second method" discipline the frequentist page used with `prop.test`. Intuitively the Savage–Dickey ratio asks: *did the data make $\theta_0$ more or less plausible than the prior did?* The posterior density at $0.5$ has collapsed to nearly zero relative to the prior, so $H_0$ is crushed.

# p-value vs posterior probability vs Bayes factor

It is worth lining the three philosophies up on the identical question — *is this coin fair?* — to see what each actually delivers.

| | **Frequentist p-value** | **Posterior probability** | **Bayes factor** |
|---|---|---|---|
| Question answered | $P(\text{data this extreme}\mid H_0)$ | $P(\theta \in \text{region}\mid\text{data})$ | how much better $H_1$ predicts than $H_0$ |
| Can state $P(H\mid\text{data})$? | No | Yes | Yes (via posterior odds) |
| Can support the null? | No — only "fail to reject" | Yes (mass in ROPE) | Yes ($\text{BF}_{10}<1$) |
| Needs a prior? | No | Yes | Yes (and is *sensitive* to it) |
| Our coin verdict | reject $H_0$ ($p\approx 0$) | $P(\theta>0.5)\approx 1$ | $\text{BF}_{10}$ huge |

All three reject the fair coin, just as the trinity did — when the data are this clear, the philosophies agree on the conclusion even as they disagree on its meaning.

::: {.callout-warning title="Bayes factors are sensitive to the prior on H₁"}
The credible interval and posterior probability are robust to the prior once data is plentiful (the washout from the priors page). The **Bayes factor is not**: widening the $H_1$ prior spreads its predictive mass thinner and can change $\text{BF}_{10}$ substantially — the *Lindley–Jeffreys paradox*. A vague prior on $H_1$ automatically penalises it. So always report the prior behind a Bayes factor, and check sensitivity, exactly as recommended for any prior.
:::

::: {.callout-tip title="Going further 🚀"}
For standard designs the `BayesFactor` package computes the marginal-likelihood ratio in one line, with sensible default priors:

```{r}
#| eval: false

library(BayesFactor)

proportionBF(y = k, N = n, p = 0.5)   # the fair-coin test of this page
ttestBF(x = group1, y = group2)       # Bayesian two-sample t-test
```

The modern alternative to Bayes factors is **predictive** model comparison, which sidesteps their prior-sensitivity: the `loo` package estimates out-of-sample fit via **WAIC** or **LOO-CV** from posterior draws, and compares models without integrating a marginal likelihood at all.

```{r}
#| eval: false

library(loo)

loo(fit1); loo(fit2)        # leave-one-out cross-validation per model
loo_compare(loo(fit1), loo(fit2))   # which predicts new data better?
```

:::

# Summary

| Tool | Statistic | What it tells you |
|---|---|---|
| **Posterior probability** | $P(\theta \in \text{region}\mid\text{data})$ | Direct probability of a hypothesis — the answer p-values only seem to give |
| **Credible interval** | is $\theta_0$ inside? | Quick interval–test duality check |
| **ROPE** | posterior mass in $[\theta_0\!-\!\delta,\theta_0\!+\!\delta]$ | Practical (not exact) equivalence; can *accept* a null |
| **Bayes factor** | $p(\text{data}\mid H_1)/p(\text{data}\mid H_0)$ | Formal model comparison; updates prior odds to posterior odds |

The Bayesian framework turns "testing" back into ordinary probability: every verdict is an area under the posterior or a ratio of how well rival models predicted the data. Unlike the frequentist trinity it can quantify evidence *for* a null and state the probability of a hypothesis outright — at the cost, as always, of a prior you must choose and defend. The next page runs this entire pipeline — prior, posterior, interval, test, and a model check — end to end on real data.