---
title: "Bayes' Theorem and the Posterior"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

The previous page promised that all of Bayesian inference is one act: take a **prior**, fold in the **likelihood**, and read off a **posterior**. This page makes that precise. We derive **Bayes' theorem**, name each piece, and then run the whole machine on the coin-flip model from the frequentist section — this time getting back not a point estimate and a standard error, but an entire distribution for $\theta$.

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart LR
  PR["Prior<br/>p(θ)"] --> M["Bayes' theorem"]
  LK["Likelihood<br/>p(data | θ)"] --> M
  M --> PO["Posterior<br/>p(θ | data)"]
  PO --> S["Summaries:<br/>posterior mean / MAP<br/>+ credible interval"]
```

# Deriving Bayes' theorem

Bayes' theorem is just the definition of conditional probability, read in a particular direction. For any two events $A$ and $B$,
$$
p(A \mid B) = \frac{p(A \cap B)}{p(B)} = \frac{p(B \mid A)\,p(A)}{p(B)} .
$$
Now replace $A$ with "the parameter takes value $\theta$" and $B$ with "we observed this data". The same identity becomes the cornerstone of inference:

$$
\boxed{\,p(\theta \mid \text{data}) = \frac{p(\text{data} \mid \theta)\,p(\theta)}{p(\text{data})}\,}
$$

Each term has a name and a job:

- $p(\theta)$ — the **prior**: our uncertainty about $\theta$ *before* the data.
- $p(\text{data} \mid \theta)$ — the **likelihood**: how plausibly each $\theta$ would have produced the observed data. *This is exactly the likelihood from the frequentist section.*
- $p(\text{data})$ — the **evidence** (or **marginal likelihood**): the probability of the data averaged over all values of $\theta$, $\;p(\text{data}) = \int p(\text{data}\mid\theta)\,p(\theta)\,d\theta$.
- $p(\theta \mid \text{data})$ — the **posterior**: our updated uncertainty about $\theta$ *after* the data.

# The proportionality shortcut

The evidence $p(\text{data})$ in the denominator does not depend on $\theta$ — it is a single number, the integral over *all* $\theta$. As far as the *shape* of the posterior in $\theta$ is concerned, it is just a normalising constant that forces the posterior to integrate to 1. So almost all the action is in the numerator:

$$
\boxed{\,p(\theta \mid \text{data}) \;\propto\; p(\text{data}\mid\theta)\,p(\theta)\,} \qquad \text{posterior} \;\propto\; \text{likelihood} \times \text{prior}.
$$

::: {.callout-important title="The evidence is the hard part"}
That harmless-looking denominator $p(\text{data}) = \int p(\text{data}\mid\theta)\,p(\theta)\,d\theta$ is an integral over the whole parameter space. For a few lucky models (this page) it has a closed form. For most realistic models it is **intractable** — and computing the posterior despite it is the entire reason the *Bayesian computation* page exists. For now we either pick models where the integral is free, or work up to the normalising constant and let R's distribution functions handle it.
:::

# The coin model, the Bayesian way

We reuse the **Bernoulli/coin-flip** model from the frequentist likelihood page: $n$ flips, $k$ heads, each flip heads with unknown probability $\theta$. The likelihood is
$$
p(\text{data}\mid\theta) \;\propto\; \theta^{k}(1-\theta)^{n-k}.
$$

For the prior we need a distribution on $\theta \in [0,1]$. The natural choice is the **Beta distribution**, $\theta \sim \text{Beta}(a, b)$, whose density is $p(\theta) \propto \theta^{a-1}(1-\theta)^{b-1}$. The two numbers $a$ and $b$ act like "prior heads" and "prior tails": $\text{Beta}(1,1)$ is flat (total ignorance), while $\text{Beta}(20,20)$ is a sharp belief the coin is near-fair.

Now multiply prior by likelihood:
$$
p(\theta \mid \text{data}) \;\propto\; \underbrace{\theta^{a-1}(1-\theta)^{b-1}}_{\text{prior}} \times \underbrace{\theta^{k}(1-\theta)^{n-k}}_{\text{likelihood}} = \theta^{(a+k)-1}(1-\theta)^{(b+n-k)-1}.
$$

The right-hand side is *another Beta density*. We can read off its parameters by inspection:

$$
\boxed{\,\theta \mid \text{data} \;\sim\; \text{Beta}\big(a + k,\; b + n - k\big)\,}
$$

No integral was harmed. The prior and posterior belong to the same family, so updating is just **arithmetic on the parameters**: add the observed heads to $a$, the observed tails to $b$.

::: {.callout-note title="In a nutshell: reading a Beta, and why the maths is so easy"}
Think of $a$ and $b$ as **prior heads and prior tails**. The bigger they are, the more sure you are, so the tighter the distribution; and the more *equal* they are, the more it concentrates around $0.5$. (So $\text{Beta}(1,1)$ is flat, $\text{Beta}(20,20)$ is a sharp spike on fairness, $\text{Beta}(70,30)$ is a confident lean toward heads.) The conjugacy step is no harder than it looks: multiplying two things of the form $\theta^{\,\cdot}(1-\theta)^{\,\cdot}$ just **adds the powers of $\theta$, and adds the powers of $(1-\theta)$** — which is exactly "add the heads, add the tails."
:::

::: {.callout-note title="Conjugacy"}
When the posterior lands in the same family as the prior, the prior is called **conjugate** to the likelihood. Beta is conjugate to the Binomial; we will meet Gamma–Poisson and Normal–Normal on the next page. Conjugacy is a convenience, not a requirement — it gives a closed-form posterior and a beautifully simple interpretation ("the prior is just extra pseudo-data"), but most real models are not conjugate and need the computation of a later page.
:::

# Worked example in R

Let us watch the prior turn into the posterior. We simulate the same coin as the frequentist pages — `set.seed(1)`, $n = 200$, true bias $0.7$ — so the two sections are looking at *identical* data.

```{r}
#| warning: false
#| message: false

set.seed(1)

# Same simulated coin as the frequentist section
trueTheta <- 0.7
n     <- 200
flips <- rbinom(n, size = 1, prob = trueTheta)
k     <- sum(flips)              # heads observed

# A weakly-informative prior: Beta(2, 2) — a gentle nudge toward 0.5, easily overruled
aPri <- 2
bPri <- 2

# Conjugate update: just add heads to a, tails to b
aPost <- aPri + k
bPost <- bPri + (n - k)

c(k = k, aPost = aPost, bPost = bPost)
```

That is the *entire* inference — two additions. Everything below is just reading information out of the resulting Beta posterior.

```{r}
#| fig-align: center

theta <- seq(0, 1, length.out = 1000)

# The three curves, each scaled to a density for comparison
priorD <- dbeta(theta, aPri, bPri)
postD  <- dbeta(theta, aPost,  bPost)
# Likelihood as a function of theta (scaled to integrate to ~1 for plotting)
lik     <- dbinom(k, n, theta);  likD <- lik / (sum(lik) * diff(theta)[1])

plot(theta, postD, type = "l", lwd = 3, col = "firebrick",
     xlab = expression(theta), ylab = "density",
     main = "Prior  ×  Likelihood  →  Posterior", xlim = c(0.4, 0.9))
lines(theta, priorD, lwd = 2, col = "darkgreen", lty = 2)
lines(theta, likD,   lwd = 2, col = "steelblue", lty = 3)
abline(v = trueTheta, col = "grey40", lty = 1)
legend("topright", bty = "n", lwd = c(2, 2, 3, 1), lty = c(2, 3, 1, 1),
       col = c("darkgreen", "steelblue", "firebrick", "grey40"),
       legend = c("prior  Beta(2,2)", "likelihood", "posterior", "true θ"))
```

The picture tells the whole story: the broad green prior, multiplied by the sharp blue likelihood, yields the red posterior — which sits essentially on top of the likelihood because, with $n = 200$, the data overwhelms the gentle prior. The posterior is our **complete** answer: not "the estimate is 0.7 ± something," but a full distribution saying how plausible *every* value of $\theta$ is.

# Summarising the posterior

A distribution is more than most reports need, so we compress it into a few numbers — the Bayesian analogues of the point estimate and confidence interval.

## Point summaries: posterior mean and MAP

Two natural "best single guesses":

- The **posterior mean** $\mathbb{E}[\theta \mid \text{data}]$ — the centre of mass. For a $\text{Beta}(a',b')$ posterior it is $a'/(a'+b')$.
- The **maximum a posteriori (MAP)** estimate — the *peak* of the posterior, the single most plausible value. With a flat prior the MAP *equals the frequentist MLE*, which is the first hint of the bridge between the two sections.

```{r}
postMean <- aPost / (aPost + bPost)
postMap  <- (aPost - 1) / (aPost + bPost - 2)   # mode of a Beta
mle      <- k / n                                  # the frequentist estimate

c(postMean = postMean, MAP = postMap, frequentist_MLE = mle)
```

All three are close to one another and to $0.7$ — with this much data and a weak prior, the Bayesian and frequentist point estimates nearly coincide.

## Interval summary: the credible interval

The Bayesian replacement for the confidence interval is the **credible interval**: an interval that contains $\theta$ with a stated posterior probability. The simplest is the **equal-tailed** interval — chop 2.5% off each tail of the posterior — which for a Beta is one call to `qbeta`:

```{r}
cred95 <- qbeta(c(0.025, 0.975), aPost, bPost)
cred95
```

::: {.callout-important title="Credible interval ≠ confidence interval"}
This interval supports the statement everyone *wants* to make: *"given the data and prior, there is a 95% probability that $\theta$ lies in this interval."* That is a probability statement **about $\theta$**, which is legitimate here only because $\theta$ is treated as random. The frequentist confidence interval from the previous section looks almost identical numerically but means something quite different — a statement about the long-run behaviour of the *procedure*, not about this particular $\theta$. Same numbers, different claims.
:::

There is a second flavour worth knowing. The **highest posterior density (HPD)** interval is the *shortest* interval containing 95% of the posterior mass — equivalently, every point inside it has higher density than every point outside. For a symmetric posterior it matches the equal-tailed one; for a skewed posterior it is narrower and is often the more honest summary.

```{r}
# Highest posterior density interval by a quick numerical search over the start tail-prob
hpdBeta <- function(a, b, level = 0.95) {
  f <- function(lowerTail) {
    lo <- qbeta(lowerTail, a, b)
    hi <- qbeta(lowerTail + level, a, b)
    hi - lo                                   # width to minimise
  }
  opt   <- optimize(f, c(0, 1 - level))
  lower <- opt$minimum
  c(qbeta(lower, a, b), qbeta(lower + level, a, b))
}

rbind(equal_tailed = cred95,
      HPD          = hpdBeta(aPost, bPost)) |> round(4)
```

Here the posterior is nearly symmetric, so the two intervals barely differ — but the HPD is the slightly shorter of the two, as it must be.

::: {.callout-tip title="Going further 🚀"}
You rarely need to hand-roll an HPD search like `hpdBeta` above. Given a vector of posterior draws (or a closed-form posterior), packages do it in one line:

```{r}
#| eval: false

HDInterval::hdi(draws, credMass = 0.95)          # from a sample of draws
bayestestR::ci(draws, ci = 0.95, method = "HDI") # tidy output, same idea
```

`HDInterval::hdi()` also accepts a distribution name (e.g. `hdi(qbeta, shape1 = aPost, shape2 = bPost)`) to get the exact-density HPD without sampling.
:::

# Summary

| Object | Symbol | What it is |
|---|---|---|
| **Prior** | $p(\theta)$ | Belief about $\theta$ before the data |
| **Likelihood** | $p(\text{data}\mid\theta)$ | The data's verdict on each $\theta$ (shared with the frequentist section) |
| **Evidence** | $p(\text{data})$ | Normalising constant; the integral that is usually hard |
| **Posterior** | $p(\theta\mid\text{data})$ | Updated belief — the complete answer |
| **Posterior mean / MAP** | $\mathbb{E}[\theta\mid\text{data}]$, mode | Point summaries of the posterior |
| **Credible interval** | — | Interval with stated *posterior probability* of containing $\theta$ |

The whole of Bayesian inference is "posterior $\propto$ likelihood $\times$ prior," followed by reading summaries off the posterior. The only thing standing between us and harder models is that the **evidence integral** is rarely free — which raises two questions we take up next: *where does the prior come from*, and *what do we do when the maths runs out?*