---
title: "dplyr: summary tables"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

We begin by loading in the cars dataset we used on the previous page (the code required is available below if needed). Please note that we want to use the version that has the manufacturers included! 

```{r}
#| code-fold: true

###| Code to create the dataset used in previous pages
cars <- mtcars

cars$am <- factor(cars$am)
cars$vs <- factor(cars$vs)

cars$manufacturer <- factor(rownames(cars))
levels(cars$manufacturer) <- list(AMC = "AMC Javelin",
                                       Cadillac = "Cadillac Fleetwood" ,
                                       Camaro = "Camaro Z28", 
                                       Datsun = "Datsun 710",
                                       Dodge = "Dodge Challenger",
                                       Duster = "Duster 360",
                                       Ferrari = "Ferrari Dino",
                                       Fiat = c("Fiat 128","Fiat X1-9"),
                                       Ford = "Ford Pantera L",
                                       Honda = "Honda Civic",
                                       Hornet = c("Hornet 4 Drive","Hornet Sportabout"),
                                       Lincoln = "Lincoln Continental",
                                       Chrysler = "Chrysler Imperial",
                                       Lotus = "Lotus Europa",
                                       Maserati = "Maserati Bora",
                                       Mazda = c("Mazda RX4","Mazda RX4 Wag"),
                                       Merc = c("Merc 230","Merc 240D","Merc 280",
                                                "Merc 280C","Merc 450SE","Merc 450SL",
                                                "Merc 450SLC"),
                                       Pontiac = "Pontiac Firebird",
                                       Porsche = "Porsche 914-2",
                                       Toyota = c("Toyota Corolla","Toyota Corona"),
                                       Valiant = "Valiant",
                                       Volvo = "Volvo 142E")
```

```{r}
###| Checking the data is correct
head(cars)
```

# A basic summary table

To create a summary table, we could manually calculate every entry and then create a dataset of these values and then create the table. However, it is much quicker (and easier) to use the `summarise()` function that comes with the `dplyr` package. 

:::{.callout-note}
The `dplyr` package will be installed/active if you have tidyverse. Alternatively, you can directly install and activate the package with the following commands:

```{r}
#| warning: false # will remove output messages

###| Package downloading and activating
# install.packages("dplyr")
library(dplyr)
```

If you are using RStudio, you can check which packages are installed using the bottom right window. 
:::

The following code gives some examples of how to extract various summary statistics using the `summarise()` command. 

```{r}
###| Basic summary table on continuous data
cars %>%
  summarise("Num. observations" = n(),
            "Mean mpg" = mean(mpg),
            "Range displacement" = max(disp)-min(disp),
            "Median horsepower" = median(hp))
```

Things get more complicated for categorical factors. We may need to use the `length()`, `which()` and `unique()` functions to find our desired statistics.

::: {.callout-note title="In a nutshell: counting TRUEs"}
`length(which(vs == 1))` counts the rows where the condition `vs == 1` is TRUE. A simpler equivalent is `sum(vs == 1)`, because R treats each TRUE as 1 and each FALSE as 0 when summing.
:::

```{r}
###| Basic summary table on categorical data
cars %>%
  summarise("vs 1" = length(which(vs == 1)),
            "am 0" = length(which(am == 0)),
            "Num. manufacturers" = length(unique(manufacturer))
  )
```

# Cylinder level exploration

Recall that we identified 3 distinct cylinder levels (check using `unique(cars$cyl)`). To do this, we will use the `group_by()` function. 

```{r}
###| Basic summary table grouped by cylinders
cars %>%
  group_by(cyl) %>%
  summarise("Num. observations" = n(),
            "Mean mpg" = mean(mpg),
            "Range displacement" = max(disp)-min(disp),
            "Median horsepower" = median(hp),
            "vs 1" = length(which(vs == 1)),
            "am 0" = length(which(am == 0)),
            "Num. manufacturers" = length(unique(manufacturer))
  )
```
:::{.callout-important}
It is important to note that this summary table had $3$ rows and not $1$! The statistics in each row correspond to the cars that had X number of cylinders. For example, of the $11$ cars with $4$ cylinders, the mean mpg was $26.6$, the displacement range $75.6$ etc... 
:::

When reading off this table, we can note a few things about the different types of cars:

:::{}
* Fewer cars had $6$ cylinders than $4$ or $8$. 
* As the number of cylinders increased, the mean mpg decreased, the displacement range increased, and the median horsepower increased. 
* No $8$ cylinder cars had `vs 1`. 
* There is no clear pattern in the `am 0` or manufacturer identifiable from this table. 

:::

# Making the tables more readable

Thus far, we have relied on the raw R output to read the tables. This is not ideal for reports or long periods of work. Instead, we can use the `kable()` function from the `kableExtra` package for neat tables. 

```{r}
###| Package
# install.packages("kableExtra")
library(kableExtra)

###| Grouped summary table
summ <- cars %>%
  group_by(cyl) %>%
  summarise("Num. observations" = n(),
            "Mean mpg" = mean(mpg),
            "Range displacement" = max(disp)-min(disp),
            "Median horsepower" = median(hp),
            "vs 1" = length(which(vs == 1)),
            "am 0" = length(which(am == 0)),
            "Num. manufacturers" = length(unique(manufacturer))
  )

###| Output
kable(summ)
```

We can continue to change various arguments of the `kable` function to get truly beautiful tables. An excellent site for this is [kableExtra](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html). A quick example of some commands are the following:

```{r}
###| Example kable extra
kable(summ, booktabs = TRUE,
      col.names = c("Cylinders","Total","Mean MPG","Displacement range",
                    "Median horsepower","Num vs 1","Num am 0",
                    "Num unique manufacturers")) %>%
  add_header_above(c(" " = 1, "Summary Statistics" = 7))
```

::: {.callout-tip title="Going further 🚀"}
Typing out one statistic per column quickly becomes unwieldy. `dplyr::across()` applies the same function(s) to many columns at once, and for polished, publication-ready tables `gt` and especially `gtsummary::tbl_summary()` build them with almost no code.

```{r}
#| eval: false
library(dplyr)
cars %>%
  group_by(cyl) %>%
  summarise(across(c(mpg, disp, hp), list(mean = mean, sd = sd)))

library(gtsummary)
tbl_summary(cars, by = cyl)   # publication-ready summary table
```
:::