---
title: "Introduction to ggplot2: The Grammar of Graphics"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

This section offers a primer on how the `ggplot2` package operates. Unlike many other plotting systems, `ggplot2` is based on the "Grammar of Graphics," which allows you to build graphs layer by layer. We will construct a complex visualisation starting from a blank canvas and adding elements—like data points, trend lines, and labels—one step at a time.

Let’s start with an example.

# 1. Getting the Data

The first step is always importing your data. For this page, we will use the **Palmer Penguins** dataset, which contains size measurements for three penguin species living on islands in the Palmer Archipelago, Antarctica.

```{r}
#| echo: true 
#| message: false
#| warning: false
library(ggplot2)
library(palmerpenguins)

penguins <- palmerpenguins::penguins
penguins <- na.omit(penguins)

head(penguins)

```

Using `head()` allows you to peek at the first few rows. This is an important habit to develop, as seeing the raw structure of your dataframe helps you spot potential relationships between variables like `flipper_length_mm` and `body_mass_g` before you even start plotting.

Next, let's create a new categorical variable to use later. We will classify penguins as "Heavy" or "Light" based on whether they weigh more than 4000 grams.

```{r}
# Create a weight category variable
penguins$weight_class <- ifelse(penguins$body_mass_g > 4000, "Heavy", "Light")

```

# 2. Building the Graph

To build a plot in `ggplot2`, you only *strictly* need the first two functions we will discuss. Every other function is an optional layer to enhance the visual.

The foundation is the `ggplot()` function. This initializes the graph by defining:

1. The **Dataset** to use.
2. The **Mapping** of variables to visual properties (aesthetics).

These mappings go inside the `aes()` function. Let's prepare to look at the relationship between **Flipper Length** (x-axis) and **Body Mass** (y-axis).

```{r}
# Specify dataset and mapping
library(ggplot2) 

ggplot(data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g))

```

**Why is the graph empty?** We told R *what* data to use and *where* to put it (the axes), but we haven't told it *how* to represent that data yet. We need a geometric object.

## 2.1 Geoms

"Geoms" (geometric objects) represent the actual marks on the plot: points, bars, lines, etc. We add these using functions starting with `geom_`. To create a scatterplot, we use `geom_point()`.

In `ggplot2`, we chain these layers together using the `+` operator.

```{r}
# Add geometric points
ggplot(data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point()

```

The plot above shows a strong positive correlation: as flipper length increases, body mass tends to increase.

We can customise the appearance of these geoms. Parameters like `color`, `size`, and `alpha` (transparency) can be set directly inside the geom function. Transparency is particularly useful when you have many points overlapping each other.

```{r}
# Make points blue, larger, and semi-transparent
ggplot(data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(color = "darkcyan", alpha = 0.6, size = 3)

```

## 2.2 Smoothers (Trend Lines)

To visualise the trend more clearly, we can add a smoothing layer using `geom_smooth()`. We can control the method (linear vs. curved), the colour, and whether to show the confidence interval. Here, we'll use a linear model (`method = "lm"`).

```{r}
#| echo: true 
#| message: false
#| warning: false
# Add a line of best fit
ggplot(data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(color = "darkcyan", alpha = 0.6, size = 3) + 
  geom_smooth(method = "lm")

```

## 2.3 Grouping with Aesthetics

While the plot above is informative, it treats all penguins as one homogeneous group. We know our dataset contains different **Species**.

We can map variables to visual characteristics like **colour** or **shape** inside the `aes()` function. This allows us to superimpose groups on a single graph. Let's map `species` to colour.

```{r}
#| echo: true 
#| message: false
#| warning: false
# Map species to color
ggplot(data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) + 
  geom_point(alpha = 0.6, size = 3) + 
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5)

```

Notice the difference:

1. `color = species` is inside `aes()` because it maps data to a visual.
2. `se = FALSE` was added to `geom_smooth` to remove the shaded confidence intervals for a cleaner look.

We can now see that Gentoo penguins are generally larger than Adelie or Chinstrap penguins.

## 2.4 Scales

"Scales" control how the data values are translated into visual properties (like specific colours or axis ticks). Scale functions always start with `scale_`.

Let's modify the X-axis breaks and manually define the colours for our species to be more distinct.

```{r}
#| echo: true 
#| message: false
#| warning: false
# Customize axes and color palette
ggplot(data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) + 
  geom_point(alpha = 0.6, size = 3) + 
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5) +
  scale_x_continuous(breaks = seq(170, 230, 10)) +
  scale_y_continuous(breaks = seq(2500, 6500, 1000)) +
  scale_color_manual(values = c("coral4", "purple", "cyan4"))

```

## 2.5 Facets

Sometimes a plot becomes too cluttered with multiple groups. **Faceting** solves this by splitting the plot into sub-plots based on a categorical variable.

Let's see if the relationship holds true across the different "Weight Classes" we defined earlier.

```{r}
#| echo: true 
#| message: false
#| warning: false
# Split the plot by weight class
ggplot(data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) + 
  geom_point(alpha = 0.6) + 
  geom_smooth(method = "lm", se = FALSE) + 
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) + 
  facet_wrap(~weight_class)

```

Faceting allows us to simultaneously view the data cut by different dimensions without overlapping points obscuring the view.

## 2.6 Labels

Good data visualisation requires clear communication. The `labs()` function allows you to customise the title, subtitle, captions, and axis labels so the viewer doesn't have to guess what "flipper_length_mm" means.

```{r}
# Add infomative labels
ggplot(data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) + 
  geom_point(alpha = 0.6) + 
  geom_smooth(method = "lm", se = FALSE) + 
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) + 
  facet_wrap(~weight_class) + 
  labs(
    title = "Penguin Size Characteristics",
    subtitle = "Comparison of Flipper Length vs Body Mass by Species",
    caption = "Source: Palmer Station LTER / palmerpenguins package",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Penguin Species"
  )

```

## 2.7 Themes

Finally, we can polish the overall look using **Themes**. Theme functions (starting with `theme_`) control non-data elements like background colours, fonts, and grid lines. `theme_minimal()` is a popular choice for a clean, modern look.

```{r}
#| echo: true 
#| message: false
#| warning: false
# Apply a minimal theme
ggplot(data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) + 
  geom_point(alpha = 0.6) + 
  geom_smooth(method = "lm", se = FALSE) + 
  scale_color_manual(values = c("darkorange", "purple", "cyan4")) + 
  facet_wrap(~weight_class) + 
  labs(
    title = "Penguin Size Characteristics",
    subtitle = "Comparison of Flipper Length vs Body Mass by Species",
    caption = "Source: Palmer Station LTER / palmerpenguins package",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Penguin Species"
  ) + 
  theme_minimal()

```

::: {.callout-tip}
You can manipulate the legend of your graph with the `theme` function. This can be added with `+` and is different from the `theme_minimal()` in the code above. For example if you want to adjust its' position, use `theme(legend.position = )`
Options for `legend.position`: 
*  `"top"`, `"bottom"`, `"left"`, `"right"` 
* `c(x, y)` for exact placement (e.g., `c(0.9, 0.2)` for bottom-right inside plot) 
* `"none"` to remove the legend
:::


# 3. Placing Mappings and Storing Objects

## 3.1 Global vs Local Mapping

In the previous examples, we defined our mapping `aes(...)` inside the main `ggplot()` function. This is a **Global** mapping, as in it applies to every layer (both points and lines).

However, you can also place mappings inside a specific Geom. This creates a **Local** mapping that applies *only* to that layer.

Observe the difference when we move `color = species` into the `geom_point()` function only:

```{r}
# Mapping color LOCALLY in geom_point only
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(aes(color = species), alpha = 0.6, size = 3) + 
  geom_smooth(method = "lm", se = FALSE, color = "black")

```

Because the color mapping was only inside `geom_point`, the `geom_smooth` did not "know" about the species groups. Consequently, it drew a single black trend line for the entire dataset, rather than one line per species.

## 3.2 Graphs as Objects

In R, a graph is just an object. You can save it to a variable, modify it later, and print it when you want to. This is great for keeping your code clean or generating multiple versions of a plot programmatically.

```{r}
# 1. Create the base plot object
p <- ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species))

# 2. Add a theme and labels to the object
pFinal <- p +
  theme_light() +
  labs(title = "Saved Graph Object")

# 3. Print the final result
pFinal

```

::: {.callout-tip title="Going further 🚀"}
Once you have saved a plot to an object, the `plotly` package can turn it interactive in a single line, giving you tooltips, zooming and panning for free.

```{r}
#| eval: false
library(plotly)
ggplotly(pFinal)
```

:::

# Further Resources

To deepen your understanding of `ggplot2`, consider these resources:

- [ggplot2 documentation](https://ggplot2.tidyverse.org)
- [R for Data Science](https://r4ds.hadley.nz/) - In particular the Data Visualisation chapter
- [The R Graph Gallery](https://r-graph-gallery.com/)