---
title: "Graph Customisation"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

# Introduction

When you first create plots in R, the default settings work well for exploring your data. However, when you're preparing visualisations for presentations, reports, or publications, you'll want to polish them to make your message clearer and more visually appealing.

In this tutorial, we'll learn how to customise various aspects of ggplot2 graphs including axes, colours, labels, legends, and themes. We'll primarily use the `palmerpenguins` dataset, which contains measurements of three penguin species from islands in Antarctica.

First, let's load our packages and data:

```{r}
#| echo: true
#| message: false
#| error: false

library(ggplot2)
library(palmerpenguins)
library(dplyr)
```

# Customising Axes

Axes help your readers understand the scale and range of your data. Let's explore how to modify them.

## Numeric Axes

You can control numeric axes using `scale_x_continuous()` and `scale_y_continuous()`. The most useful options are:

- `breaks`: Where to place tick marks and labels
- `limits`: The minimum and maximum values to display

```{r}
#| echo: true
#| warning: false
#| message: false

# Basic plot
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point() +
  scale_x_continuous(breaks = seq(30, 60, by = 5),
                     limits = c(30, 60)) +
  scale_y_continuous(breaks = seq(13, 22, by = 2),
                     limits = c(13, 22)) +
  labs(title = "Penguin Bill Dimensions",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)")
```

:::{.callout-tip}
The `seq()` function creates a sequence of numbers. For example, `seq(30, 60, by = 5)` creates: 30, 35, 40, 45, 50, 55, 60.
:::

## Formatting Numbers

The `scales` package provides helpful functions for formatting axis labels:

- `comma`: Adds commas to large numbers (1000 becomes 1,000)
- `dollar`: Adds currency symbols
- `percent`: Converts decimals to percentages

In the code chunk below, we create a random dataset by sampling from normal and uniform distributions for demonstrative purposes. Don't worry, the penguins will come back in the next example!

```{r}
# Create sample data
set.seed(42)
pop <- data.frame(
  city = paste("City", 1:40),
  population = rnorm(40, 250000, 80000),
  employment_rate = runif(40, 0.55, 0.85),
  median_income = rnorm(40, 65000, 15000)
)

ggplot(pop, aes(x = population, y = employment_rate)) +
  geom_point(color = "steelblue", size = 3, alpha = 0.6) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Population vs Employment Rate",
       x = "City Population",
       y = "Employment Rate")
```

## Categorical Axes

For categorical variables, use `scale_x_discrete()` or `scale_y_discrete()`. Returning to our penguins dataset:

```{r}
# Apply categorical variable scaling
ggplot(penguins, aes(x = species)) +
  geom_bar(fill = "coral") +
  scale_x_discrete(labels = c("Adélie\nPenguin", 
                              "Chinstrap\nPenguin", 
                              "Gentoo\nPenguin")) +
  labs(title = "Number of Penguins by Species",
       x = "Species",
       y = "Count")
```

:::{.callout-note}
`\n` creates a line break in text labels.
:::

## Date Axes

Working with dates requires `scale_x_date()` or `scale_y_date()`. We use the economics dataset for this:

```{r}
# Using the economics dataset
ggplot(economics, aes(x = date, y = unemploy / 1000)) +
  geom_line(color = "darkblue", linewidth = 1) +
  scale_x_date(date_breaks = "10 years",
               date_labels = "%Y") +
  labs(title = "US Unemployment Over Time",
       x = "Year",
       y = "Unemployed (thousands)")
```

:::{.callout-note}
We use `date_` for breaks and labels for dates.
:::

Common date format codes:

| Code | Meaning | Example |
|------|---------|---------|
| %d | Day of month | 01-31 |
| %m | Month number | 01-12 |
| %b | Abbreviated month | Jan |
| %B | Full month | January |
| %y | 2-digit year | 24 |
| %Y | 4-digit year | 2024 |

# Working with Colours

Colours can make your plots more engaging and help distinguish between groups.

## Setting Colours Manually

Use `color` for points and lines, and `fill` for bars and areas:

```{r}
#| echo: true
#| warning: false
#| message: false

# Single color
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(color = "darkgreen", size = 2) +
  labs(title = "Penguin Bill Measurements")
```

To assign specific colours to categories, use `scale_color_manual()` or `scale_fill_manual()`:

```{r}
#| echo: true
#| warning: false
#| message: false

ggplot(penguins, aes(x = species, fill = species)) +
  geom_bar() +
  scale_fill_manual(values = c("Adelie" = "darkorange",
                                "Chinstrap" = "purple",
                                "Gentoo" = "cyan4")) +
  labs(title = "Penguin Species Count",
       x = "Species",
       y = "Count",
       fill = "Species")
```

## Using Colour Palettes

Pre-designed colour palettes might be a starting point to inspire the final aesthetic you choose for your graph.

### ColorBrewer Palettes

ColorBrewer provides carefully designed colour schemes:

```{r}
#| echo: true
#| warning: false
#| message: false

ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Penguin Species by Island",
       x = "Species",
       y = "Count",
       fill = "Island")
```

### Viridis Palettes

Viridis palettes are *colourblind-friendly* and print well in greyscale:

```{r}
#| echo: true
#| warning: false
#| message: false

ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar() +
  scale_fill_viridis_d() +
  labs(title = "Penguin Species by Island",
       x = "Species",
       y = "Count",
       fill = "Island")
```

:::{.callout-note}
Use `_c` for continuous variables (numbers) and `_d` for discrete variables (categories).
:::

# Customising Points and Lines

## Point Shapes

Change point shapes with the `shape` parameter:

```{r}
#| echo: true
#| warning: false
#| message: false

ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, 
                     shape = species)) +
  geom_point(size = 3) +
  labs(title = "Bill Dimensions by Species",
       shape = "Species")
```

Shapes 21-25 allow both fill and border colours:

```{r}
#| echo: true
#| warning: false
#| message: false

ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, 
                     fill = species)) +
  geom_point(shape = 21, size = 3, color = "black") +
  labs(title = "Bill Dimensions by Species")
```

## Line Types

You can modify line appearance with `linetype`:

```{r}
ggplot(economics, aes(x = date)) +
  geom_line(aes(y = unemploy), linetype = "solid", color = "blue") +
  geom_line(aes(y = unemploy * 1.1), linetype = "dashed", color = "red") +
  labs(title = "Unemployment Trends",
       y = "Number Unemployed")
```

# Adding Annotations

Annotations can highlight important features in your data.

## Text Annotations

Use `annotate()` to add text anywhere on your plot:

```{r}
#| echo: true
#| warning: false
#| message: false

pengSumm <- penguins %>%
  group_by(species) %>%
  summarise(mean_mass = mean(body_mass_g, na.rm = TRUE))

ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot(fill = "lightblue") +
  annotate("text", x = 2, y = 6000, 
           label = "Gentoo penguins are\nthe heaviest species",
           color = "darkred", size = 4) +
  labs(title = "Body Mass Distribution by Species",
       x = "Species",
       y = "Body Mass (g)")
```

## Adding Reference Lines

Use `geom_hline()` and `geom_vline()` for horizontal and vertical lines:

```{r}
#| echo: true
#| warning: false
#| message: false

meanMass <- mean(penguins$body_mass_g, na.rm = TRUE)

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g,
                     color = species)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = meanMass,
             linetype = "dashed", color = "darkred") +
  annotate("text", x = 175, y = meanMass + 200,
           label = "Overall Mean", color = "darkred") +
  labs(title = "Penguin Measurements with Mean Body Mass",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)")
```

## Highlighting Specific Groups

The `gghighlight` package makes it easy to emphasise particular data:

```{r}
#| echo: true
#| warning: false
#| message: false

library(gghighlight)

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(size = 2, color = "coral") +
  gghighlight(species == "Gentoo") +
  labs(title = "Highlighting Gentoo Penguins",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)")
```

# Using Themes

Themes control the overall appearance of your plot (backgrounds, gridlines, fonts, etc.).

## Built-in Themes

ggplot2 includes several pre-made themes:

```{r}
#| echo: true
#| warning: false
#| message: false

p <- ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar() +
  labs(title = "Penguins by Species and Island")

# Try different themes
p + theme_minimal()  # Clean and simple
```

```{r}
#| warning: false
p + theme_classic()  # Traditional look
```

```{r}
#| warning: false
p + theme_dark()  # Dark background
```

## Custom Themes with ggthemes

The `ggthemes` package offers many professional themes:

```{r}
#| echo: true
#| warning: false
#| message: false

library(ggthemes)

base <- ggplot(penguins, aes(x = flipper_length_mm,
                                   y = body_mass_g,
                                   color = species)) +
  geom_point(size = 2) +
  labs(title = "Penguin Physical Measurements",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)")

# Economist style
base + theme_economist() + scale_color_economist()
```

```{r}
#| echo: true
#| warning: false
#| message: false

# FiveThirtyEight style
base + theme_fivethirtyeight()
```

# Combining Multiple Plots

The `patchwork` package lets you arrange multiple plots together:

```{r}
#| echo: true
#| warning: false
#| message: false

library(patchwork)

# Create three plots
p1 <- ggplot(penguins, aes(x = species, fill = species)) +
  geom_bar() +
  scale_fill_viridis_d() +
  labs(title = "Species Count") +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  labs(title = "Body Mass by Species",
       y = "Body Mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")

p3 <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, 
                            color = species)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_d() +
  labs(title = "Flipper Length vs Body Mass",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)") +
  theme_minimal()

# Combine plots: | means side-by-side, / means stacked
(p1 | p2) / p3 +
  plot_annotation(
    title = "Palmer Penguins: A Multi-faceted View",
    theme = theme(plot.title = element_text(size = 16, face = "bold"))
  )
```

The `&` operator applies settings to all plots:

```{r}
#| echo: true
#| warning: false
#| message: false

(p1 | p2) / p3 & theme_light()
```

::: {.callout-tip title="Going further 🚀"}
`patchwork` can do more than grids. Use `inset_element()` to drop a small plot inside a larger one (handy for a zoomed-in corner), and `plot_layout(guides = "collect")` to gather duplicate legends into one shared legend. The `cowplot` package is a popular alternative for arranging and aligning plots.

```{r}
#| eval: false
library(patchwork)
p3 + inset_element(p1, left = 0.6, bottom = 0.6, right = 1, top = 1)
(p1 | p2) / p3 + plot_layout(guides = "collect")

library(cowplot)
plot_grid(p1, p2, p3, labels = "AUTO")
```
:::

# Further Resources

- [ggplot2 documentation](https://ggplot2.tidyverse.org/)
- [R Graphics Cookbook](https://r-graphics.org/)
- [Data Visualization: A Practical Introduction](https://socviz.co/)
- [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/)