---
title: "Visualising Multiple Variables"
format: 
  html:
    fig-width: 10
    fig-height: 5
    code-fold: show
    code-tools: true
    toc: true
    code-copy: true
    number_sections: true
---

# Introduction

So far, you've learned to visualise single variables (like histograms) and relationships between two variables (like scatterplots). But real-world questions often involve multiple variables at once. For example: "How do penguin body characteristics vary by species, island, AND sex?"

This chapter covers two techniques for displaying multiple variables:

1. **Grouping**: Using colours, shapes, and sizes to distinguish groups within a single plot
2. **Faceting**: Creating separate mini-plots for different groups

Let's get started with our packages and data:

```{r}
#| echo: true 
#| message: false
#| warning: false

library(ggplot2)
library(palmerpenguins)
library(dplyr)

# View our data
glimpse(penguins)
```

# Grouping: Multiple Variables in One Plot

Grouping lets you add more variables to a plot by mapping them to visual properties like colour, shape, size, or transparency.

## Starting Simple

Let's begin with a basic relationship: flipper length vs. body mass.

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  labs(title = "Penguin Body Measurements",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)")
```

## Adding Colour to Show Groups

Now let's add species as a third variable using colour:

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(penguins, aes(x = flipper_length_mm, 
                     y = body_mass_g,
                     color = species)) +
  geom_point() +
  labs(title = "Penguin Measurements by Species",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)",
       color = "Species")
```

Each species now has its own colour, making patterns much clearer!

## Adding Shape for Another Variable

We can add a fourth variable (sex) using different point shapes:

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(penguins, aes(x = flipper_length_mm,
                     y = body_mass_g,
                     color = species,
                     shape = sex)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "Penguin Measurements by Species and Sex",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)",
       color = "Species",
       shape = "Sex")
```

:::{.callout-important}
* Mappings to variables go **inside** `aes()` (like `color = species`)
* Fixed values go **outside** `aes()` (like `size = 3`)
:::

## Creating Bubble Plots with Size

You can map a continuous variable to point size, creating a "bubble plot":

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(penguins, aes(x = flipper_length_mm,
                     y = body_mass_g,
                     color = species,
                     size = bill_length_mm)) +
  geom_point(alpha = 0.6) +
  labs(title = "Four Variables in One Plot",
       subtitle = "Size represents bill length",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)",
       color = "Species",
       size = "Bill Length (mm)")
```

This plot shows FOUR variables at once! But be careful, too many variables can make plots hard to read.

## Adding Trend Lines by Group

Let's explore the relationship between bill length and depth, with separate trend lines for each species:

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(penguins, aes(x = bill_length_mm,
                     y = bill_depth_mm,
                     color = species)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  labs(title = "Bill Dimensions with Species-Specific Trends",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)",
       color = "Species") +
  theme_minimal()
```

The `geom_smooth()` automatically creates separate lines for each colour group.

## A Polished Example

Here's a well-formatted plot using multiple customisation techniques:

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(penguins, aes(x = flipper_length_mm,
                     y = body_mass_g,
                     color = sex)) +
  geom_point(alpha = 0.4, size = 2.5) +
  geom_smooth(method = "lm", 
              formula = y ~ poly(x, 2),
              se = FALSE,
              linewidth = 1.3) +
  scale_color_brewer(palette = "Set1",
                     labels = c("Female", "Male", "Unknown")) +
  labs(title = "Body Mass vs. Flipper Length by Sex",
       subtitle = "Palmer Station, Antarctica (2007-2009)",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)",
       color = "Sex") +
  theme_minimal()
```

# Faceting: Small Multiples

Instead of cramming everything into one plot, faceting creates separate subplots for each group. This is often clearer than using many colours and shapes.

## Basic Faceting with `facet_wrap()`

Let's compare body mass distributions across species:

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(fill = "steelblue", bins = 20) +
  facet_wrap(~species, ncol = 1) +
  labs(title = "Body Mass Distribution by Species",
       x = "Body Mass (g)",
       y = "Count")
```

The `~species` notation means "create separate plots for each species." The `ncol = 1` arranges them in a single column.

## Faceting with Two Variables

Use `facet_grid()` to create a grid based on two variables:

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(penguins, aes(x = body_mass_g / 1000)) +
  geom_histogram(fill = "coral", bins = 15) +
  facet_grid(sex ~ species) +
  labs(title = "Body Mass by Species and Sex",
       x = "Body Mass (kg)",
       y = "Count")
```

The formula `sex ~ species` means rows are sex and columns are species.

## Combining Grouping and Faceting

You can use both techniques together for even more insight:

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(penguins, aes(x = bill_length_mm,
                     y = bill_depth_mm,
                     color = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~species, ncol = 1) +
  labs(title = "Bill Dimensions by Species and Sex",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)",
       color = "Sex")
```

Now we can see how the relationship between bill dimensions differs across species AND sex.

## Making Facets More Readable

Let's improve the previous plot with better formatting:

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(penguins, aes(x = bill_length_mm,
                     y = bill_depth_mm,
                     color = sex)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  facet_wrap(~species, ncol = 3) +
  scale_color_brewer(palette = "Set1",
                     labels = c("Female", "Male", "Unknown")) +
  theme_minimal() +
  labs(title = "Bill Dimensions Vary by Species and Sex",
       subtitle = "Measurements from Palmer Station, Antarctica",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)",
       color = "Sex") +
  theme(legend.position = "bottom")
```

## Faceting with Time Series

Faceting works great for comparing trends over time. Let's use the `economics` dataset to compare different economic indicators:

```{r}
#| echo: true 
#| message: false
#| warning: false

# Prepare data for multiple time series
econ <- economics %>%
  select(date, unemploy, psavert, pop) %>%
  tidyr::pivot_longer(cols = -date,
                      names_to = "variable",
                      values_to = "value")

# Create labels for facets
econLabels <- c(
  "unemploy" = "Unemployment (thousands)",
  "psavert" = "Personal Savings Rate (%)",
  "pop" = "Population (thousands)"
)

ggplot(econ, aes(x = date, y = value)) +
  geom_line(color = "darkblue") +
  facet_wrap(~variable,
             ncol = 1,
             scales = "free_y",
             labeller = labeller(variable = econLabels)) +
  theme_minimal() +
  labs(title = "US Economic Indicators Over Time",
       x = "Year",
       y = NULL) +
  theme(strip.text = element_text(face = "bold", size = 10))
```

The `scales = "free_y"` option lets each facet have its own y-axis scale, which is crucial when variables have different ranges.

## Many Facets: Small Multiples

When you have many groups, faceting creates a "small multiples" display. Here's an example with different islands:

```{r}
#| echo: true 
#| message: false
#| warning: false

# Compare trends across islands
ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = species), size = 2, alpha = 0.7) +
  facet_wrap(~island) +
  scale_color_viridis_d() +
  theme_minimal() +
  labs(title = "Bill Dimensions Across Different Islands",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)",
       color = "Species")
```

# Choosing Between Grouping and Faceting

**Use Grouping when:**
* You want to compare groups directly on the same axes
* You have 2-4 groups
* The differences between groups are subtle

**Use Faceting when:**
* You have many groups (5+)
* Comparisons are clearer with separate plots
* Groups have very different ranges or patterns
* You want to reduce visual clutter

**Use Both when:**
* You need to show complex relationships
* You have multiple categorical variables
* Your story requires showing differences at multiple levels

# Common Mistakes to Avoid

1. **Too many groups**: More than 5 colours becomes hard to distinguish
2. **Unclear legends**: Always label your aesthetic mappings clearly
3. **Overcrowding**: If your plot looks messy, try faceting instead
4. **Inconsistent scales**: Be careful with `scales = "free"` in facets, it can mislead
5. **Missing context**: Always include informative titles and labels

# Complete Example: Putting It All Together

Here's an example using multiple techniques:

```{r}
#| echo: true 
#| message: false
#| warning: false

# Create a multi-variable visualization
ggplot(penguins, aes(x = flipper_length_mm, 
                     y = body_mass_g,
                     color = species,
                     shape = sex)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(aes(group = species), 
              method = "lm",
              se = FALSE,
              linewidth = 1,
              show.legend = FALSE) +
  facet_wrap(~island) +
  scale_color_viridis_d(option = "plasma", end = 0.9) +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold", size = 11)) +
  labs(title = "Penguin Physical Characteristics Across Islands",
       subtitle = "Relationship between flipper length and body mass",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)",
       color = "Species",
       shape = "Sex",
       caption = "Data: Palmer Station LTER")
```

This plot shows FIVE variables: flipper length, body mass, species (colour + trend lines), sex (shape), and island (facets)!

::: {.callout-tip title="Going further 🚀"}
When one of your variables is time, the `gganimate` package can turn it into motion instead of yet another aesthetic. Adding `transition_states()` makes the plot step through each value (for example each year), animating how the relationship changes.

```{r}
#| eval: false
library(gganimate)
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, colour = species)) +
  geom_point() +
  transition_states(year)
```
:::

# Practice Exercises

Try creating these visualisations yourself:

1. Create a scatterplot of bill length vs. depth, coloured by island, faceted by species
2. Make a histogram of body mass, faceted by both species and sex using `facet_grid()`
3. Plot flipper length over time (year), with different colours for species
4. Create a grouped boxplot showing body mass by species, with separate boxes for each sex (Hint: use `interaction()` or `facet_wrap()`)
5. Challenge: Recreate the final "putting it all together" example with different variables (make it look nice and informative!).

# Further Resources

- [ggplot2 Book - Faceting](https://ggplot2-book.org/facet.html)
- [R Graphics Cookbook - Multiple Groups](https://r-graphics.org/)
- [Data Visualization: A Practical Introduction](https://socviz.co/)