---
title: "Specialised Visualisation Techniques"
format: 
  html:
    fig-width: 10
    fig-height: 6
    code-fold: show
    code-tools: true
    toc: true
    code-copy: true
    number_sections: true
---

# Introduction

This chapter covers specialised visualisation techniques that don't fit neatly into other categories but are incredibly useful for specific types of data and questions. Think of these as tools for your "advanced visualisation toolkit"—you won't use them every day, but when you need them, they're perfect for the job.

We'll explore:
- Visualising three-dimensional relationships
- Displaying flows and transitions
- Creating comparison matrices
- Specialised chart types for specific purposes

Let's load our packages:

```{r}
#| echo: true 
#| message: false
#| warning: false

library(ggplot2)
library(palmerpenguins)
library(dplyr)
```

# Bubble Charts: Three Variables on Two Axes

When you want to show three continuous variables simultaneously, bubble charts are an excellent choice. They're essentially scatterplots where point size represents a third variable.

## Basic Bubble Chart

Let's explore the relationship between penguin flipper length, body mass, and bill length:

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(penguins, aes(x = flipper_length_mm, 
                     y = body_mass_g,
                     size = bill_length_mm)) +
  geom_point(alpha = 0.6) +
  labs(title = "Penguin Physical Characteristics",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)",
       size = "Bill Length (mm)")
```

## Enhanced Bubble Chart

Let's improve this with better styling and add a fourth variable (species) using colour:

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(penguins, aes(x = flipper_length_mm,
                     y = body_mass_g,
                     size = bill_length_mm,
                     color = species)) +
  geom_point(alpha = 0.6, shape = 16) +
  scale_size_continuous(range = c(2, 12)) +
  scale_color_viridis_d() +
  labs(title = "Four Variables in One Visualization",
       subtitle = "Bubble size represents bill length",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)",
       size = "Bill Length (mm)",
       color = "Species") +
  theme_minimal()
```

The `scale_size_continuous(range = c(2, 12))` controls the minimum and maximum bubble sizes.

:::{.callout-note}
Bubble charts are controversial because humans are better at judging length than area. Use them when the approximate relationship is more important than exact values.
:::

# Correlation Heatmaps

Heatmaps display data values as coloured cells, making patterns immediately visible. They're particularly useful for comparing many variables at once.

## Creating a Correlation Matrix

Let's visualise correlations between penguin measurements:

```{r}
#| echo: true 
#| message: false
#| warning: false

# Prepare data - select numeric columns only
pnum <- penguins %>%
  select(bill_length_mm, bill_depth_mm, 
         flipper_length_mm, body_mass_g) %>%
  na.omit()

# Calculate correlation matrix
corMat <- cor(pnum)

# Convert to long format for ggplot
library(tidyr)
corLong <- corMat %>%
  as.data.frame() %>%
  mutate(var1 = rownames(.)) %>%
  pivot_longer(cols = -var1, 
               names_to = "var2", 
               values_to = "correlation")

# Create heatmap
ggplot(corLong, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(correlation, 2)), 
            color = "black", size = 4) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation Matrix of Penguin Measurements",
       x = "", y = "", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Strong positive correlations appear in red, negative correlations in blue, and weak correlations in white.

# Scatterplot Matrices

A scatterplot matrix shows relationships between multiple variables in a grid format. It's like seeing all possible pairs of scatterplots at once.

```{r}
#| echo: true 
#| message: false
#| warning: false

library(GGally)

# Select variables for the matrix
sub <- penguins %>%
  select(bill_length_mm, bill_depth_mm, 
         flipper_length_mm, body_mass_g, species) %>%
  na.omit()

# Create scatterplot matrix
ggpairs(sub, 
        aes(color = species, alpha = 0.5),
        columns = 1:4) +
  labs(title = "Relationships Between Penguin Measurements") +
  theme_bw()
```

**Reading the matrix**:

* **Diagonal**: Distribution of each variable
* **Lower triangle**: Scatterplots (row variable on y-axis, column variable on x-axis)
* **Upper triangle**: Correlation coefficients

## Custom Scatterplot Matrix

You can customise the appearance with your own functions:

```{r}
#| echo: true 
#| message: false
#| warning: false

# Custom function for density plots
myDensity <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) + 
    geom_density(alpha = 0.6, fill = "steelblue")
}

# Custom function for scatterplots
myScatter <- function(data, mapping, ...) {
  ggplot(data = data, mapping = mapping) + 
    geom_point(alpha = 0.4, color = "steelblue", size = 1.5) + 
    geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 0.8)
}

# Create customized matrix
pnumOnly <- penguins %>%
  select(bill_length_mm, bill_depth_mm, 
         flipper_length_mm, body_mass_g) %>%
  na.omit()

ggpairs(pnumOnly,
        lower = list(continuous = myScatter),
        diag = list(continuous = myDensity)) +
  labs(title = "Penguin Measurements: Detailed Relationships") +
  theme_minimal()
```

# Alluvial Diagrams: Visualizing Flows

Alluvial diagrams (also called Sankey diagrams) show how categorical variables relate to each other through flowing connections. They're perfect for showing composition changes across groups.

## Basic Alluvial Diagram

Let's examine how penguins are distributed across species, islands, and sex:

```{r}
#| echo: true 
#| message: false
#| warning: false

library(ggalluvial)

# Prepare data
flows <- penguins %>%
  filter(!is.na(sex)) %>%
  group_by(species, island, sex) %>%
  summarise(count = n(), .groups = "drop")

# Create alluvial diagram
ggplot(flows,
       aes(axis1 = species,
           axis2 = island,
           axis3 = sex,
           y = count)) +
  geom_alluvium(aes(fill = species), alpha = 0.7) +
  geom_stratum(alpha = 0.8) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Species", "Island", "Sex"),
                   expand = c(0.1, 0.1)) +
  scale_fill_viridis_d() +
  labs(title = "Penguin Distribution Across Categories",
       y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

**How to read this**:

* Start on the left with species
* Follow the flows to see how each species is distributed across islands
* Continue following to see the sex distribution
* Flow width represents the number of penguins

## Enhanced Alluvial Diagram

Let's create a more detailed version with better styling:

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(flows,
       aes(axis1 = species,
           axis2 = island,
           axis3 = sex,
           y = count)) +
  geom_alluvium(aes(fill = species), 
                color = "black", 
                alpha = 0.8,
                curve_type = "sigmoid") +
  geom_stratum(alpha = 0.9, color = "white", linewidth = 1) +
  geom_text(stat = "stratum", 
            aes(label = after_stat(stratum)),
            size = 4, fontface = "bold") +
  scale_x_discrete(limits = c("Species", "Island", "Sex"),
                   expand = c(0.15, 0.15)) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "How Penguins are Distributed",
       subtitle = "Following the flow from species through island to sex",
       y = "Number of Penguins",
       fill = "Species") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 12, face = "bold"))
```

# Waterfall Charts: Showing Cumulative Effects

Waterfall charts illustrate the cumulative effect of sequential positive and negative values. They're commonly used in finance but work for any cumulative process.

```{r}
#| echo: true 
#| message: false
#| warning: false

library(waterfalls)

# Create sample budget data
budget <- data.frame(
  category = c("Research Grant", "Equipment", "Software", 
               "Lab Supplies", "Travel", "Conference Fees"),
  amount = c(50000, -15000, -3000, -8000, -5000, -2000)
)

# Create basic waterfall chart
waterfall(budget) +
  labs(title = "Research Budget Breakdown",
       x = "", y = "Amount ($)") +
  theme_minimal()
```

## Waterfall Chart with Net Total

Let's add a total column to show the final balance:

```{r}
#| echo: true 
#| message: false
#| warning: false

waterfall(budget,
          calc_total = TRUE,
          total_axis_text = "Remaining",
          total_rect_text_color = "black",
          total_rect_color = "gold") +
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "Research Project Financial Flow",
       subtitle = "Tracking income and expenses",
       x = "", y = "") +
  theme_minimal()
```

This shows clearly how the initial grant is allocated across different expenses.

# Advanced Heatmap: Time Series

Heatmaps can be good for visualising changes over time. Let's create one using economic data:

```{r}
#| echo: true 
#| message: false
#| warning: false

# Prepare time series data
econMonth <- economics %>%
  mutate(year = lubridate::year(date),
         month = lubridate::month(date, label = TRUE)) %>%
  filter(year >= 2000, year <= 2015) %>%
  select(year, month, unemploy) %>%
  mutate(unemploy_scaled = scale(unemploy)[,1])

# Create heatmap
ggplot(econMonth, aes(x = month, y = factor(year), fill = unemploy_scaled)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(low = "darkgreen", mid = "white", high = "darkred",
                       midpoint = 0,
                       name = "Unemployment\n(Scaled)") +
  labs(title = "US Unemployment Patterns by Month and Year",
       subtitle = "Red indicates higher unemployment, green indicates lower",
       x = "Month", y = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

This reveals seasonal patterns and trends across years.

# Choosing the Right Specialized Chart

| Chart Type | Best For | Avoid When |
|------------|----------|------------|
| **Bubble Chart** | 3-4 continuous variables | Need precise values |
| **Heatmap** | Many variables, patterns in matrices | Few data points |
| **Scatterplot Matrix** | Exploring relationships between 3-6 variables | More than 7 variables |
| **Alluvial Diagram** | Categorical flows, composition changes | Simple counts are sufficient |
| **Waterfall Chart** | Cumulative effects, sequential changes | All values are similar |


# Common Mistakes

1. **Overcomplicating**: Don't use specialised charts when simple ones work better
2. **Too many variables**: Even specialised charts have limits (typically 4-8 variables)
3. **Poor scaling**: Always consider rescaling variables for fair comparisons
4. **Missing context**: These charts often need extra explanation, so do add subtitles and annotations
5. **Wrong tool**: Make sure your chart type matches your data structure

The key is knowing when to use each technique. Start simple, and only use these specialised charts when they truly add insight that simpler visualisations can't provide.

# Practice Exercises

1. Create a bubble chart with penguin data using bill dimensions and body mass
2. Make a correlation heatmap for the `economics` dataset numeric variables
3. Build an alluvial diagram showing penguin distribution across all three categorical variables
4. Create a scatterplot matrix coloured by penguin sex
5. Make a waterfall chart showing hypothetical monthly savings (mix of income and expenses)

# Further Resources

- [ggplot2 Extensions Gallery](https://exts.ggplot2.tidyverse.org/)
- [R Graph Gallery](https://r-graph-gallery.com/)
- [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/) by Claus Wilke