---
title: "Visualising Time"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

A graph is a powerful vehicle for displaying change. While standard bar charts compare static categories, time-dependent graphs capture the *motion* of data.

In this chapter, we move beyond simple line charts to explore more sophisticated ways to visualise temporal change.

# Time Series

A time series is a sequence of observations indexed in time order.

## Simple Time Series

::: {.callout-tip title="Math to Code Mindset"}

* **Input:** Time $t$ and value $y$
* **Transformation:** Sort by $t$
* **Output:** Line segments connecting consecutive points

:::

```{r}
#| echo: true 
#| message: false
#| warning: false

library(ggplot2)
library(scales)

ggplot(economics, aes(x = date, y = psavert)) +
geom_line(color = "indianred3", linewidth = 1) +
geom_smooth(method = "loess", se = FALSE) +
scale_x_date(date_breaks = "5 years", labels = date_format("%b-%y")) +
labs(
title = "Personal Savings Rate (1967–2015)",
y = "Savings Rate",
x = ""
) +
theme_minimal()
```

:::{.callout-important}
Dates must be stored as `Date` or `POSIXct` objects.
Text values are sorted alphabetically, not chronologically.
:::

## Multivariate Time Series

```{r}
#| echo: true 
#| message: false
#| warning: false

library(quantmod)
library(dplyr)

getSymbols(c("AAPL", "META"), from = "2023-01-01", to = "2023-07-31")
```

::: {.callout-note title="In a nutshell: getSymbols works by side effect"}
`quantmod::getSymbols("AAPL")` fetches live data from the internet and, rather than returning it, quietly creates an object named `AAPL` for you (notice there is no `<-` assignment). This means it needs an internet connection to run, and because the data is live the results may differ slightly each time you run it.
:::

```{r}
#| echo: true
#| message: false
#| warning: false


mts <- bind_rows(
data.frame(Date = index(AAPL), Close = as.numeric(AAPL$AAPL.Close), Company = "Apple"),
data.frame(Date = index(META), Close = as.numeric(META$META.Close), Company = "Meta")
)

ggplot(mts, aes(x = Date, y = Close, color = Company)) +
geom_line(linewidth = 1) +
scale_color_brewer(palette = "Dark2") +
labs(
title = "Tech Stock Performance",
subtitle = "Apple vs Meta (Jan–July 2023)",
y = "Closing Price ($)",
x = ""
) +
theme_minimal()
```

# Dumbbell Charts

A dumbbell chart visualises **change between two time points**.

::: {.callout-tip title="Math to Code Mindset"}

* **Input:** Category and two values
* **Transformation:** Pivot to wide format
* **Output:** A segment plus two points

:::

```{r}
#| echo: true 
#| message: false
#| warning: false

library(tidyr)
data(gapminder, package = "gapminder")

pdat <- gapminder %>%
filter(continent == "Americas", year %in% c(1952, 2007)) %>%
select(country, year, lifeExp) %>%
pivot_wider(names_from = year, values_from = lifeExp, names_prefix = "y")

ggplot(pdat, aes(y = reorder(country, y1952))) +
geom_segment(
aes(x = y1952, xend = y2007, yend = reorder(country, y1952)),
color = "grey70",
linewidth = 1
) +
geom_point(aes(x = y1952), color = "red", size = 3) +
geom_point(aes(x = y2007), color = "blue", size = 3) +
labs(
title = "Change in Life Expectancy (1952 vs 2007)",
subtitle = "Red = 1952, Blue = 2007",
x = "Life Expectancy (Years)",
y = ""
) +
theme_minimal()
```

# Slope Graphs (ggplot-only)

Slope graphs track **relative position across discrete time points**.

Key idea:

> Fix the *x-position* to time, and the *y-position* to **rank**.

```{r}
#| echo: true 
#| message: false
#| warning: false

library(forcats)

slope <- gapminder %>%
filter(
year %in% c(1992, 1997, 2002, 2007),
country %in% c(
"Panama", "Costa Rica", "Nicaragua",
"Honduras", "El Salvador", "Guatemala", "Belize"
)
) %>%
group_by(year) %>%
mutate(rank = rank(-lifeExp, ties.method = "first")) %>%
ungroup() %>%
mutate(year = factor(year))

ggplot(slope, aes(x = year, y = rank, group = country)) +
geom_line(color = "grey50", linewidth = 1) +
geom_point(size = 3) +
scale_y_reverse(breaks = 1:7) +
labs(
title = "Life Expectancy Rankings",
subtitle = "Central America (1992–2007)",
x = "Year",
y = "Rank (Higher = Better)"
) +
theme_minimal()
```

::: {.callout-note}
Crossing lines indicate **rank reversals** — for example, Guatemala overtakes Honduras in the early 2000s.
:::

# Area Charts

Area charts emphasise **magnitude** by filling the area beneath a line.

## Stacked Area Charts

```{r}
#| echo: true 
#| message: false
#| warning: false

data(uspopage, package = "gcookbook")

ggplot(
uspopage,
aes(x = Year, y = Thousands / 1000, fill = forcats::fct_rev(AgeGroup))
) +
geom_area(color = "black", linewidth = 0.2) +
scale_fill_brewer(palette = "Set2") +
labs(
title = "US Population Composition",
y = "Population (Millions)",
fill = "Age Group",
x = ""
) +
theme_minimal()
```


# Stream Graphs

::: {.callout-note title="In a nutshell: this section is advanced/optional"}
Stream graphs are a more specialised technique. If you are still finding your feet, feel free to skip this section, the earlier charts cover everything you need, and you can return here later.
:::

A stream graph is a **centred stacked area chart**.
Instead of stacking values from zero upward, the entire stack is **recentred around a moving baseline**, reducing visual bias caused by accumulation from the bottom.

## From Stacked Areas to Streams

::: {.callout-tip title="Math to Code Mindset"}

* **Input:** Time $t$, group $g$, value $y_{tg}$
* **Normalise:** $p_{tg} = y_{tg} / \sum_g y_{tg}$
* **Stack:** cumulative sums within each $t$
* **Recentre:** shift stack midpoint to zero
* **Output:** Filled ribbons around a central baseline

:::

---

## Data Transformation

```{r}
#| echo: true 
#| message: false
#| warning: false

library(dplyr)

stream <- uspopage %>%
group_by(Year) %>%
mutate(
prop = Thousands / sum(Thousands),
ymax = cumsum(prop),
ymin = ymax - prop
) %>%
ungroup() %>%
group_by(Year) %>%
mutate(
center = (max(ymax) + min(ymin)) / 2,
ymin = ymin - center,
ymax = ymax - center
)
```
### What this does mathematically

For each year:

1. Convert population counts to **proportions**
2. Stack age groups using cumulative sums
3. Compute the vertical midpoint of the stack
4. Shift the entire stack so its midpoint is zero

This removes the fixed baseline that causes distortion in stacked area charts.

---

## Stream Graph Rendering 

```{r}
#| echo: true 
#| message: false
#| warning: false

ggplot(stream) +
geom_ribbon(
aes(
x = Year,
ymin = ymin,
ymax = ymax,
fill = forcats::fct_rev(AgeGroup)
),
color = "black",
linewidth = 0.15
) +
scale_fill_brewer(palette = "Set2") +
labs(
title = "US Population Demographics",
subtitle = "Stream graph constructed from centered stacked areas",
y = "Centered Proportion",
x = "Year",
fill = "Age Group"
) +
theme_minimal()
```
---

::: {.callout-note}

A stream graph is **not a new chart type** — it is a stacked area chart with a **moving, centred baseline**.

:::

The widening of upper bands toward the right visualises the ageing of the US population.



