---
title: "Univariate Data"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

A sensible starting point for any data analysis is to look at each important variable on its own. Univariate graphs do exactly this: they show the distribution of a single variable at a time. That variable might be **categorical** (for example diet, taxonomic order, or conservation status) or **quantitative** (such as sleep duration, body weight, or brain weight).

Throughout this chapter we use the `msleep` dataset, which contains sleep times and weights for 83 different mammals. We focus on three variables: the diet (`vore`) and taxonomic order (`order`) of the animals, and the total amount of sleep they get (`sleep_total`).

---

## Categorical variables

Both the diet of the mammals and their taxonomic order are categorical variables. The distribution of a single categorical variable is most commonly shown using a bar chart. Other options include pie charts, and—less frequently—tree maps or waffle charts.

### Bar charts

Below, we plot a simple bar chart to display the distribution of mammals by diet (`vore`).

```{r}
#| echo: true
#| warning: false
#| message: false

library(ggplot2)
data(msleep, package = "ggplot2")

# plot the distribution of diet (vore)
# We filter out NAs for a cleaner plot
library(dplyr)
mclean <- msleep %>% filter(!is.na(vore))

ggplot(mclean, aes(x = vore)) +
  geom_bar()

```

From this plot we can immediately see that most mammals in this dataset are herbivores, followed closely by omnivores and carnivores, with insectivores being the smallest group.

---

#### Percentages

Instead of showing raw counts, bars can also represent percentages. In a bar chart, `aes(x = vore)` is shorthand for `aes(x = vore, y = after_stat(count))`, where `count` is the frequency in each category. By working with this quantity explicitly, we can convert counts into percentages.

::: {.callout-note title="In a nutshell: what after_stat means"}
`after_stat()` tells `ggplot2` to compute the bar heights itself from the data (here, by counting how many mammals fall in each category), rather than you supplying a `y` value directly.
:::

```{r}
# plot the distribution as percentages
ggplot(mclean,
       aes(x = vore, y = after_stat(count/sum(count)))) +
  geom_bar() +
  labs(x = "Diet",
       y = "Percent",
       title  = "Mammals by diet") +
  scale_y_continuous(labels = scales::percent)

```

Here, the `scales` package is used to format the y-axis labels as percentages.

---

#### Sorting categories

It is often helpful to order categories by their frequency. To do this, we first calculate the counts explicitly and then reorder the categories using `reorder()`. Setting `stat = "identity"` tells `ggplot2` not to compute counts internally.

```{r}
# calculate number of mammals in each diet category
pdat <- mclean %>%
  count(vore)

```

The resulting dataset is shown below:

| vore | n |
| --- | --- |
| carni | 19 |
| herbi | 32 |
| insecti | 5 |
| omni | 20 |

We can now use this dataset to create a sorted bar chart.

```{r}
# plot the bars in ascending order
ggplot(pdat,
       aes(x = reorder(vore, n), y = n)) +
  geom_bar(stat="identity") +
  labs(x = "Diet",
       y = "Frequency",
       title  = "Mammals by diet")

```

The bars are ordered from smallest to largest frequency. Using `reorder(vore, -n)` would reverse this order.

---

#### Labelling bars

Sometimes it is useful to display the numerical values directly on the bars.

```{r}
# plot the bars with numeric labels
ggplot(pdat,
       aes(x = vore, y = n)) +
  geom_bar(stat="identity") +
  geom_text(aes(label = n), vjust=-0.5) +
  labs(x = "Diet",
       y = "Frequency",
       title  = "Mammals by diet")

```

Here, `geom_text()` adds the labels, and `vjust` controls their vertical positioning.

Putting these ideas together, we can create a bar chart showing percentages, ordered from most to least frequent, and labelled accordingly.

```{r}
#| echo: true
#| warning: false
#| message: false

library(scales)
pdat <- mclean %>%
  count(vore) %>%
  mutate(pct = n / sum(n),
         pctlabel = paste0(round(pct*100), "%"))

# plot the bars as percentages,
# in descending order with bar labels
ggplot(pdat,
       aes(x = reorder(vore, -pct), y = pct)) +
  geom_bar(stat="identity", fill="indianred3", color="black") +
  geom_text(aes(label = pctlabel), vjust=-0.25) +
  scale_y_continuous(labels = percent) +
  labs(x = "Diet",
       y = "Percent",
       title  = "Mammals by diet")

```

---

#### Overlapping labels

When there are many categories or when labels are long, axis labels can overlap. This is clear when plotting the taxonomic order of the mammals.

```{r}
# basic bar chart with overlapping labels
ggplot(msleep, aes(x=order)) +
  geom_bar() +
  labs(x = "Taxonomic Order",
       y = "Frequency",
       title = "Mammals by Order")

```

One solution is to flip the axes.

```{r}
# horizontal bar chart
ggplot(msleep, aes(x = order)) +
  geom_bar() +
  labs(x = "",
       y = "Frequency",
       title = "Mammals by Order") +
  coord_flip()

```

Alternatively, labels can be rotated.

```{r}
# bar chart with rotated labels
ggplot(msleep, aes(x=order)) +
  geom_bar() +
  labs(x = "",
       y = "Frequency",
       title = "Mammals by Order") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

A third option is to stagger the labels by inserting line breaks.

```{r}
# bar chart with staggered labels
lbls <- paste0(c("","\n"), unique(msleep$order))
# Note: In real usage, ensure labels match factor levels order
ggplot(msleep,
       aes(x=factor(order, labels = lbls))) +
  geom_bar() +
  labs(x = "",
       y = "Frequency",
       title = "Mammals by Order")

```

In practice, it is usually best to avoid rotated labels where possible, as they are harder to read and increase cognitive load for the reader.

---

### Pie charts

Pie charts are often debated in statistics. For comparing category frequencies, bar charts are generally superior because humans judge lengths more accurately than areas. However, if the goal is to show how each category relates to the whole, and there are only a few categories, pie charts can still be useful.

A pie chart in `ggplot2` is essentially a bar chart plotted on a polar coordinate system.

```{r}
# Prepare data for pie chart
pdat <- mclean %>%
  count(vore) %>%
  mutate(prop = n / sum(n)) %>%
  mutate(label = scales::percent(prop))

# Create pie chart
ggplot(pdat, aes(x = "", y = prop, fill = vore)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(title = "Mammals by diet")

```

We use `coord_polar("y")` to transform the bar chart into a circle. `theme_void()` is used to remove the background, axis grid, and labels that are not necessary for a pie chart.

To place labels inside the slices, we can use `geom_text()` with `position_stack()`.

```{r}
# Create pie chart with labels
ggplot(pdat, aes(x = "", y = prop, fill = vore)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5),
            color = "white",
            size = 4) +
  theme_void() +
  labs(title = "Mammals by diet") +
  theme(legend.position = "right")

```

This plot makes it clear, for example, that roughly 42% of the mammals in this dataset are herbivores.

---

### Tree maps

Tree maps provide another alternative to pie charts and are particularly useful when there are many categories, such as the taxonomic orders in our dataset.

```{r}
#| echo: true
#| warning: false
#| message: false

library(treemapify)

# create a treemap of mammal orders
pdat <- msleep %>%
  count(order)

ggplot(pdat,
       aes(fill = order, area = n)) +
  geom_treemap() +
  labs(title = "Mammals by Order")

```

Adding labels makes the plot more informative.

```{r}
# create a treemap with tile labels
ggplot(pdat,
       aes(fill = order, area = n, label = order)) +
  geom_treemap() +
  geom_treemap_text(colour = "white", place = "centre") +
  labs(title = "Mammals by Order") +
  theme(legend.position = "none")

```

---

### Waffle charts

A waffle chart (also called a grid plot or square pie chart) represents proportions using a grid of squares. While dedicated packages exist, we can create a waffle chart in standard `ggplot2` by treating the data as a grid of tiles (dedicated packages sometimes get removed so it's better to have a more sustainable way to make graphs that don't rely on external packages).

We begin by sorting the data and assigning logical `x` and `y` coordinates to each mammal to create a 10-column grid.

```{r}
# Prepare data: Sort by diet to cluster colors
waffle <- mclean %>%
  arrange(vore) %>%
  mutate(
    # Create x and y coordinates for a 10-column grid
    x = (row_number() - 1) %% 10,
    y = (row_number() - 1) %/% 10
  )

# Create the chart using geom_tile
ggplot(waffle, aes(x = x, y = y, fill = vore)) +
  geom_tile(color = "white") +  # white borders create the "grid" look
  coord_fixed() +               # ensures squares remain square
  theme_void() +                # removes axes and backgrounds
  labs(title = "Mammals by Diet",
       caption = "1 square = 1 mammal")

```

This chart displays every individual mammal as a single square, making it easy to visualise the sample size alongside the proportions.

---

## Quantitative variables

In the `msleep` dataset, `sleep_total` (total sleep in hours) is a quantitative variable. Common ways to visualise the distribution of a single quantitative variable include histograms, kernel density plots, and dot plots.

### Histograms

Histograms are the most widely used tool for visualising quantitative data. The range of values is divided into adjacent bins of equal width, and the number of observations in each bin is shown as a bar.

```{r}
#| echo: true
#| warning: false
#| message: false

library(ggplot2)

# plot the sleep distribution using a histogram
ggplot(msleep, aes(x = sleep_total)) +
  geom_histogram() +
  labs(title = "Mammal sleep distribution",
       x = "Total Sleep (hours)")

```

The distribution appears somewhat bimodal, with many mammals sleeping around 10 hours, but a significant number sleeping much less or much more.

---

#### Bins and bandwidths

The choice of the number of bins (or equivalently the bin width) has a large impact on the appearance of a histogram. The default is 30 bins, but it is often useful to experiment.

```{r}
# plot the histogram with 15 bins
ggplot(msleep, aes(x = sleep_total)) +
  geom_histogram(fill = "cornflowerblue",
                 color = "white",
                 bins = 15) +
  labs(title="Mammal sleep distribution",
       subtitle = "number of bins = 15",
       x = "Total Sleep (hours)")

```

Alternatively, you can specify the bin width directly.

```{r}
# plot the histogram with a binwidth of 2 hours
ggplot(msleep, aes(x = sleep_total)) +
  geom_histogram(fill = "cornflowerblue",
                 color = "white",
                 binwidth = 2) +
  labs(title="Mammal sleep distribution",
       subtitle = "binwidth = 2 hours",
       x = "Total Sleep (hours)")

```

As before, the y-axis can show counts or percentages.

```{r}
library(scales)
ggplot(msleep,
       aes(x = sleep_total, y= after_stat(count/sum(count)))) +
  geom_histogram(fill = "cornflowerblue",
                 color = "white",
                 binwidth = 2) +
  labs(title="Mammal sleep distribution",
       y = "Percent",
       x = "Total Sleep (hours)") +
  scale_y_continuous(labels = percent)

```

---

### Kernel density plots

A kernel density plot provides a smooth alternative to a histogram. The idea is to estimate the underlying probability density function of a continuous variable, producing a smooth curve whose total area equals one.

```{r}
# Create a kernel density plot of sleep
ggplot(msleep, aes(x = sleep_total)) +
  geom_density() +
  labs(title = "Mammal sleep distribution")

```

The area under the curve between two time points represents the proportion of mammals in that sleep range.

```{r}
# Create a kernel density plot of sleep
ggplot(msleep, aes(x = sleep_total)) +
  geom_density(fill = "indianred3") +
  labs(title = "Mammal sleep distribution")

```


---

#### Smoothing parameter

The amount of smoothing is controlled by the bandwidth parameter `bw`. Larger values produce smoother curves, while smaller values reveal more detail.

```{r}
# default bandwidth for the sleep variable
bw.nrd0(msleep$sleep_total)

```

```{r}
# Create a kernel density plot of sleep
ggplot(msleep, aes(x = sleep_total)) +
  geom_density(fill = "deepskyblue",
               bw = 1) +
  labs(title = "Mammal sleep distribution",
       subtitle = "bandwidth = 1")

```

Kernel density plots are useful for identifying where observations are concentrated, although the interpretation of the y-axis can be less intuitive for non-statisticians, so do keep this in mind.

---

### Dot plots

A dot plot is another alternative to the histogram. The data are grouped into bins, but each individual observation is shown as a dot. This approach works best for relatively small datasets like `msleep`.

```{r}
#| echo: true
#| warning: false
#| message: false

# plot the sleep distribution using a dotplot
ggplot(msleep, aes(x = sleep_total)) +
  geom_dotplot() +
  labs(title = "Mammal sleep distribution",
       y = "Proportion",
       x = "Total Sleep (hours)")

```

Many additional options are available; see `?geom_dotplot` for further details. 

These are just a couple of ways you could visualise one variable, do explore to find more!

::: {.callout-tip title="Going further 🚀"}
Two modern packages give you slicker distribution displays. `ggbeeswarm` spreads overlapping points sideways so none are hidden, and `ggridges` draws a density curve per group, stacked into a ridgeline plot.

```{r}
#| eval: false
library(ggbeeswarm)
ggplot(msleep, aes(x = vore, y = sleep_total)) +
  geom_beeswarm()

library(ggridges)
ggplot(msleep, aes(x = sleep_total, y = vore)) +
  geom_density_ridges()
```
:::

# Further Resources

- [R for Data Science](https://r4ds.hadley.nz/)
- [ggplot2 documentation](https://ggplot2.tidyverse.org/)