---
title: "Fitting a model with categorical data"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

Thus far, we have fit models with continuous data. However, we will often encounter variables that have a finite number of categories (i.e. categorical variables). We can either assign a reference category (treatment coding) or have no intercept term.

# Set up

To explore categorical data, let us look at the agriculture dataset.

```{r}
###| Load in the data
agriculture <- read.csv("../Data/agriculture.csv")

###| First 6 rows
head(agriculture)
```

In this dataset, there are 4 variables: Variety, Block, SIZE and YIELD. From the first 6 rows, we see that Variety is a categorical variable with categories at least A-F. However, as we can only see 1 for Block, it is unclear what type of variable it is. Instead, let us look at the summary of the dataset, using the `summary()` function.

```{r}
###| Dataset summary
summary(agriculture)
```

From this, we see that SIZE and YIELD are both continuous variables (the summary reports their minimum, median, mean and maximum). We also see that Block takes values from 1 to 8 and Variety is always a character. However, we do not know what values Variety takes.

To fix this, we will alter the data using the `factor()` command.

```{r}
###| Factor command
agriculture$Variety <- factor(agriculture$Variety)

###| First 6 rows
head(agriculture)

###| Summary
summary(agriculture)
```

Now, the Variety column is treated as a categorical variable, rather than as characters/strings. If we look at the new summary of the data, we observe that there are 6 categories (A-F) and each category has 8 observations.

# Fitting a model

Now, let's fit a model that uses Variety and Block to predict SIZE. That is, we fit a model of the following from: 

\begin{align}
SIZE_i = & \alpha \times \mathbb{I}{(\text{observation i has Variety A})} + \beta \times \mathbb{I}{(\text{observation i has Variety B})} \\ & + \gamma \times \mathbb{I}{(\text{observation i has Variety C})} + \delta \times \mathbb{I}{(\text{observation i has Variety D})} \\ & + \theta \times \mathbb{I}{(\text{observation i has Variety E})} + \zeta \times \mathbb{I}{(\text{observation i has Variety F})} \\ & + \eta \times Block_i + \hat{\epsilon}_i,
\end{align}

where $\hat{\epsilon}_i$ is the error of the ith observation.

::: {.callout-note title="In a nutshell: indicator functions and 0 +"}
An **indicator function** $\mathbb{I}(\text{Variety A})$ is just a switch: it equals 1 if that row really is Variety A, and 0 otherwise. So for any single row, only one of the six Variety switches is "on", and that turns its coefficient into the estimate for that variety. Writing `0 +` in the formula **removes the intercept**, so instead of comparing every variety to a baseline, R gives each variety its own outright mean — that is what makes the heavy-looking algebra below work.
:::

```{r}
###| SIZE ~ Variety + Block
m1 <- lm(SIZE ~ 0 + Variety + Block, data = agriculture) # Set intercept to 0

###| Output
m1
```

Reading off of the output, we obtain the following model:

\begin{align}
SIZE_i = &57.6 \times \mathbb{I}{(\text{observation i has Variety A})} + 62.9 \times \mathbb{I}{(\text{observation i has Variety B})} \\ & + 69.6 \times \mathbb{I}{(\text{observation i has Variety C})} + 89.5 \times \mathbb{I}{(\text{observation i has Variety D})} \\ & + 73.5 \times \mathbb{I}{(\text{observation i has Variety E})} + 79.9 \times \mathbb{I}{(\text{observation i has Variety F})} \\ & + 9.1 \times Block_i + \hat{\epsilon}_i.
\end{align}

We interpret the Block variable as before: holding every other variable constant, a 1 unit increase in the Block is expected to result in a 9.1 unit increase in SIZE.

:::{.callout-important}
However, we cannot interpret the categorical variable, Variety, in the same way. Instead, we will compare what happens as we change category. 
:::

For example,

<div>

-   For a Variety A observation, and holding Block constant, we expect the SIZE to be 5.3 units smaller than if it was Variety B.
-   For a Variety F observation, and holding Block constant, we expect the SIZE to be 22.3 units larger than if it was Variety A.

</div>

# Reference categories

Using no intercept (by adding the 0 term in the `lm()` command), we can compare across categories easily. However, suppose we want to **compare everything to category A**. Instead of finding the estimates and then subtracting them, we can ask R to do this directly by setting up reference categories using `relevel()`. 

Now, our model looks like:

```{=tex}
\begin{align}
SIZE_i = & \tilde{\alpha} + \tilde{\beta} \times \mathbb{I}{(\text{observation i has Variety B})} \\ & + \tilde{\gamma} \times \mathbb{I}{(\text{observation i has Variety C})} + \tilde{\delta} \times \mathbb{I}{(\text{observation i has Variety D})} \\ & + \tilde{\epsilon} \times \mathbb{I}{(\text{observation i has Variety E})} + \tilde{\zeta} \times \mathbb{I}{(\text{observation i has Variety F})} \\ & + \tilde{\eta} \times Block_i + \hat{\epsilon}_i,
\end{align}
```

We code this up as follows: 

Note that Variety A is already the first (default) reference level, so the `relevel()` call below does not actually change anything here; it simply demonstrates the syntax. To make a different category the reference, you would change `ref` to that category (for example, `ref = "C"`).

```{r}
###| Setting Category A as the reference category
agriculture$Variety <- relevel(agriculture$Variety, ref = "A")

###| New model
m2 <- lm(SIZE ~ Variety + Block, data = agriculture) # includes an intercept (Variety A is the reference category)

###| Output
m2
```

Now, our model is: \begin{align}
SIZE_i = &57.6 + 5.3 \times \mathbb{I}{(\text{observation i has Variety B})} \\ & + 12 \times \mathbb{I}{(\text{observation i has Variety C})} + 31.9 \times \mathbb{I}{(\text{observation i has Variety D})} \\ & + 15.7 \times \mathbb{I}{(\text{observation i has Variety E})} + 22.3 \times \mathbb{I}{(\text{observation i has Variety F})} \\ & + 9.1 \times Block_i + \hat{\epsilon}_i.
\end{align}

Both of these models are equivalent! They are reparametrisations of each other, with $\alpha = \tilde{\alpha}$, $\beta = \tilde{\alpha} + \tilde{\beta}$, $\gamma = \tilde{\alpha} + \tilde{\gamma}$, ..., $\zeta = \tilde{\alpha} + \tilde{\zeta}$ and $\eta = \tilde{\eta}$. Based on the context, you should choose to include or not to include the intercept term. 

# Interaction terms

What effect does the Variety category have on the Block estimator? We explore this using interaction terms with a model of the form:

\begin{align}
SIZE_i = & \alpha + \beta \times \mathbb{I}{(\text{observation i has Variety B})} \\ & + ... + \zeta \times \mathbb{I}{(\text{observation i has Variety F})} \\ & + \eta \times Block_i + \\ & \tilde{\beta} \times \mathbb{I}{(\text{observation i has Variety B})} \times Block_i \\ & + ... + \tilde{\zeta} \times \mathbb{I}{(\text{observation i has Variety F})} \times Block_i \\ & \hat{\epsilon}_i,
\end{align} where $\hat{\epsilon}_i$ is the error of the ith observation.

```{r}
###| Model with interaction terms
m3 <- lm(SIZE ~ Variety + Block + Variety:Block, data = agriculture) 

###| Output
m3
```

We interpret the model as follows:

<div>

-   When Block = 0, the observation being Variety B rather than Variety A is associated with a SIZE that is 29.9 units smaller. Because of the interaction term, this gap changes with Block.\
-   When Block = 0, the observation being Variety C rather than Variety A is associated with a SIZE that is 24.4 units smaller. Because of the interaction term, this gap changes with Block.\
-   When Block = 0, the observation being Variety D rather than Variety A is associated with a SIZE that is 23.7 units smaller. Because of the interaction term, this gap changes with Block.\
-   When Block = 0, the observation being Variety E rather than Variety A is associated with a SIZE that is 20.3 units smaller. Because of the interaction term, this gap changes with Block.\
-   When Block = 0, the observation being Variety F rather than Variety A is associated with a SIZE that is 3.3 units larger. Because of the interaction term, this gap changes with Block.
-   If the observation is Variety A, increasing block size by 1 will increase the SIZE by 2.3.
-   If the observation is Variety B, increasing block size by 1 will increase the SIZE by 7.8.
-   If the observation is Variety C, increasing block size by 1 will increase the SIZE by 8.1.
-   If the observation is Variety D, increasing block size by 1 will increase the SIZE by 12.3.
-   If the observation is Variety E, increasing block size by 1 will increase the SIZE by 8.0.
-   If the observation is Variety F, increasing block size by 1 will increase the SIZE by 2.3.

</div>


:::{.callout-important}
When using interaction terms, it is important to remember the principle of parsimony. If we decide to model the interaction between variable A and variable B, we must include both variables in the model.
:::

::: {.callout-tip title="Going further 🚀"}
Reading effects straight off a table of coefficients is fiddly, especially with interactions. The `emmeans` package gives the estimated mean SIZE for each Variety directly, and `interactions::interact_plot()` draws how the Block slope changes by Variety. You can also just feed new rows to `predict()`.

```{r}
#| eval: false
library(emmeans)
emmeans(m3, ~ Variety)             # estimated mean SIZE per Variety

library(interactions)
interact_plot(m3, pred = Block, modx = Variety)  # visualise the interaction
```
:::