---
title: "Loading in data"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

## Why loading data matters for modelling

Before you fit a model, you must make sure the data is loaded correctly and is in the right format. If the input data is wrong, any model you fit may give misleading results.

For modelling, the key checks are:

- the dataset has the expected rows and columns,
- each column is the right type (numeric, factor, character),
- missing values are handled appropriately,
- categorical variables are encoded correctly,
- any transformations needed for the model are prepared beforehand.

---

## Loading data in R

### Base R: `read.csv()`

You can load a CSV file using the base R function `read.csv()`.

```{r}
# Load data from a local file in the Data folder
df <- read.csv("../Data/agriculture.csv", stringsAsFactors = FALSE)

# Look at the first few rows
head(df)

# Check the structure
str(df)
```

### Tidyverse: `readr::read_csv()`

If you prefer the tidyverse style, use `read_csv()` from the `readr` package.

```{r}
library(readr)
df <- read_csv("../Data/agriculture.csv")

# Inspect the dataset
str(df)
```

:::{.callout-note}
Using `read_csv()` often gives cleaner default column types and better handling of missing values than `read.csv()`.
:::

---

## Inspecting the dataset before modelling

Once the file is loaded, perform a few simple checks.

```{r}
# Number of observations and variables
dim(df)

# Column names and sample values
names(df)
summary(df)

# Check for missing values
colSums(is.na(df))
```

### Identify variable types

For modelling, it is important to distinguish between numeric and categorical variables.

```{r}
# Example of converting a column to a factor
if ("Variety" %in% names(df)) {
  df$Variety <- as.factor(df$Variety)
}

# Confirm column types again
sapply(df, class)
```

If a variable contains categories such as group labels, season names, or treatment types, convert it to a factor before fitting a model.

---

## Preparing data for a linear model

### Remove or flag missing values

Different modelling functions handle missing values differently. For a simple model, remove rows with missing values or use explicit imputation.

```{r}
# Remove rows that contain any missing values
dfClean <- na.omit(df)

# Confirm no missing values remain
colSums(is.na(dfClean))
```

### Create model-ready variables

Some variables need transformation before modelling.

```{r}
# Example: if a numeric variable is stored as text, convert it explicitly
if ("Yield" %in% names(dfClean)) {
  dfClean$Yield <- as.numeric(dfClean$Yield)
}

# Example: create a new variable for a model
if ("SoilQuality" %in% names(dfClean)) {
  dfClean$highQuality <- dfClean$SoilQuality > 7
}
```

:::{.callout-warning}
Always check the output of conversions such as `as.numeric()` because invalid text values can become `NA` silently.
:::

---

## Quick modelling check

A final modelling check should confirm the data is ready for `lm()` or similar functions.

```{r}
# Example using a simple linear model if the variables exist
if (all(c("Yield", "Nitrogen") %in% names(dfClean))) {
  mod <- lm(Yield ~ Nitrogen, data = dfClean)
  summary(mod)
}
```

This is a useful way to verify that the data types and values are appropriate for the model.

---

## Check your understanding

1. Load a dataset from the `Data` folder and inspect the first 10 rows.
2. Convert one categorical variable to a factor and one numeric variable to numeric type.
3. Remove rows with missing values and report how many observations remain.
4. Fit a simple linear model using one numeric predictor and one numeric response.

```{r}
#| code-fold: true
###| Answer example
library(readr)
dfEx <- read_csv("../Data/agriculture.csv")

# Inspect the data
head(dfEx, 10)

# Convert variables if needed
if ("Variety" %in% names(dfEx)) {
  dfEx$Variety <- as.factor(dfEx$Variety)
}

if ("Yield" %in% names(dfEx)) {
  dfEx$Yield <- as.numeric(dfEx$Yield)
}

# Clean missing values
dfExClean <- na.omit(dfEx)
print(dim(dfExClean))

# Fit a model if the columns exist
if (all(c("Yield", "Nitrogen") %in% names(dfExClean))) {
  modEx <- lm(Yield ~ Nitrogen, data = dfExClean)
  summary(modEx)
}
```
