---
title: "Diagnostics and Inference for GLMs"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

Fitting a GLM is one line; **trusting** it is the real work — exactly the lesson from the linear-modelling and frequentist sections, now in the GLM world. A linear model is judged by residual sums of squares and the four assumption plots; a GLM is judged by its close cousins, built from the **deviance** and the likelihood. This page gathers the tools that apply to *every* GLM: measuring fit, comparing models, and quantifying uncertainty.

We use the Poisson model of `warpbreaks` from the previous page as the running example.

```{r}
#| warning: false
#| message: false

model <- glm(breaks ~ wool + tension, data = warpbreaks, family = poisson)
```

# Deviance: the GLM's residual sum of squares

A linear model measures lack of fit with the residual sum of squares. A GLM uses the **deviance**, defined from the log-likelihood as
$$
D = 2\big[\,\ell_{\text{saturated}} - \ell_{\text{model}}\,\big],
$$
the gap between your fitted model and a "perfect" *saturated* model that reproduces the data exactly. Smaller deviance means a better fit. Every `glm` reports two of them:

- the **null deviance** — the fit of the intercept-only model (no predictors);
- the **residual deviance** — the fit of *your* model.

The drop from null to residual is how much your predictors explain — the GLM version of "explained variation."

```{r}
c(nullDeviance     = model$null.deviance, nullDf = model$df.null,
  residualDeviance = model$deviance,      residualDf = model$df.residual)
```

::: {.callout-note title="In a nutshell: deviance is just badness-of-fit"}
Think of deviance as the GLM's "leftover error", playing the role the residual sum of squares plays for a linear model: **smaller is better**, and adding a useful predictor should make it drop. The *null deviance* is where you start (nothing but an average); the *residual deviance* is where your model gets to. A big drop means your predictors earned their place.
:::

# A rough goodness-of-fit check

For a Poisson or binomial model (with reasonably large counts per row), a well-fitting model has a **residual deviance roughly equal to its degrees of freedom**. A residual deviance much *larger* than its df is the classic fingerprint of **overdispersion** or a missing predictor.

```{r}
c(residualDeviance = model$deviance,
  df               = model$df.residual,
  ratio            = model$deviance / model$df.residual)   # want ≈ 1
```

A ratio comfortably above 1 here echoes the overdispersion warning from the Poisson page — a reminder to check the variance assumption before believing the standard errors. (This deviance-based check and the Pearson dispersion estimate from the previous page tell the same story by slightly different routes.)

# Residuals for GLMs

Raw residuals $y_i - \hat\mu_i$ are not very useful for a GLM because the variance changes with the mean. Instead R gives two standardised kinds:

- **Deviance residuals** (the default) — each observation's signed contribution to the total deviance; they sum in squares to $D$.
- **Pearson residuals** — $(y_i - \hat\mu_i)/\sqrt{\widehat{\text{Var}}(\hat\mu_i)}$, the residual scaled by its expected standard deviation.

```{r}
#| fig-align: center

plot(fitted(model), residuals(model, type = "deviance"),
     pch = 19, col = "steelblue",
     xlab = "Fitted mean (count)", ylab = "Deviance residual",
     main = "Deviance residuals vs fitted values")
abline(h = 0, col = "firebrick", lwd = 2, lty = 2)
```

Read this like the linear-model residual plot: you want a structureless band around zero. **Curved** trends suggest the wrong link or a missing term; a **funnel** suggests the variance assumption is off (overdispersion again). Do not expect the tidy cloud of a linear model — for low counts the discreteness shows as stripes, which is normal.

::: {.callout-note title="In a nutshell: GLM residual plots look different"}
GLM residual plots are *lumpier* than linear-model ones — counts and 0/1 outcomes are discrete, so residuals fall into bands and stripes. Don't panic at the pattern itself; look for the *trends* (a slope or a curve) and *fans* (spreading variance) that signal a real problem.
:::

# Comparing nested models: the likelihood-ratio test

To ask "do these extra predictors earn their place?", compare two **nested** GLMs by how much the deviance drops. Because deviance is built from the log-likelihood, the drop is a **likelihood-ratio test statistic** — the very same idea (and the same $\chi^2$ reference distribution, via Wilks' theorem) as the likelihood-ratio test in the Frequentist section. `anova()` with `test = "Chisq"` does it.

```{r}
modelSmall <- glm(breaks ~ wool,            data = warpbreaks, family = poisson)
modelFull  <- glm(breaks ~ wool + tension,  data = warpbreaks, family = poisson)

anova(modelSmall, modelFull, test = "Chisq")
```

The deviance drop from adding `tension` is tested against a $\chi^2$ with degrees of freedom equal to the number of extra parameters. A small p-value says the larger model fits significantly better — `tension` belongs in the model. This is the GLM workhorse for testing whole groups of terms (e.g. all levels of a factor at once).

# Model selection with AIC

For comparing models that are **not** nested — different predictors, or a Poisson versus a negative binomial — the likelihood-ratio test does not apply, and we use the **Akaike Information Criterion**:
$$
\text{AIC} = -2\,\ell(\hat\theta) + 2p,
$$
log-likelihood penalised by the number of parameters $p$. It rewards fit and punishes complexity; **lower is better**.

```{r}
AIC(modelSmall, modelFull)
```

The model with the lower AIC is preferred — here, confirming that the richer model is worth its extra parameters. AIC is the principled, automatic version of the **principle of parsimony** from the linear-modelling section.

# Confidence intervals for the coefficients

Finally, uncertainty. R offers two flavours, the same pairing as the frequentist asymptotics page:

- `confint()` — **profile-likelihood** intervals, which respect the true (possibly skewed) shape of the likelihood. Preferred for GLMs.
- `confint.default()` — fast **Wald** intervals ($\hat\beta \pm 1.96\,\text{se}$), which assume the likelihood is symmetric.

```{r}
#| warning: false
#| message: false

# On the link (log) scale; exponentiate to get rate ratios with their CIs
exp(confint(model))
```

These are the interval companions to the rate ratios from the Poisson page — and being profile-based, they stay faithful to the likelihood even when it is asymmetric.

::: {.callout-tip title="Going further 🚀"}
The one-call diagnostic suites that exist for linear models work for GLMs too: **`performance::check_model()`** produces a full panel of GLM-appropriate diagnostic plots, **`performance::check_overdispersion()`** runs a formal dispersion test, and **`DHARMa`** simulates *scaled residuals* that behave like uniform residuals for any GLM — the most reliable modern way to spot misfit, overdispersion and zero-inflation. `broom::tidy(model, exponentiate = TRUE, conf.int = TRUE)` returns all the rate/odds ratios and intervals as a tidy data frame ready for a report.
:::

# Summary

| Question | Tool |
|---|---|
| How badly does it fit? | **Deviance** (null vs residual); ratio to df $\approx 1$ |
| Are the residuals healthy? | **Deviance / Pearson residuals** vs fitted |
| Do these extra terms help? | **Likelihood-ratio test**, `anova(..., test = "Chisq")` |
| Which non-nested model? | **AIC** (lower is better) |
| How uncertain are the effects? | **Profile** CIs, `confint()` |
| Is it overdispersed? | Residual deviance $\gg$ df; `check_overdispersion()` |

Every one of these is a likelihood quantity in disguise — which is why the Frequentist Statistics section underpins this whole topic. With fitting, interpretation and diagnostics in hand, the final page lets you put them together yourself.