---
title: "Logistic Regression (Binary Data)"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

Our first GLM tackles the most common non-normal response of all: a **binary** outcome — success/failure, yes/no, $1/0$. The model is **logistic regression**, the Binomial-distribution, logit-link member of the GLM family from the previous page. We will fit one, read its coefficients as **odds**, turn it into predicted probabilities, and use it to **classify**.

# The model

For a binary response $Y_i \in \{0,1\}$, the only thing to model is the probability of a success, $p_i = P(Y_i = 1)$. We attach it to a linear predictor through the **logit** link:
$$
\underbrace{\log\!\frac{p_i}{1-p_i}}_{\text{log-odds of success}} = \beta_0 + \beta_1 x_{i1} + \dots + \beta_p x_{ip}.
$$
Inverting the link gives the probability itself, guaranteed to land in $[0,1]$:
$$
p_i = \frac{1}{1 + e^{-(\beta_0 + \beta_1 x_{i1} + \dots)}}.
$$
That second formula is the famous **S-shaped (sigmoid) curve**: as the linear predictor sweeps from $-\infty$ to $+\infty$, the probability rises smoothly from 0 to 1, never escaping the interval.

::: {.callout-note title="In a nutshell: odds and log-odds"}
The **odds** of an event are $\dfrac{p}{1-p}$ — the chance it happens divided by the chance it doesn't. A probability of $0.8$ is odds of $0.8/0.2 = 4$ ("4 to 1 on"). The **log-odds** (or *logit*) is just $\log$ of that. We model the log-odds because, unlike a probability, it can be any real number — exactly what a linear predictor needs.
:::

# Fitting a logistic regression

We use R's built-in `mtcars` and ask: from a car's **weight**, can we predict whether it has a **manual** gearbox (`am = 1`) rather than an automatic (`am = 0`)? Fitting is one call to `glm()` with `family = binomial` — note how closely it mirrors `lm()`.

```{r}
#| warning: false
#| message: false

model <- glm(am ~ wt, data = mtcars, family = binomial)
summary(model)
```

Read the `summary` exactly as you would a linear model's, with **one crucial difference: the coefficients are on the log-odds scale.** The `wt` coefficient is negative and significant — heavier cars have *lower* log-odds of being manual, which matches intuition (manual cars in this data, like the Lotus and Honda, tend to be light).

# Interpreting the coefficients: odds ratios

A coefficient of $\beta_1$ says "a one-unit increase in $x_1$ adds $\beta_1$ to the **log-odds**" — true, but hard to feel. Exponentiating turns it into something concrete: $e^{\beta_1}$ is the **odds ratio**, the factor by which the odds of success are multiplied for each one-unit increase in the predictor.

```{r}
# Coefficients on the odds scale, with 95% confidence intervals
exp(cbind(OddsRatio = coef(model), confint(model)))
```

The odds ratio for `wt` is well below 1: each extra 1000 lb of weight **multiplies** the odds of a manual gearbox by that factor — i.e. sharply reduces them. An odds ratio of 1 would mean "no effect"; above 1 means the predictor raises the odds, below 1 means it lowers them.

::: {.callout-note title="In a nutshell: reading an odds ratio"}
- Odds ratio $= 1$ → the predictor makes no difference.
- Odds ratio $= 2$ → each unit *doubles* the odds of success.
- Odds ratio $= 0.5$ → each unit *halves* the odds.

It is a **multiplicative** effect on the odds, not an additive effect on the probability — which is why the same coefficient changes the probability a lot in the middle of the S-curve and barely at all near 0 or 1.
:::

# Predicted probabilities and the S-curve

To get back to plain probabilities, predict with `type = "response"` (which applies the inverse link for you). Plotting these against weight reveals the characteristic sigmoid.

```{r}
#| fig-align: center

# A smooth grid of weights, and the model's predicted P(manual)
wtGrid <- data.frame(wt = seq(min(mtcars$wt), max(mtcars$wt), length.out = 200))
wtGrid$pHat <- predict(model, newdata = wtGrid, type = "response")

plot(mtcars$wt, mtcars$am, pch = 19, col = "steelblue",
     xlab = "Weight (1000 lbs)", ylab = "P(manual gearbox)",
     main = "Logistic fit: probability of a manual gearbox vs weight")
lines(wtGrid$wt, wtGrid$pHat, lwd = 3, col = "firebrick")
abline(h = 0.5, col = "grey50", lty = 3)
legend("topright", bty = "n", lwd = c(NA, 3), pch = c(19, NA),
       col = c("steelblue", "firebrick"),
       legend = c("observed (0/1)", "fitted probability"))
```

The fitted curve glides from near 1 for light cars down to near 0 for heavy ones, passing through $0.5$ around the weight where manual and automatic become equally likely — and, unlike a straight line, it never strays outside $[0,1]$.

# More than one predictor

Nothing changes with extra predictors — we just add them to the linear predictor, exactly as in multiple linear regression. Let us add horsepower.

```{r}
model2 <- glm(am ~ wt + hp, data = mtcars, family = binomial)
round(exp(coef(model2)), 4)      # odds ratios for each predictor
```

Each odds ratio is now interpreted **holding the other predictor fixed** — the same "all else equal" reading as in linear regression, but multiplicative on the odds.

# From probabilities to a classification

A logistic model outputs probabilities; to make a **decision** we pick a threshold (commonly $0.5$) and predict "success" above it. Comparing predictions to the truth gives a **confusion matrix**.

```{r}
pHat      <- predict(model2, type = "response")
predicted <- ifelse(pHat > 0.5, 1, 0)

table(predicted = predicted, actual = mtcars$am)
```

The diagonal entries are correct classifications, the off-diagonal ones are errors. The $0.5$ cut-off is a *choice*, not a law: raising it makes the model more cautious about declaring a success (fewer false positives, more false negatives), and the right trade-off depends on the costs of each kind of mistake.

::: {.callout-tip title="Going further 🚀"}
A single threshold hides how the model performs across *all* cut-offs. The **ROC curve** plots the true-positive against the false-positive rate as the threshold varies, and the **area under it (AUC)** summarises overall discrimination in one number:

```{r}
#| eval: false
library(pROC)
roc(mtcars$am, pHat) |> auc()      # 0.5 = random, 1 = perfect
```

For binary responses recorded as *proportions* (e.g. 8 successes out of 10 trials per row), fit with `glm(cbind(successes, failures) ~ ..., family = binomial)`. And when the outcome has **more than two categories**, the natural extensions are `nnet::multinom()` (unordered) and `MASS::polr()` (ordered) — same linear-predictor idea, more than one logit.
:::

# Summary

| Idea | Logistic regression |
|---|---|
| Response | Binary $0/1$ (or a proportion) |
| Distribution / link | Binomial / logit |
| Model | $\log\frac{p}{1-p} = X\beta$, so $p = 1/(1+e^{-X\beta})$ |
| Coefficient $\beta_j$ | Additive effect on the **log-odds** |
| $e^{\beta_j}$ | **Odds ratio** — multiplicative effect on the odds |
| `glm(y ~ x, family = binomial)` | Fits it by maximum likelihood |
| `predict(..., type = "response")` | Returns probabilities; threshold to classify |

Logistic regression is the GLM template with a Binomial response and a logit link — the linear predictor unchanged, the link doing the work of keeping every prediction a genuine probability. Next we swap the distribution again, for **counts**.