---
title: "Hypothesis Testing"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

So far this section has been about **estimation**: the likelihood gave us a point estimate $\hat\theta$, and asymptotics and simulation gave us its uncertainty as a confidence interval. **Hypothesis testing** is the other side of the same coin. Instead of asking *"what is $\theta$?"* it asks a sharper question: *"is the data compatible with one specific claim about $\theta$?"*

Remarkably, this needs no new machinery. The score $U(\theta)$, the Fisher information $I(\theta)$, asymptotic normality and Wilks' theorem — every tool is already on the table from the likelihood and asymptotics pages. A test is just those tools pointed in a new direction.

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart LR
  D["Data"] --> E["Estimation<br/>'What is θ?'<br/>→ θ̂ and a CI"]
  D --> H["Testing<br/>'Is θ = θ₀ plausible?'<br/>→ a p-value and a decision"]
  E -. "two sides of one coin<br/>(duality)" .-> H
```

# The logic of a hypothesis test

A hypothesis test pits two rival claims against each other.

- The **null hypothesis** $H_0$ is the specific, "boring" claim we put on trial — usually that a parameter equals some fixed value, $\theta = \theta_0$ (e.g. "the coin is fair, $\theta = 0.5$"). It is the *status quo* we will only abandon with evidence.
- The **alternative hypothesis** $H_1$ is everything else we would accept instead (e.g. $\theta \neq 0.5$).

The test works like a courtroom that **presumes the null is true** and asks whether the data is too surprising for that presumption to survive.

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart TD
  A["Assume H₀ is true"] --> B["Compute a test statistic<br/>measuring how far the data<br/>strays from H₀"]
  B --> C["How likely is a value this extreme<br/>IF H₀ were true? → the p-value"]
  C --> D{"p-value < α?"}
  D -->|"Yes"| E["Reject H₀<br/>(the data is too surprising)"]
  D -->|"No"| F["Do not reject H₀<br/>(data is compatible with it)"]
```

The four moving parts:

1. **Test statistic** — a single number summarising how far the observed data departs from what $H_0$ predicts. Large values count against $H_0$.
2. **Null distribution** — the distribution of that statistic *if $H_0$ were true*. For the tests below it is (asymptotically) a chi-squared or a standard normal.
3. **p-value** — the probability, computed under $H_0$, of seeing a statistic at least as extreme as the one we observed. A *small* p-value means "this data would be very unusual if $H_0$ were true," which is evidence against $H_0$.
4. **Significance level $\alpha$** — the threshold (commonly $0.05$) we fix *in advance*. Reject $H_0$ when $p < \alpha$.

::: {.callout-important title="What a p-value is — and is not"}
The p-value is $P(\text{data this extreme} \mid H_0)$. It is **not** the probability that $H_0$ is true — that would be a statement about $\theta$, which a frequentist refuses to make, since $\theta$ is fixed, not random. "$p = 0.03$" means *"if the null held, data like this would arise only 3% of the time,"* not *"there is a 3% chance the null is true."*
:::

::: {.callout-note title="One-sided vs two-sided alternatives"}
The alternative above, $H_1: \theta \neq \theta_0$, is **two-sided** — it counts a departure in *either* direction as evidence against the null. Sometimes only one direction is of interest (e.g. "is the new treatment *better*?", $H_1: \theta > \theta_0$), giving a **one-sided** test that puts the whole rejection region in one tail and is more powerful *when the truth really does lie that way*. The signed Wald statistic $Z_{\text{W}}$ below carries this direction, so one-sided tests come straight from it. The squared trinity statistics we focus on, however, are inherently two-sided — squaring throws the sign away. Choose the direction *before* seeing the data: picking a one-sided test after peeking is a classic way to manufacture significance.
:::

# Two ways to be wrong: errors and power

Because the data is random, any decision rule sometimes errs. There are exactly two ways:

| | **$H_0$ is really true** | **$H_0$ is really false** |
|---|---|---|
| **We reject $H_0$** | Type I error (false positive), prob. $\alpha$ | Correct decision |
| **We do not reject $H_0$** | Correct decision | Type II error (false negative), prob. $\beta$ |

- The **significance level $\alpha$** *is* the Type I error rate — the chance of crying wolf when the null is true. We control it directly by choosing the cut-off.
- The **power** of a test is $1 - \beta$: the probability of correctly rejecting a *false* null. Power rises with the sample size, the effect size (how far the truth is from $\theta_0$), and the significance level.

::: {.callout-note title="The asymmetry of testing"}
A test never "accepts" $H_0$ — it only *fails to reject* it. Absence of evidence is not evidence of absence: a non-significant result may mean the null is true, or merely that we had too little data (low power) to detect a real effect. This is why we say "do not reject," not "accept."
:::

::: {.callout-warning title="α is per-test: beware multiple comparisons"}
The 5% Type I error rate applies to *one* test. Run many tests and the false positives accumulate: testing 20 true nulls at $\alpha = 0.05$ gives roughly a $1 - 0.95^{20} \approx 64\%$ chance of at least one spurious "discovery." Hunting through a dataset for *any* significant result — *p-hacking* — exploits exactly this. When you perform many tests, adjust for it (e.g. the **Bonferroni** correction uses $\alpha/m$ for $m$ tests), and treat a lone significant result among many with suspicion.
:::

# A health warning: p-hacking and bad practice

The machinery on this page is only trustworthy if the test is decided *before* the data is examined. The Type I error guarantee — "at most a 5% chance of a false positive" — silently assumes you run **one pre-specified test** on the data. Almost every way of bending that assumption inflates the real false-positive rate far above the advertised $\alpha$, usually without the analyst noticing. Collectively these are **p-hacking** (or "questionable research practices"), and they are the single biggest reason published findings fail to replicate.

The usual culprits:

- **Fishing for significance.** Trying many variables, subgroups, or transformations and reporting only the combination that crossed $p < 0.05$ — the multiple-comparisons trap above, dressed up as a single result.
- **Optional stopping.** Peeking at the data and stopping collection the moment $p < 0.05$. If you keep testing as data trickles in, you are *guaranteed* to eventually cross the threshold by chance, even under a true null.
- **HARKing** (Hypothesising After the Results are Known). Running the test, seeing what came out, and then presenting that as the hypothesis you meant to test all along — turning an exploratory accident into a fake confirmation.
- **Selective reporting / dropping data.** Quietly excluding "inconvenient" observations, or burying the analyses that didn't work, until the survivors look significant.

::: {.callout-tip title="Going further 🚀"}
The antidotes are mostly about *discipline*, not statistics: **pre-register** your hypothesis and analysis plan before collecting data; report *every* test you ran, not just the significant ones; treat anything found by exploration as a hypothesis to confirm on *fresh* data, never as a result; and remember that **a p-value is only interpretable for a test you committed to in advance**. A good rule of thumb: if you would not have run exactly this test had the data looked different, the p-value no longer means what it claims.
:::

# The trinity: Wald, score and likelihood-ratio tests

For parametric models there are three classical large-sample tests, sometimes called the **holy trinity**. All three test the same $H_0: \theta = \theta_0$, and all three are asymptotically $\chi^2_1$ under the null — but they measure the discrepancy between $\hat\theta$ and $\theta_0$ in three geometrically different ways on the log-likelihood curve.

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart TD
  H0["H₀: θ = θ₀"]
  H0 --> W["<b>Wald</b><br/>How far is θ̂ from θ₀?<br/>(horizontal distance,<br/>scaled by curvature at θ̂)"]
  H0 --> S["<b>Score</b><br/>How steep is ℓ at θ₀?<br/>(slope at the null,<br/>scaled by info at θ₀)"]
  H0 --> L["<b>Likelihood-ratio</b><br/>How much higher is ℓ(θ̂)<br/>than ℓ(θ₀)?<br/>(vertical drop)"]
  W --> C["All three →  χ²₁  under H₀"]
  S --> C
  L --> C
```

## The Wald test

Start at the peak and measure how many standard errors the null value sits away from the MLE. Using $\operatorname{se}(\hat\theta) = 1/\sqrt{I(\hat\theta)}$,
$$
\boxed{\,Z_{\text{W}} = \frac{\hat\theta - \theta_0}{\operatorname{se}(\hat\theta)}\,}\qquad\text{or, squared,}\qquad W_{\text{W}} = I(\hat\theta)\,(\hat\theta - \theta_0)^2 \;\overset{\text{approx.}}{\sim}\; \chi^2_1 .
$$
It only needs the fit at $\hat\theta$ — the same ingredients as the Wald *confidence interval*. **Cheap, but it assumes the likelihood is symmetric (parabolic) around the peak.**

## The score test

This one never fits the alternative at all — it stays at $\theta_0$ and asks how steep the log-likelihood still is there. If $\theta_0$ were the true peak, the slope (score) would be near zero; a steep slope says we are far from the maximum. Standardising the score by its standard deviation $\sqrt{I(\theta_0)}$,
$$
\boxed{\,W_{\text{S}} = \frac{U(\theta_0)^2}{I(\theta_0)} \;\overset{\text{approx.}}{\sim}\; \chi^2_1\,}
$$
Its great practical advantage: **it only requires the model fitted under $H_0$**, never the unrestricted MLE. When the null model is far easier to fit than the full one, the score test is the natural choice.

## The likelihood-ratio test

This is the test we already half-built on the asymptotics page. Compare the height of the log-likelihood at its peak with its height at the null:
$$
\boxed{\,W_{\text{LR}} = 2\big[\,\ell(\hat\theta) - \ell(\theta_0)\,\big] \;\overset{\text{approx.}}{\sim}\; \chi^2_1\,}
$$
By **Wilks' theorem** (derived on the asymptotics page) this is $\chi^2_1$ under $H_0$. It uses the genuine shape of the likelihood rather than a parabolic approximation, so it is generally the most accurate of the three — at the cost of fitting *both* the null and the full model.

::: {.callout-note title="In a nutshell: one question, three rulers"}
Do not let three formulas fool you into thinking there are three different ideas. All three ask the very same thing — *"is $\theta_0$ far from the peak of the log-likelihood?"* — they just **measure the distance differently**: Wald reads it sideways (how far $\hat\theta$ sits from $\theta_0$), the score reads the slope at $\theta_0$, and the likelihood-ratio reads the drop in height. When the peak is a tidy parabola, all three rulers give the same answer.
:::

::: {.callout-note title="A note on notation: I(θ̂)"}
On the likelihood page we distinguished the *observed* information $J(\theta) = -\ell''(\theta)$ from the *expected* (Fisher) information $I(\theta) = \mathbb{E}[-\ell''(\theta)]$. In the formulas above $I(\hat\theta)$ and $I(\theta_0)$ are the information evaluated at a single point — for the coin model below they coincide because the expectation is available in closed form, but in general practitioners plug in whichever is to hand (often the observed information straight from the fit, e.g. `optim`'s Hessian). The asymptotics are the same either way.
:::

::: {.callout-note title="Three views of the same hill"}
Picture the log-likelihood as a hill with its summit at $\hat\theta$. The **Wald** test measures the *horizontal* distance from the summit to $\theta_0$; the **likelihood-ratio** test measures the *vertical* drop from the summit down to the height above $\theta_0$; the **score** test measures the *slope* of the hillside at $\theta_0$. When the hill is a perfect parabola all three agree exactly — they differ only by how skewed the real likelihood is, which is also why they can disagree in small samples.
:::

# Confidence intervals and tests are dual

There is a deep equivalence, already hinted at on the asymptotics page:

$$
\theta_0 \text{ lies in the 95\% CI} \iff \text{the test of } H_0:\theta=\theta_0 \text{ is } \textbf{not} \text{ rejected at } \alpha = 0.05 .
$$

A confidence interval is exactly *the set of null values a test would not reject*, and a test is exactly *the question of whether a value falls in the interval*. This is why the Wald CI ($\hat\theta \pm 1.96\,\operatorname{se}$) and the Wald test ($|Z_{\text{W}}| > 1.96$) are the same statement, and why inverting the likelihood-ratio statistic on the asymptotics page gave a confidence interval. Estimation and testing are two readings of one picture.

# Worked example in R

We reuse the **coin-flip** model from the likelihood and asymptotics pages, and test whether the coin is fair:
$$
H_0: \theta = 0.5 \qquad\text{vs}\qquad H_1: \theta \neq 0.5 .
$$
Everything is in closed form: $\hat\theta = k/n$, score $U(\theta) = \tfrac{k}{\theta} - \tfrac{n-k}{1-\theta}$, information $I(\theta) = \tfrac{n}{\theta(1-\theta)}$.

```{r}
#| warning: false
#| message: false

set.seed(1)

# Simulate coin flips with a true (but "unknown") bias
trueTheta <- 0.7
n     <- 200
flips <- rbinom(n, size = 1, prob = trueTheta)
k     <- sum(flips)

theta0 <- 0.5            # the null value: "the coin is fair"

# Building blocks
ll    <- function(theta) k * log(theta) + (n - k) * log(1 - theta)
score <- function(theta) k / theta - (n - k) / (1 - theta)
info  <- function(theta) n / (theta * (1 - theta))

mle <- k / n
c(k = k, mle = mle)
```

Now compute all three test statistics. Each is compared against a $\chi^2_1$ distribution, whose upper 5% cut-off is $\approx 3.84$.

```{r}
# 1. Wald: distance of the MLE from theta0, scaled by info AT THE MLE
wWald  <- info(mle) * (mle - theta0)^2

# 2. Score: steepness of the log-likelihood AT theta0, scaled by info AT theta0
wScore <- score(theta0)^2 / info(theta0)

# 3. Likelihood-ratio: vertical drop from the peak to the null
wLR    <- 2 * (ll(mle) - ll(theta0))

# Each statistic, with its p-value from the chi-squared(1) null distribution
results <- rbind(
  Wald            = c(statistic = wWald,  pValue = 1 - pchisq(wWald,  df = 1)),
  Score           = c(statistic = wScore, pValue = 1 - pchisq(wScore, df = 1)),
  LikelihoodRatio = c(statistic = wLR,    pValue = 1 - pchisq(wLR,    df = 1))
)
round(results, 4)
```

All three statistics sit far above the $3.84$ cut-off and the p-values are essentially zero, so we **decisively reject** the null that the coin is fair — unsurprising, since the data was generated with $\theta = 0.7$. Note how closely the three statistics agree: with $n = 200$ the likelihood is nearly parabolic, exactly the regime where the trinity coincides.

## Seeing the three tests on the log-likelihood

The picture below makes the geometry of the *Note* above concrete. The **likelihood-ratio** statistic is the vertical gap between the two dashed horizontal lines; the **Wald** statistic comes from the horizontal gap $\hat\theta - \theta_0$; the **score** statistic comes from the slope of the tangent line drawn at $\theta_0$.

```{r}
#| fig-align: center

grid <- seq(0.35, 0.85, length.out = 400)

plot(grid, ll(grid), type = "l", lwd = 2, col = "steelblue",
     xlab = expression(theta), ylab = expression(ell(theta)),
     main = "The three tests as features of the log-likelihood")

# Vertical lines at the MLE and the null
abline(v = mle,    col = "firebrick", lwd = 2, lty = 2)
abline(v = theta0, col = "darkgreen", lwd = 2, lty = 3)

# Heights at the peak and at the null: their vertical gap is the LR statistic
abline(h = ll(mle),    col = "firebrick", lwd = 1, lty = 2)
abline(h = ll(theta0), col = "darkgreen", lwd = 1, lty = 3)

# Tangent (slope = score) at theta0: a near-vertical slope means strong evidence
slope <- score(theta0)
curve(ll(theta0) + slope * (x - theta0), add = TRUE,
      col = "darkorange", lwd = 2, lty = 1, from = 0.42, to = 0.58)

legend("bottom", bty = "n", lwd = 2, lty = c(1, 2, 3, 1),
       col = c("steelblue", "firebrick", "darkgreen", "darkorange"),
       legend = c("log-likelihood",
                  expression(paste("MLE  ", hat(theta))),
                  expression(paste("null  ", theta[0])),
                  "tangent at the null (score)"))
```

## Cross-check against R's built-in test

R already knows how to test a proportion. `prop.test` (a score-type test, with an optional continuity correction) and the exact `binom.test` should broadly agree with our hand-rolled statistics:

```{r}
prop.test(k, n, p = theta0, correct = FALSE)$p.value   # score-type
binom.test(k, n, p = theta0)$p.value                   # exact
```

Both return essentially zero, confirming our trinity by an independent route.

# Seeing power in action

Earlier we *defined* power as $1-\beta$ and claimed it rises with the sample size and the effect size. With the coin model in hand we can stop asserting this and simply **draw it**. Testing $H_0: \theta = 0.5$ two-sided at $\alpha = 0.05$, we reject when the estimate strays more than $1.96$ null standard errors from $0.5$. For any *true* $\theta$ the estimate $\hat\theta$ is approximately $\mathcal N\big(\theta,\, \theta(1-\theta)/n\big)$, so the probability of landing in the rejection region — the power — is a quantity we can compute directly.

```{r}
#| fig-align: center

# Power of the two-sided test of H0: theta = 0.5, at level alpha = 0.05
powerCoin <- function(theta, n, theta0 = 0.5, alpha = 0.05) {
  se0  <- sqrt(theta0 * (1 - theta0) / n)         # SE under the null
  crit <- qnorm(1 - alpha / 2)                    # 1.96
  lower <- theta0 - crit * se0                    # rejection boundaries
  upper <- theta0 + crit * se0
  se <- sqrt(theta * (1 - theta) / n)             # SE under the true theta
  # P(reject) = P(thetaHat below lower) + P(thetaHat above upper)
  pnorm(lower, theta, se) + (1 - pnorm(upper, theta, se))
}

grid <- seq(0.3, 0.7, length.out = 400)

plot(NA, xlim = range(grid), ylim = c(0, 1),
     xlab = expression(paste("true  ", theta)), ylab = "power  (1 - β)",
     main = "Power of the coin-fairness test")
for (i in seq_along(c(20, 50, 200))) {
  nn <- c(20, 50, 200)[i]
  lines(grid, powerCoin(grid, nn), lwd = 2, col = i + 1)
}
abline(h = 0.05, col = "grey50", lty = 3)   # at theta = 0.5 power = alpha
abline(v = 0.5,  col = "grey50", lty = 3)
legend("top", bty = "n", lwd = 2, col = c(2, 3, 4),
       legend = c("n = 20", "n = 50", "n = 200"))
```

Two things the picture makes obvious. **At $\theta = 0.5$ the curve dips to $\alpha = 0.05$** — when the null is *true*, "power" is just the Type I error rate, the 5% chance of rejecting anyway. And **the curves climb steeply as the truth moves away from $0.5$ and as $n$ grows**: a bigger sample, or a more biased coin, makes a real departure easier to detect. This is exactly why a non-significant result is not proof of fairness — with $n = 20$, even a genuinely biased $\theta = 0.6$ coin is caught less than half the time.

# Testing when there is no formula

Just as with confidence intervals, the trinity relies on the asymptotic $\chi^2_1$ approximation, which can be poor in small samples or for awkward models. When it cannot be trusted, the same **simulation** ideas from the previous page apply: generate many data sets *under the null*, recompute the statistic on each, and read the p-value off the simulated null distribution directly — a *Monte Carlo* or *parametric bootstrap* test. The principle is identical to the coverage study on the simulation page: when the formula is shaky, simulate the repetitions the formula was only approximating.

::: {.callout-tip title="Going further 🚀"}
**Planning the sample size.** Power is most useful *before* collecting data — to ask "how big must $n$ be to detect the effect I care about?" The base `power.prop.test` does this for two proportions, and the **`pwr`** package covers many other designs:

```{r}
power.prop.test(p1 = 0.5, p2 = 0.6, power = 0.8)$n   # n per group
```

```{r}
#| eval: false
library(pwr)
pwr.t.test(d = 0.5, power = 0.8)   # n per group for a two-sample t-test
```

**The trinity on real fitted models.** You rarely code the statistics by hand. For a likelihood-ratio test between two nested fits use **`lmtest::lrtest`**, and for a Wald test on coefficients use **`aod::wald.test`**:

```{r}
#| eval: false
library(lmtest); library(aod)
lrtest(modelNull, modelFull)                              # likelihood-ratio
wald.test(b = coef(fit), Sigma = vcov(fit), Terms = 2)    # Wald
```
:::

# Summary

| Test | Statistic | Fits needed | Strength |
|---|---|---|---|
| **Wald** | $I(\hat\theta)(\hat\theta - \theta_0)^2$ | MLE only | Cheap; assumes a symmetric likelihood |
| **Score** | $U(\theta_0)^2 / I(\theta_0)$ | Null model only | No need to fit the alternative |
| **Likelihood-ratio** | $2[\ell(\hat\theta) - \ell(\theta_0)]$ | Both models | Most accurate; uses the true likelihood shape |

| Concept | Meaning |
|---|---|
| Null $H_0$ / alternative $H_1$ | The claim on trial vs everything else |
| p-value | $P(\text{data this extreme}\mid H_0)$ — *not* $P(H_0\text{ true})$ |
| Significance level $\alpha$ | Pre-chosen threshold = Type I error rate |
| Power $1-\beta$ | Probability of rejecting a *false* null |
| CI–test duality | The CI is the set of null values a test would not reject |

All of this is the likelihood machinery from earlier in the section, simply turned from *"what is $\theta$?"* to *"is $\theta = \theta_0$ believable?"* — the two questions every frequentist analysis ends up asking.
