---
title: "Likelihood, Score and Fisher Information"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

On the previous page we said statistics is the **inverse problem**: given data, recover the model. The **likelihood function** is the precise mathematical tool that lets us do this. Everything in frequentist parametric inference — estimates, standard errors, confidence intervals, tests — flows from it.

This page builds the chain of ideas step by step:

$$
\text{likelihood} \;\to\; \text{log-likelihood} \;\to\; \text{score} \;\to\; \text{MLE} \;\to\; \text{Fisher information} \;\to\; \text{standard error}.
$$

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart LR
  L["Likelihood<br/>L(θ)"] --> LL["Log-likelihood<br/>ℓ(θ)"]
  LL -->|"differentiate"| S["Score<br/>U(θ) = ℓ'(θ)"]
  S -->|"set = 0 and solve"| MLE["MLE<br/>θ̂"]
  LL -->|"second derivative"| I["Information<br/>curvature of ℓ"]
  I -->|"1 / √I"| SE["Standard error<br/>of θ̂"]
```

# The likelihood function

Suppose our data $x_1, \dots, x_n$ are modelled as coming from a distribution with probability (mass or density) function $f(x \mid \theta)$, where $\theta$ is the unknown parameter we want to recover. Another way of looking at this is that we have a **family of distributions** $\{f(x \mid \theta) : \theta \in \Theta\}$, and we want to find the member of that family that best explains the data we actually observed. (Throughout, we assume the observations are independent and identically distributed, written $X_1, \dots, X_n \sim^{iid} f(\cdot \mid \theta)$ — drawn separately from the same distribution.)

For a *fixed* value of $\theta$, the probability of seeing this particular data set (assuming the observations are independent) is

$$
f(x_1, \dots, x_n \mid \theta) = \prod_{i=1}^{n} f(x_i \mid \theta).
$$

The key move is to **flip what we treat as variable**. The data is already observed and fixed; the parameter is what we do not know. So we read the same expression as a function of $\theta$:

$$
\boxed{\,L(\theta) = \prod_{i=1}^{n} f(x_i \mid \theta)\,}
$$

::: {.callout-important title="Likelihood is not probability"}
$L(\theta)$ has the same formula as a probability, but it is **not** a probability distribution over $\theta$. We are not saying "$\theta$ is random." We are asking: *for each candidate value of $\theta$, how plausibly would it have produced the data we actually saw?* It does **not** integrate to 1 over $\theta$, and $\theta$ is still a fixed unknown — exactly the frequentist stance from the previous page.
:::

**What it represents:** $L(\theta)$ measures how well each candidate parameter value *explains the observed data*. A large $L(\theta)$ means "this $\theta$ makes the data we saw look unsurprising." This is precisely the backwards reasoning we wanted: high likelihood points to plausible models.

# The log-likelihood

In practice we almost never work with $L(\theta)$ directly. Instead we take its logarithm:

$$
\boxed{\,\ell(\theta) = \log L(\theta) = \sum_{i=1}^{n} \log f(x_i \mid \theta)\,}
$$

We do this for three reasons:

1. **Products become sums.** The $\prod$ turns into a $\sum$, which is far easier to differentiate.
2. **Numerical stability.** Multiplying many small probabilities together underflows to zero on a computer; adding their logs does not.
3. **The maximiser is unchanged.** Because $\log$ is strictly increasing, the $\theta$ that maximises $\ell(\theta)$ is exactly the $\theta$ that maximises $L(\theta)$. We lose nothing by switching.

**What it represents:** the same information as the likelihood — how well each $\theta$ explains the data — but on an additive, well-behaved scale.

# The score function

To find the $\theta$ that best explains the data, we look for the peak of $\ell(\theta)$. As with any optimisation, we differentiate. The derivative of the log-likelihood with respect to $\theta$ is called the **score function**:

$$
\boxed{\,U(\theta) = \frac{\partial \ell(\theta)}{\partial \theta}\,}
$$

**What it represents:** the score is the **slope** of the log-likelihood. It tells us, at any candidate $\theta$:

- a **positive** score means $\ell$ is still rising — larger $\theta$ would explain the data better;
- a **negative** score means we have gone too far — smaller $\theta$ is better;
- a **zero** score means we are at a turning point — a candidate for the best explanation.

A useful and important fact is that, when evaluated at the *true* parameter, the score has mean zero:
$$
\mathbb{E}[\,U(\theta)\,] = 0.
$$
On average the data does not pull the estimate in any particular direction away from the truth — a property we will lean on shortly.

::: {.callout-note title="In a nutshell: why the score averages to zero"}
At the *true* $\theta$ you are already standing on the peak of the curve, on average. A given data set might nudge the slope a little uphill (pushing your estimate higher) or a little downhill (pushing it lower), but it is **equally likely to nudge either way**. Average over all the data sets the truth could produce and those nudges cancel — so the expected slope is exactly zero.
:::

# The maximum likelihood estimate (MLE)

The **maximum likelihood estimate**, written $\hat{\theta}$, is the value of $\theta$ that maximises the likelihood — equivalently, where the score is zero:

$$
\boxed{\,U(\hat{\theta}) = 0\,} \qquad \text{(and } \ell''(\hat{\theta}) < 0 \text{, confirming a maximum).}
$$

**What it represents:** $\hat{\theta}$ is our single best point estimate — the parameter value that makes the observed data *most plausible*. It is the frequentist's answer to "what model produced this data?" Sometimes we can solve $U(\hat{\theta}) = 0$ by hand; when we cannot, we solve it numerically (for example with the Newton–Raphson method from the *Mathematics to Code* section).

# Fisher information

A point estimate alone is not enough — we also need to know **how trustworthy it is**. This is where *information* comes in, and it is the deepest idea on this page.

Look again at the shape of the log-likelihood near its peak:

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart LR
  A["Sharp, narrow peak<br/>large curvature"] --> A2["the data strongly<br/>pins down θ →<br/>HIGH information,<br/>SMALL standard error"]
  B["Flat, broad peak<br/>small curvature"] --> B2["many θ explain the data<br/>almost equally well →<br/>LOW information,<br/>LARGE standard error"]
```

The **curvature** of the log-likelihood at its peak — how sharply it bends — measures how much the data tells us about $\theta$. Curvature is the *second* derivative, so we define:

## Observed information

$$
\boxed{\,J(\theta) = -\frac{\partial^2 \ell(\theta)}{\partial \theta^2}\,}
$$

The minus sign makes $J$ positive at a maximum (where the second derivative is negative). Evaluated at the MLE, $J(\hat{\theta})$ is the **observed information**: the curvature of the log-likelihood for *this particular data set*.

## Expected (Fisher) information

If instead we average the curvature over all data sets the model could produce, we get the **expected** or **Fisher information**:

$$
\boxed{\,I(\theta) = \mathbb{E}\!\left[-\frac{\partial^2 \ell(\theta)}{\partial \theta^2}\right] = \mathbb{E}\!\left[\,U(\theta)^2\,\right]\,}
$$

The second equality (the variance of the score, since the score has mean zero) is a remarkable identity: the *average curvature* equals the *variability of the slope*. Both capture the same thing.

**What it represents:** Fisher information quantifies **how much the data tells us about $\theta$**. High information ⇒ a sharp likelihood peak ⇒ the data strongly constrains $\theta$ ⇒ a precise estimate. Low information ⇒ a flat peak ⇒ many values of $\theta$ are nearly as plausible ⇒ an imprecise estimate. Information also **adds up across independent observations**: $n$ independent data points carry $n$ times the information of one, which is exactly why bigger samples give sharper estimates.

# Putting it together: from information to a standard error

The reason Fisher information matters so much is the central result of this whole section. For large samples, the MLE is approximately normally distributed:

$$
\hat{\theta} \;\overset{\text{approx.}}{\sim}\; \mathcal{N}\!\left(\theta,\; \frac{1}{I(\theta)}\right).
$$

::: {.callout-note title="In a nutshell: these are stated results, not skipped steps"}
Two facts on this page are presented as *results to use*, not derived here: the identity $I(\theta) = \mathbb{E}[U(\theta)^2]$ (curvature equals the variance of the score), and the normal approximation $\hat\theta \sim \mathcal N(\theta, 1/I(\theta))$. If the proofs feel missing, you have not lost the plot — they are justified on the **Asymptotics** page (via the Central Limit Theorem). Take them on trust for now.
:::

In words: the MLE is approximately unbiased, and its **variance is the reciprocal of the Fisher information**. Therefore the **standard error** of the estimate is

$$
\boxed{\,\operatorname{se}(\hat{\theta}) \approx \frac{1}{\sqrt{I(\hat{\theta})}}\,}
$$

This closes the loop:

- more data, or a sharper likelihood, ⇒ **more information**;
- more information ⇒ **smaller standard error**;
- smaller standard error ⇒ a **more precise estimate** and a **narrower confidence interval**.

An approximate 95% confidence interval follows immediately:
$$
\hat{\theta} \pm 1.96 \times \operatorname{se}(\hat{\theta}).
$$

# More than one parameter: the vector case

::: {.callout-warning title="Prerequisites — and a safe exit"}
This section uses **vectors, matrices, and the matrix inverse**. If you are comfortable with those, read on. If not, you can safely **skip to the worked example below** — everything in the rest of the course that you need follows from the single-parameter intuition (curvature ⇒ precision) you already have. The matrix version is the same idea wearing bigger clothes, not a new idea.
:::

So far $\theta$ has been a single number. Most real models have **several** parameters at once — a Normal has a mean *and* a variance, a Gamma has a shape *and* a rate, a regression has a whole vector of coefficients. The good news is that *nothing conceptual changes*: every object above simply gains a dimension. We now write the parameter as a vector $\boldsymbol\theta = (\theta_1, \dots, \theta_p)$, and the scalars become vectors and matrices.

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart LR
  S1["Score: a single derivative<br/>U(θ) = ℓ'(θ)"] --> S2["Score vector (gradient):<br/>one partial derivative<br/>per parameter"]
  I1["Information: a single number<br/>I(θ) = −ℓ''(θ)"] --> I2["Information MATRIX:<br/>all second partials,<br/>p × p"]
```

## The score becomes a gradient vector

Instead of one derivative, we take a **partial derivative with respect to each parameter**. The score is now a *vector* — the gradient of the log-likelihood:
$$
\mathbf{U}(\boldsymbol\theta) =
\begin{pmatrix} \dfrac{\partial \ell}{\partial \theta_1} \\[2mm] \vdots \\[1mm] \dfrac{\partial \ell}{\partial \theta_p} \end{pmatrix}.
$$
The MLE $\hat{\boldsymbol\theta}$ is the point where **every** component is zero simultaneously, $\mathbf{U}(\hat{\boldsymbol\theta}) = \mathbf{0}$ — a *system* of $p$ equations in $p$ unknowns. This is exactly why multi-parameter models usually need a numerical optimiser: you can rarely solve that system by hand.

## The information becomes a matrix

The curvature is no longer a single second derivative but a whole **matrix** of them — the *Hessian* of the log-likelihood, averaged:
$$
\mathbf{I}(\boldsymbol\theta) = \mathbb{E}\!\left[-\frac{\partial^2 \ell}{\partial \theta_j\, \partial \theta_k}\right]_{j,k=1,\dots,p}.
$$
The **diagonal** entries measure how sharply each parameter is pinned down on its own; the **off-diagonal** entries measure how the parameters' estimates are *correlated* — something a one-parameter analysis cannot see. A large off-diagonal means the data struggles to tell two parameters apart (they trade off against each other).

## Standard errors come from the inverse matrix

The reciprocal $1/I$ becomes a **matrix inverse**. The asymptotic result generalises to a multivariate normal:
$$
\hat{\boldsymbol\theta} \;\overset{\text{approx.}}{\sim}\; \mathcal{N}_p\!\left(\boldsymbol\theta,\; \mathbf{I}(\boldsymbol\theta)^{-1}\right).
$$
The matrix $\mathbf{V} = \mathbf{I}(\hat{\boldsymbol\theta})^{-1}$ is the **variance–covariance matrix** of the estimates. Each parameter's standard error is the square root of the corresponding diagonal entry:
$$
\boxed{\,\operatorname{se}(\hat{\theta}_j) \approx \sqrt{\big[\mathbf{I}(\hat{\boldsymbol\theta})^{-1}\big]_{jj}}\,}
$$

::: {.callout-important title="Invert first, then take the diagonal"}
A common mistake is to take $1/\sqrt{I_{jj}}$ — inverting the diagonal entries directly. That is **wrong** unless the off-diagonals are zero. You must invert the *whole* matrix first and only then read off the diagonal, because each parameter's uncertainty depends on how it is entangled with the others.
:::

In short: **score → gradient vector, information → matrix, divide → matrix inverse.** Every idea on this page carries over; it just wears bigger clothes. The *Simulation* page works a full two-parameter example end to end, fitting a Gamma distribution with `optim` and reading the standard errors off the inverse Hessian exactly as described here.

# A worked example in R

Let us make all of this concrete with the simplest possible model: **Bernoulli data** (coin flips), where each observation is 1 (heads) with probability $\theta$ and 0 (tails) otherwise.

For a sample with $k$ heads out of $n$ flips, the pieces work out neatly by hand:

| Quantity | Formula |
|---|---|
| Log-likelihood | $\ell(\theta) = k\log\theta + (n-k)\log(1-\theta)$ |
| Score | $U(\theta) = \dfrac{k}{\theta} - \dfrac{n-k}{1-\theta}$ |
| MLE | $\hat{\theta} = \dfrac{k}{n}$ (the sample proportion) |
| Fisher information | $I(\theta) = \dfrac{n}{\theta(1-\theta)}$ |

Now let us verify this numerically. We will simulate some flips and find the MLE both algebraically and by optimising the log-likelihood directly.

```{r}
#| warning: false
#| message: false

set.seed(1)

# Simulate n coin flips with a true (but "unknown") bias
trueTheta <- 0.7
n <- 200
flips <- rbinom(n, size = 1, prob = trueTheta)
k <- sum(flips)              # number of heads observed

# The log-likelihood as a function of theta
ll <- function(theta) {
  k * log(theta) + (n - k) * log(1 - theta)
}

# 1. MLE by formula: the sample proportion
mle <- k / n

# 2. MLE by numerical optimisation of the log-likelihood
opt <- optimise(ll, interval = c(0.001, 0.999), maximum = TRUE)
mleNum <- opt$maximum

c(trueTheta = trueTheta,
  mle       = mle,
  mleNum    = mleNum)
```

The algebraic MLE and the numerical maximiser agree, and both land close to the true value of 0.7. Now the **Fisher information** and the **standard error**:

```{r}
# Fisher information evaluated at the MLE
info <- n / (mle * (1 - mle))

# Standard error = 1 / sqrt(information)
se <- 1 / sqrt(info)

# Approximate 95% confidence interval
ci95 <- mle + c(-1, 1) * 1.96 * se

list(fisherInfo = info,
     se         = se,
     ci95       = ci95)
```

Finally, we can *see* the curvature idea by plotting the log-likelihood. The peak sits at the MLE, and the **sharpness** of that peak is exactly the Fisher information — the quantity that becomes our standard error.

```{r}
#| fig-align: center

grid <- seq(0.4, 0.95, length.out = 400)

plot(grid, ll(grid), type = "l", lwd = 2,
     col = "steelblue",
     xlab = expression(theta),
     ylab = expression(paste("log-likelihood  ", ell(theta))),
     main = "Log-likelihood for the coin-flip example")

abline(v = mle,       col = "firebrick", lwd = 2, lty = 2)
abline(v = trueTheta, col = "darkgreen", lwd = 2, lty = 3)

legend("bottomright",
       legend = c("log-likelihood",
                  expression(paste("MLE  ", hat(theta))),
                  expression(paste("true  ", theta))),
       col = c("steelblue", "firebrick", "darkgreen"),
       lwd = 2, lty = c(1, 2, 3), bty = "n")
```

Try increasing `n` in the simulation: the peak becomes **sharper**, the Fisher information **grows**, and the standard error **shrinks** — the data is telling us more and more about $\theta$, exactly as the theory promises.

::: {.callout-tip title="Going further 🚀"}
You rarely have to do the optimisation *and* the standard errors by hand. Hand a negative log-likelihood to `stats4::mle` (or the friendlier `bbmle::mle2`) and you get `coef()`, `vcov()`, `confint()` and a `summary()` for free:

```{r}
#| eval: false
library(stats4)
nll <- function(theta) -(k * log(theta) + (n - k) * log(1 - theta))
fit <- mle(nll, start = list(theta = 0.5))
vcov(fit)      # variance = 1 / information
confint(fit)   # profile-likelihood interval, not just Wald
```

And when differentiating the log-likelihood twice by hand is painful, let `numDeriv::hessian()` build the information matrix numerically — no calculus required:

```{r}
#| eval: false
library(numDeriv)
J <- numDeriv::hessian(nll, x = mle)   # observed information
solve(J)                               # variance-covariance matrix
```
:::

# Summary

| Object | Symbol | What it represents |
|---|---|---|
| **Likelihood** | $L(\theta)$ | How well each $\theta$ explains the observed data |
| **Log-likelihood** | $\ell(\theta)$ | The same, on an additive, numerically stable scale |
| **Score** | $U(\theta) = \ell'(\theta)$ | The slope of $\ell$; points toward better $\theta$ |
| **MLE** | $\hat{\theta}$ | The $\theta$ that best explains the data ($U(\hat\theta)=0$) |
| **Observed information** | $J(\hat{\theta}) = -\ell''(\hat{\theta})$ | Curvature of $\ell$ for *this* data set |
| **Fisher information** | $I(\theta) = \mathbb{E}[-\ell''(\theta)]$ | Expected curvature — how much the data tells us about $\theta$ |
| **Standard error** | $1/\sqrt{I(\hat{\theta})}$ | The precision of the estimate |

These six quantities are the engine of frequentist parametric inference. Every estimate, standard error, confidence interval and likelihood-ratio test you meet later is built from exactly this chain.