---
title: "Worked Examples: The Whole Pipeline End to End"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

Every previous page introduced one link in a chain. This page joins them all up on **real data**. We will take two genuine datasets that ship with R and run the *complete frequentist pipeline* on each: choose a model, estimate its parameters two ways, quantify the uncertainty, test a hypothesis, and — crucially — **check whether the model actually fits**. No new theory; just the machinery from the rest of the section, applied honestly.

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart LR
  M["Pick a model<br/>f(x | θ)"] --> E["Estimate θ<br/>MoM &amp; MLE"]
  E --> U["Quantify uncertainty<br/>Fisher info → SE → CI"]
  U --> T["Test a hypothesis<br/>Wald / score / LR"]
  T --> C["Check the fit<br/>plots &amp; diagnostics"]
  C --> D["Conclusion"]
```

The two case studies are chosen to make a point. The first looks like a textbook success — until the diagnostics reveal it is not. The second is a two-parameter model where the algebra runs out and `optim` takes over. Together they show that fitting a model is easy; *trusting* it is the real work.

# Case study 1 — Great discoveries per year (Poisson)

R's built-in `discoveries` dataset records the number of "great" inventions and scientific discoveries in each year from 1860 to 1959. Counts of rare events in fixed intervals are the textbook setting for the **Poisson** distribution, $f(y \mid \lambda) = e^{-\lambda}\lambda^{y}/y!$, with a single rate parameter $\lambda$.

```{r}
#| warning: false
#| message: false

y <- as.numeric(discoveries)
n <- length(y)

c(nYears = n, totalDiscoveries = sum(y),
  mean = mean(y), variance = var(y))
```

Let us first just *look* at the data — always the first step.

```{r}
#| fig-align: center

plot(1860:1959, y, type = "h", lwd = 2, col = "steelblue",
     xlab = "Year", ylab = "Number of great discoveries",
     main = "Great discoveries per year, 1860–1959")
points(1860:1959, y, pch = 19, col = "steelblue", cex = 0.6)
abline(h = mean(y), col = "firebrick", lwd = 2, lty = 2)
legend("topright", bty = "n", lwd = 2, lty = c(1, 2),
       col = c("steelblue", "firebrick"),
       legend = c("yearly count", expression(paste("mean  ", bar(y)))))
```

## Estimation: method of moments and maximum likelihood

The Poisson is the rare case where the two estimators from the intro page **coincide exactly**. The theoretical mean is $\lambda$, so the method of moments sets $\hat\lambda_{\text{MoM}} = \bar y$. And the log-likelihood
$$
\ell(\lambda) = -n\lambda + \Big(\textstyle\sum_i y_i\Big)\log\lambda - \sum_i \log(y_i!)
$$
has score $U(\lambda) = \sum_i y_i/\lambda - n$, which is zero at $\hat\lambda_{\text{MLE}} = \bar y$ too. Both recipes land on the sample mean.

```{r}
ll <- function(lambda) sum(dpois(y, lambda, log = TRUE))

lamMom <- mean(y)                 # method of moments
lamHat <- mean(y)                 # MLE (identical here)

# Fisher information I(λ) = n/λ  ⇒  se = sqrt(λ̂ / n)
info <- n / lamHat
se   <- 1 / sqrt(info)

c(lamMom = lamMom, lamMle = lamHat,
  fisherInfo = info, stdError = se)
```

## Uncertainty: two confidence intervals

We build both intervals from the asymptotics page — the symmetric **Wald** interval and the shape-respecting **likelihood-ratio** interval — and plot the log-likelihood with both overlaid.

```{r}
# Wald interval
waldCI <- lamHat + c(-1, 1) * 1.96 * se

# Likelihood-ratio interval: invert 2[ℓ(λ̂) − ℓ(λ)] ≤ χ²₁,₀.₉₅
grid <- seq(lamHat - 4 * se, lamHat + 4 * se, length.out = 4000)
W    <- 2 * (ll(lamHat) - sapply(grid, ll))
lrCI <- range(grid[W <= qchisq(0.95, df = 1)])

rbind(Wald = waldCI, LikelihoodRatio = lrCI)
```

```{r}
#| fig-align: center

llVals <- sapply(grid, ll)
plot(grid, llVals, type = "l", lwd = 2, col = "steelblue",
     xlab = expression(lambda), ylab = expression(ell(lambda)),
     main = "Log-likelihood for the discovery rate, with 95% CIs")

cutoff <- ll(lamHat) - qchisq(0.95, 1) / 2
abline(h = cutoff,   col = "grey50",   lty = 3)
abline(v = lamHat,   col = "black",   lwd = 2)
abline(v = lrCI,     col = "firebrick", lwd = 2, lty = 2)
abline(v = waldCI,   col = "darkorange",lwd = 2, lty = 4)
legend("bottomright", bty = "n", lwd = 2, lty = c(1, 2, 4),
       col = c("black", "firebrick", "darkorange"),
       legend = c(expression(hat(lambda)), "Likelihood-ratio CI", "Wald CI"))
```

Because $n = 100$ and the likelihood is nearly parabolic, the two intervals almost coincide — the regime where asymptotics are reliable.

## A hypothesis test

Suppose a historian claims the long-run rate is exactly **three discoveries per year**. We test $H_0: \lambda = 3$ against $H_1: \lambda \neq 3$ using all three members of the trinity from the testing page.

```{r}
lambda0 <- 3

# Wald: distance of the MLE from λ₀, scaled by info at the MLE
wWald  <- info * (lamHat - lambda0)^2

# Score: steepness at λ₀, scaled by info at λ₀  (U(λ₀) = Σy/λ₀ − n,  I(λ₀) = n/λ₀)
U0     <- sum(y) / lambda0 - n
wScore <- U0^2 / (n / lambda0)

# Likelihood-ratio: vertical drop from the peak to λ₀
wLR    <- 2 * (ll(lamHat) - ll(lambda0))

round(rbind(
  Wald            = c(stat = wWald,  p = 1 - pchisq(wWald,  1)),
  Score           = c(stat = wScore, p = 1 - pchisq(wScore, 1)),
  LikelihoodRatio = c(stat = wLR,    p = 1 - pchisq(wLR,    1))
), 4)
```

All three statistics are modest and the p-values comfortably exceed $0.05$: the data is **compatible** with a rate of three per year. We do not reject $H_0$ — though, per the testing page, that is "fail to reject," not proof the rate is exactly three.

## The crucial step: does the Poisson model even fit?

A Poisson distribution has a defining fingerprint: **its mean equals its variance.** We already saw the sample variance is noticeably *larger* than the sample mean — a warning sign called **overdispersion**. The clearest way to see the consequence is to overlay the fitted Poisson frequencies on the observed ones.

```{r}
#| fig-align: center

kmax <- max(y)
observed <- as.numeric(table(factor(y, levels = 0:kmax)))
expected <- n * dpois(0:kmax, lamHat)

mat <- rbind(Observed = observed, `Fitted Poisson` = expected)
barplot(mat, beside = TRUE, names.arg = 0:kmax,
        col = c("steelblue", "firebrick"),
        xlab = "Discoveries in a year", ylab = "Number of years",
        main = "Observed counts vs the fitted Poisson model")
legend("topright", bty = "n", fill = c("steelblue", "firebrick"),
       legend = c("Observed", "Fitted Poisson"))
```

The model is **too thin in the tails**: it expects almost no years with 0 or with 6+ discoveries, yet both occur more often than predicted. This is exactly what overdispersion looks like. A quick dispersion check confirms it:

```{r}
c(mean = mean(y), variance = var(y),
  dispersionRatio = var(y) / mean(y))   # ≈ 1 for a true Poisson
```

::: {.callout-important title="The estimate was fine; the model was not"}
Every number we computed — the MLE, its standard error, the confidence intervals, the hypothesis test — was *arithmetically* correct. But they all assumed a Poisson, and the Poisson does not describe this data well. A richer model (e.g. the **negative binomial**, which allows the variance to exceed the mean) would be more honest. This is the central lesson of a worked example: **estimation is mechanical, but it only means something once the model passes its diagnostics.**
:::

::: {.callout-tip title="Going further 🚀"}
Diagnosing overdispersion is only half the job — let's *fix* it. The **negative binomial** adds a dispersion parameter so the variance can exceed the mean, and `MASS::glm.nb` fits it in one line. Comparing models by **AIC** (lower is better) makes the improvement concrete:

```{r}
#| eval: false
library(MASS)
poisFit  <- glm(y ~ 1, family = poisson)        # the model we just fitted
nbFit    <- glm.nb(y ~ 1)                        # negative binomial alternative
AIC(poisFit, nbFit)                              # lower AIC = better fit
```

For fitting *distributions* (rather than regressions), `fitdistrplus::fitdist(y, "nbinom")` gives the fit, standard errors, and a panel of diagnostic plots via `plot(fit)` — a great one-stop package to explore.
:::

## A bootstrap sanity check

Finally, the simulation page told us we can always check an asymptotic standard error by resampling. The bootstrap distribution of $\hat\lambda$ should be centred at the estimate with spread matching our formula `se`.

```{r}
#| fig-align: center

set.seed(1)
B <- 5000
bootLambda <- replicate(B, mean(sample(y, n, replace = TRUE)))

hist(bootLambda, breaks = 40, col = "lightsteelblue", border = "white",
     main = expression(paste("Bootstrap distribution of  ", hat(lambda))),
     xlab = expression(hat(lambda)))
abline(v = lamHat, col = "firebrick", lwd = 2)
abline(v = waldCI, col = "darkgreen", lwd = 2, lty = 2)

c(formulaSe = se, bootstrapSe = sd(bootLambda))
```

The two standard errors agree — the *uncertainty* calculation is trustworthy even though the *model* is imperfect. (Note the bootstrap resamples the real, overdispersed data, so it honestly reflects the true variability, slightly above the Poisson formula.)

# Case study 2 — Lengths of major rivers (Gamma)

Our second dataset, `rivers`, gives the lengths (in miles) of 141 major rivers in North America. Lengths are positive and **right-skewed**, so a natural model is the two-parameter **Gamma** distribution with shape $\alpha$ and rate $\beta$ — exactly the intractable, multi-parameter case the simulation page tackled with `optim`.

```{r}
#| fig-align: center

x <- rivers
n <- length(x)

hist(x, breaks = 20, col = "lightsteelblue", border = "white", freq = FALSE,
     xlab = "River length (miles)", ylab = "Density",
     main = "Lengths of 141 major North American rivers")
rug(x)
```

## Fitting two parameters with `optim`

There is no closed form for the Gamma MLEs, so we minimise the **negative** log-likelihood numerically, seeding the search with **method-of-moments** starting values (the intro page's promised use of MoM) and asking for the Hessian to get standard errors.

```{r}
#| warning: false
#| message: false

nll <- function(par) {
  shape <- par[1]; rate <- par[2]
  if (shape <= 0 || rate <= 0) return(1e10)
  -sum(dgamma(x, shape = shape, rate = rate, log = TRUE))
}

# Method-of-moments start:  shape = mean²/var,  rate = mean/var
m <- mean(x); v <- var(x)
start <- c(shape = m^2 / v, rate = m / v)

fit <- optim(start, nll, hessian = TRUE)

shapeHat <- fit$par[1]; rateHat <- fit$par[2]
V  <- solve(fit$hessian)        # variance–covariance = inverse observed information
se <- sqrt(diag(V))

ci <- cbind(estimate = fit$par,
            lower = fit$par - 1.96 * se,
            upper = fit$par + 1.96 * se)
round(ci, 4)
```

Both parameters are estimated with their Wald intervals. The fitted **shape is above 1**, which matters: a Gamma with shape $> 1$ has its density rising from zero, so the model expects very few extremely short rivers — sensible, since no major river has near-zero length.

How well does the fitted curve track the data?

```{r}
#| fig-align: center

hist(x, breaks = 20, col = "lightsteelblue", border = "white", freq = FALSE,
     xlab = "River length (miles)", ylab = "Density",
     main = "River lengths with the fitted Gamma density")
xs <- seq(min(x), max(x), length.out = 400)
lines(xs, dgamma(xs, shape = shapeHat, rate = rateHat),
      col = "firebrick", lwd = 3)
legend("topright", bty = "n", lwd = 3, col = "firebrick",
       legend = "fitted Gamma")
```

## Seeing the two-parameter likelihood

With two parameters the log-likelihood is a **surface**, not a curve. A contour plot shows the peak (the MLE) and how the parameters trade off — the elongated, tilted contours reveal that shape and rate are **strongly correlated**, the off-diagonal effect the likelihood page warned about.

```{r}
#| fig-align: center

shapeSeq <- seq(shapeHat - 3.5 * se[1], shapeHat + 3.5 * se[1], length.out = 90)
rateSeq  <- seq(rateHat  - 3.5 * se[2], rateHat  + 3.5 * se[2], length.out = 90)
surf <- outer(shapeSeq, rateSeq,
              Vectorize(function(s, r) -nll(c(s, r))))

contour(shapeSeq, rateSeq, surf, nlevels = 25,
        xlab = "shape  α", ylab = "rate  β",
        main = "Log-likelihood surface for the Gamma fit")
points(shapeHat, rateHat, pch = 19, col = "firebrick", cex = 1.3)
text(shapeHat, rateHat, labels = "  MLE", pos = 4, col = "firebrick")
```

We can also collapse the surface onto a single parameter via the **profile likelihood**: for each candidate shape, maximise over the rate. This is the honest one-dimensional picture of the evidence about the shape, and its $\chi^2_1$ cut-off gives a confidence interval that respects the likelihood's true form.

```{r}
#| fig-align: center

shapeGrid <- seq(shapeHat - 3.5 * se[1], shapeHat + 3.5 * se[1], length.out = 150)
profile <- sapply(shapeGrid, function(s) {
  o <- optimize(function(r) -sum(dgamma(x, shape = s, rate = r, log = TRUE)),
                interval = c(1e-6, 10 * rateHat))
  -o$objective
})

plot(shapeGrid, profile, type = "l", lwd = 2, col = "steelblue",
     xlab = "shape  α", ylab = "profile log-likelihood",
     main = "Profile likelihood for the Gamma shape")
abline(h = max(profile) - qchisq(0.95, 1) / 2, col = "grey50", lty = 3)
abline(v = shapeHat, col = "firebrick", lwd = 2, lty = 2)
abline(v = 1,         col = "darkgreen", lwd = 2, lty = 4)
legend("bottomright", bty = "n", lwd = 2, lty = c(2, 4, 3),
       col = c("firebrick", "darkgreen", "grey50"),
       legend = c(expression(hat(alpha)), "shape = 1 (Exponential)",
                  "95% LR cut-off"))
```

## A nested hypothesis test: is the simpler Exponential enough?

The Exponential distribution is just a Gamma with **shape fixed at 1**. That makes "is the Exponential adequate?" a clean hypothesis test, $H_0: \alpha = 1$, between two *nested* models — the ideal job for the **likelihood-ratio test**. We compare the full two-parameter Gamma fit against the one-parameter Exponential fit.

```{r}
llGamma <- -fit$value                                   # full model
rateExp <- 1 / mean(x)                                  # Exponential MLE
llExp   <- sum(dexp(x, rate = rateExp, log = TRUE))     # restricted model

LR <- 2 * (llGamma - llExp)
pValue <- 1 - pchisq(LR, df = 1)                        # 1 extra parameter

c(llExponential = llExp, llGamma = llGamma,
  lrStatistic = LR, pValue = pValue)
```

The likelihood-ratio statistic is large and the p-value tiny, so we **reject the Exponential** in favour of the Gamma: the extra shape parameter earns its place. This matches the profile plot above, where shape $= 1$ falls well below the $\chi^2_1$ cut-off line — the test and the picture tell the same story, just as the CI–test duality promises.

## Diagnostics: a quantile–quantile plot

A density overlay is reassuring but coarse. The sharper check is a **Q–Q plot**: sort the data and plot it against the quantiles the fitted Gamma predicts. Points on the diagonal mean the model captures the whole distribution; systematic departure — especially in the tail — flags where it fails.

```{r}
#| fig-align: center

theoretical <- qgamma(ppoints(n), shape = shapeHat, rate = rateHat)
empirical   <- sort(x)

plot(theoretical, empirical, pch = 19, col = "steelblue", cex = 0.7,
     xlab = "Gamma theoretical quantiles", ylab = "Observed river length",
     main = "Gamma Q–Q plot for river lengths")
abline(0, 1, col = "firebrick", lwd = 2)
```

The bulk of the points hug the line, so the Gamma is a solid description of typical rivers — but the longest rivers tend to drift **above** the diagonal, meaning the real upper tail is a little heavier than the Gamma allows. A perfectly honest report would note that the model is good for the body of the distribution but slightly underestimates the chance of an exceptionally long river.

## Bootstrapping the joint uncertainty

The Hessian-based standard errors are themselves asymptotic. Following the simulation page, we **bootstrap the entire two-parameter fit** — re-running `optim` on each resample — and scatter the resulting estimates. The cloud's shape *is* the joint sampling distribution, and its tilt is the shape–rate correlation made visible.

```{r}
#| warning: false
#| message: false
#| fig-align: center

set.seed(42)
B <- 800
boot <- matrix(NA, nrow = B, ncol = 2, dimnames = list(NULL, c("shape", "rate")))
for (b in 1:B) {
  xb <- sample(x, n, replace = TRUE)
  ob <- optim(fit$par, function(par) {
    if (any(par <= 0)) return(1e10)
    -sum(dgamma(xb, shape = par[1], rate = par[2], log = TRUE))
  })
  boot[b, ] <- ob$par
}

plot(boot[, "shape"], boot[, "rate"], pch = 19,
     col = adjustcolor("steelblue", 0.35), cex = 0.6,
     xlab = "shape  α", ylab = "rate  β",
     main = "Bootstrap distribution of the Gamma MLE")
points(shapeHat, rateHat, pch = 19, col = "firebrick", cex = 1.4)
legend("topleft", bty = "n", pch = 19, col = c("steelblue", "firebrick"),
       legend = c("bootstrap estimates", "MLE"))

# Bootstrap SEs next to the Hessian-based ones
rbind(hessianSe   = se,
      bootstrapSe = apply(boot, 2, sd))
```

The bootstrap standard errors closely track the Hessian-based ones — the asymptotic two-parameter fit is trustworthy here — and the diagonal stripe in the scatter is the strong positive correlation between shape and rate that the contour plot first hinted at.

# What every worked example has in common

Strip away the specifics and both case studies followed the **identical pipeline**, which is the whole of this section in one loop:

```{mermaid}
%%| echo: false
%%| fig-align: center

flowchart TD
  A["Look at the data"] --> B["Choose a model f(x | θ)"]
  B --> C["Estimate θ — MoM for a start, MLE for the answer"]
  C --> D["Uncertainty — Fisher information → SE → confidence interval"]
  D --> E["Test the question of interest — Wald / score / LR"]
  E --> F["CHECK THE FIT — overlays, Q–Q plots, dispersion, bootstrap"]
  F --> G{"Does it pass?"}
  G -->|"Yes"| H["Report estimate, CI and test"]
  G -->|"No"| B
```

| Step | Tool (and the page it came from) | Case 1 (Poisson) | Case 2 (Gamma) |
|---|---|---|---|
| Estimate | MoM & MLE *(intro, likelihood)* | $\hat\lambda = \bar y$ | `optim` on the NLL |
| Standard error | Fisher information *(likelihood)* | $\sqrt{\hat\lambda/n}$ | inverse Hessian |
| Confidence interval | Wald & LR *(asymptotics)* | both, nearly equal | Wald + profile |
| Hypothesis test | the trinity *(testing)* | $H_0:\lambda=3$, not rejected | $H_0:\alpha=1$, rejected |
| Model check | diagnostics & bootstrap *(simulation)* | overdispersion found | tail slightly heavy |

The estimates and intervals were the easy part — the same handful of formulas every time. The judgement lay in the **last row**: the Poisson's machinery worked perfectly on a model that did not fit, while the Gamma survived its diagnostics with only a minor caveat. That is the real skill frequentist statistics teaches — not turning the crank, but knowing whether to trust what comes out.
