---
title: "Python for R Users: Setup and Basics"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

You have spent this course learning R. **Python** is the other great language of data science, and a well-rounded statistician is fluent in both. This short section is a **bridge for R users**: rather than teaching programming from scratch, it maps what you *already know in R* onto Python, then introduces the two libraries that power data work — **NumPy** and **pandas** — so that the Python code in the Machine Learning section becomes readable and runnable. It also feeds directly into modules like ST236 (Python for Data Analytic Tasks).

::: {.callout-note title="In a nutshell: this is translation, not a fresh start"}
You are not learning to program again — you already can. You are learning to *say the same things in a new language*. An R vector becomes a NumPy array; a `data.frame` becomes a pandas DataFrame; `dplyr` becomes pandas. Keep your R knowledge in mind and read each Python idea as "the Python version of the R thing I know."
:::

# Getting Python running

To follow along on your own machine you need a Python installation with the scientific libraries. Two easy routes:

- **Google Colab** ([colab.research.google.com](https://colab.research.google.com)) — a free, browser-based notebook with NumPy and pandas already installed. Nothing to set up; ideal for experimenting.
- **A local install** — the **Anaconda** (or Miniconda) distribution bundles Python with NumPy, pandas and the rest of the scientific stack in one download.

The pages of this section actually *run* their Python code through R's **`reticulate`** package, which lets a single document execute both R and Python. If you want to reproduce that locally, install `reticulate` and point it at a Python with the libraries:

```{r}
#| eval: false
install.packages("reticulate")
library(reticulate)
py_install(c("numpy", "pandas"))   # install the libraries into reticulate's Python
```

Everything below this point is genuinely executed, so the output you see is what Python really returns.

# The five things that will trip up an R user

Before any code, internalise the handful of differences that cause 90% of beginner mistakes coming from R:

::: {.callout-important title="Five Python surprises for R users"}
1. **Indexing starts at 0**, not 1. The first element of a list is `x[0]`. And slices are **end-exclusive**: `x[0:2]` gives elements 0 and 1, *not* 2.
2. **Assignment is `=`**, not `<-`. (`<-` is not valid Python.)
3. **Indentation is part of the syntax.** Python has no `{ }`; a consistent indent (usually 4 spaces) defines a block. Get it wrong and the code errors.
4. **Methods, not functions.** Where R writes `mean(x)`, Python often writes `x.mean()` — the function "belongs to" the object and is called with a dot.
5. **The naming convention is `snake_case`** (`my_data`), not the `camelCase` or `dot.case` common in R.
:::

# Assignment and printing

R returns and prints the last value automatically; in Python you usually call `print()` to see output.

```{python}
x = 5
y = 10
print(x + y)
```

The same in R, for comparison:

```{r}
x <- 5
y <- 10
x + y
```

# Types

Python's basic types mirror R's: integers, floating-point numbers, strings and booleans. The `type()` function reports them (note Python's booleans are capitalised, `True`/`False`).

```{python}
print(type(7))          # int
print(type(7.5))        # float
print(type("stats"))    # str
print(type(True))       # bool
```

# Lists: Python's everyday container

A Python **list** is the closest everyday analogue of an R vector — an ordered collection — though it can hold mixed types like an R *list*. The crucial difference is **0-based, end-exclusive** indexing.

```{python}
my_list = [10, 20, 30, 40, 50]

print(my_list[0])      # first element  -> 10   (R would be my_list[1])
print(my_list[1:3])    # indices 1 and 2 -> [20, 30]   (end 3 excluded)
print(my_list[-1])     # last element   -> 50
print(len(my_list))    # length         -> 5

my_list.append(60)     # add to the end (a method!)
print(my_list)
```

::: {.callout-note title="In a nutshell: lists do not do maths"}
Beware: a plain Python list does **not** behave like an R vector under arithmetic. `[1, 2, 3] * 2` gives `[1, 2, 3, 1, 2, 3]` (it *repeats* the list), not `[2, 4, 6]`. For element-wise maths you need a **NumPy array** — which is exactly why NumPy exists, and the subject of the next page.
:::

# Dictionaries: the named-list analogue

A **dictionary** stores key–value pairs, like an R named list. You look values up by their key.

```{python}
prices = {"apple": 0.50, "banana": 0.30, "cherry": 0.80}

print(prices["apple"])        # look up by key -> 0.5
print(list(prices.keys()))    # the keys
print(list(prices.values()))  # the values
```

# Loops and conditionals

The logic is identical to R; only the syntax (a colon and indentation instead of braces) changes. Note `range(n)` counts `0, 1, …, n-1`.

```{python}
for i in range(4):        # 0, 1, 2, 3
    print(i)
```

```{python}
score = 72

if score >= 70:
    grade = "First"
elif score >= 60:
    grade = "2:1"
else:
    grade = "lower"

print(grade)
```

Python's logical operators are the words **`and`**, **`or`**, **`not`** (where R uses `&`, `|`, `!`), and equality is `==`.

# Functions

Defined with `def`, a colon, and an indented body. The `return` statement hands back the result (just like R's `return()`, but here it is the norm). Default arguments work as in R.

```{python}
def power(x, p=2):        # p defaults to 2
    return x ** p          # ** is exponentiation (R's ^)

print(power(5))            # 25
print(power(2, 10))        # 1024
```

# Importing libraries

Where R attaches a package with `library()`, Python **imports** a module, conventionally giving it a short alias. You then reach its functions through that alias with a dot.

```{python}
import numpy as np

print(np.sqrt(16))        # use numpy's sqrt via the alias 'np'  -> 4.0
```

This `import … as …` line will start almost every Python data script you write — `import numpy as np` and `import pandas as pd` are practically universal.

::: {.callout-tip title="Going further 🚀"}
A few idioms that make Python feel natural once you know R's equivalents: **list comprehensions** (`[x**2 for x in range(5)]`) are Python's compact answer to `sapply`/`vapply`; **f-strings** (`f"grade: {grade}"`) are its `paste0`/`sprintf`; and `help(np.sqrt)` (or `?` in a notebook) is its `?`. Because this document runs through `reticulate`, you can even pass objects between the languages — an R data frame `myData` is visible in Python as `r.myData`, and a Python object `result` is visible in R as `py$result`.
:::

# Where this section is going

:::{}
* **NumPy** — arrays and vectorised computing: the analogue of R's vectorised vectors and matrices, and the foundation of all numerical Python.
* **pandas** — DataFrames for data wrangling: the analogue of `data.frame`/tibble and `dplyr`, for filtering, transforming and summarising tabular data.
* **R vs Python** — a reference page of the key differences and gotchas, and guidance on when to reach for which language.
:::

With these three, the Python half of the Machine Learning section stops being decorative and becomes code you can read, run and adapt.
