---
title: "NumPy: Arrays and Vectorised Computing"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

You already think in terms of vectors: in R you write `c(1, 2, 3) * 2` and the whole thing scales at once, no loop required. Python's plain lists do **not** behave that way, but the **NumPy** library gives you back exactly the vectorised, fast, mathematical objects you are used to. This page is the translation guide: you know this in R, here is the NumPy version.

Throughout, we import NumPy under its conventional alias `np`:

```{python}
import numpy as np
np.__version__
```

# Why NumPy?

In R, arithmetic on a vector is *element-wise* and instant. Python's built-in `list` does something quite different: multiplying a list by a number **repeats** the list rather than scaling its contents.

```{python}
plain_list = [1, 2, 3]
plain_list * 2          # repetition, NOT element-wise maths!
```

Compare with R, where multiplication does what a statistician expects:

```{r}
plainVector <- c(1, 2, 3)
plainVector * 2          # element-wise: 2 4 6
```

The fix is the NumPy **array** (`ndarray`). It is the direct analogue of R's atomic vector: element-wise arithmetic, vectorised maths functions, and fast C-level loops under the hood.

```{python}
arr = np.array([1, 2, 3])
arr * 2                  # element-wise: array([2, 4, 6])
```

::: {.callout-note title="In a nutshell: a NumPy array is R's vector"}
A Python `list` is just a container and does not do maths element-wise (`[1,2,3] * 2` repeats it). Whenever you would reach for an R vector, reach for a **NumPy array** instead: `np.array([...])`. Everything else on this page follows from that one idea.
:::

# Creating arrays

There are a handful of constructors you will use constantly. Each one maps cleanly onto an R function you already know.

```{python}
np.array([1, 2, 3])      # like R's c(1, 2, 3)
```

`np.arange` is R's `seq` with the "by" argument: it generates a sequence with a fixed step. Note the **end is exclusive**, so this stops before 10.

```{python}
np.arange(0, 10, 2)      # like seq(0, 8, by = 2) -> 0 2 4 6 8
```

`np.linspace` is R's `seq` with `length.out`: give it a start, an end (this time **inclusive**), and how many points you want, evenly spaced.

```{python}
np.linspace(0, 1, 5)     # like seq(0, 1, length.out = 5)
```

`np.zeros` and `np.ones` build arrays filled with a constant, much like R's `rep`. For a 2-D shape you pass a tuple `(rows, cols)`.

```{python}
np.zeros(3)              # like rep(0, 3)
```

```{python}
np.ones((2, 2))          # a 2x2 matrix of ones; like matrix(1, 2, 2)
```

Here is the R side of the same ideas for reference:

```{r}
seq(0, 8, by = 2)        # arange
seq(0, 1, length.out = 5)# linspace
rep(0, 3)                # zeros
```

# Vectorised arithmetic and broadcasting

This is where NumPy feels like home. Operators and maths functions apply element-wise, exactly as in R, with no explicit loop.

```{python}
arr = np.array([1, 2, 3, 4])
arr * 2                  # scale every element
```

```{python}
arr + arr                # element-wise addition of two arrays
```

```{python}
np.sqrt(arr)             # vectorised maths function, like sqrt() in R
```

When you combine an array with a single number, NumPy **broadcasts** the scalar across every element. This is precisely R's recycling of a length-1 vector.

```{python}
arr + 10                 # 10 is broadcast to every element
```

The R analogue is identical in spirit:

```{r}
x <- c(1, 2, 3, 4)
sqrt(x)                  # vectorised
x + 10                   # length-1 vector recycled
```

# Aggregations (summaries)

Reducing an array to a single summary number works just like R's `mean`, `sum`, and friends, with one notable difference: in NumPy these are usually **methods** attached to the array (written `arr.mean()`), whereas in R they are **functions** that take the vector as an argument (`mean(x)`).

```{python}
arr = np.array([2, 4, 6, 8, 10])
arr.mean()               # R: mean(x)
```

```{python}
arr.sum(), arr.min(), arr.max()   # R: sum(x), min(x), max(x)
```

```{python}
arr.std()                # population standard deviation by default
```

The R equivalents read left-to-right as function calls:

```{r}
x <- c(2, 4, 6, 8, 10)
mean(x); sum(x); min(x); max(x); sd(x)
```

::: {.callout-tip title="Going further 🚀"}
A subtle gotcha: `arr.std()` divides by `n` (the population formula), whereas R's `sd(x)` divides by `n - 1` (the sample formula). To match R, use `arr.std(ddof=1)`, where `ddof` is the "delta degrees of freedom" subtracted from `n`.
:::

# 2-D arrays (matrices)

A nested list becomes a 2-D array, the analogue of R's `matrix`. The `.shape` attribute reports the dimensions as a `(rows, cols)` tuple, much like R's `dim()`.

```{python}
m = np.array([[1, 2],
              [3, 4]])
m.shape                  # (2, 2); like dim() in R
```

Element access uses square brackets with a comma, and here is the single biggest trap for an R user: **NumPy indices start at 0**, not 1. So the top-left element is `m[0, 0]`, and `m[0, 1]` is the first row, second column.

```{python}
m[0, 1]                  # row 0, column 1 -> 2 (R's m[1, 2])
```

Matrix multiplication uses the `@` operator, which plays the role of R's `%*%`. (Plain `*` would multiply element-wise instead.)

```{python}
m @ m                    # matrix product; R: m %*% m
```

::: {.callout-note title="In a nutshell: counting starts at 0, and slice ends are exclusive"}
NumPy is **0-indexed**: the first element is `arr[0]`, so an R index `i` becomes a NumPy index `i - 1`. Slices are also **half-open** (the end is excluded): `arr[1:3]` returns the elements at positions 1 and 2 only, *not* position 3. R's `x[2:3]` (inclusive) is NumPy's `arr[1:3]`.
:::

```{python}
a = np.array([10, 20, 30, 40, 50])
a[1:3]                   # indices 1 and 2 -> array([20, 30]); 3 is excluded
```

# Boolean masking

Filtering with a logical condition is one of R's most loved features (`x[x > 2]`), and NumPy does the same thing with the same syntax. A comparison produces a boolean array, which you then use as an index.

```{python}
arr = np.array([1, 2, 3, 4, 5])
arr > 2                  # a boolean mask: array([False, False, True, True, True])
```

```{python}
arr[arr > 2]             # keep only the elements where the mask is True
```

`np.where` is the vectorised "if-else": choose between two values per element based on a condition, much like R's `ifelse`.

```{python}
np.where(arr > 2, arr, 0)   # keep values > 2, replace the rest with 0
```

The R version is reassuringly familiar:

```{r}
x <- c(1, 2, 3, 4, 5)
x[x > 2]                 # logical indexing
ifelse(x > 2, x, 0)      # vectorised if-else
```

# Going further: speed and the wider ecosystem

::: {.callout-tip title="Going further 🚀"}
Just as in R, the golden rule is **vectorise instead of looping**. A Python `for` loop over millions of elements is slow because each step runs in the Python interpreter; a NumPy expression like `arr * 2` runs the loop in compiled C, exactly as R's vectorised operations drop into C/Fortran. So prefer `arr.sum()` over a hand-written loop, and prefer boolean masks over filtering one element at a time.

NumPy is also the **foundation of the scientific Python stack**. The `scipy` library builds on it for optimisation, integration, and statistics; **pandas** data frames store their columns as NumPy arrays; and **scikit-learn** expects NumPy arrays as the inputs and outputs of its models. Learn NumPy well and the rest of the ecosystem falls into place.
:::

# Summary: R to NumPy cheat sheet

| Task | R | NumPy |
|------|------|-------|
| Create a vector/array | `c(1, 2, 3)` | `np.array([1, 2, 3])` |
| Sequence by step | `seq(0, 8, by = 2)` | `np.arange(0, 10, 2)` |
| Sequence by length | `seq(0, 1, length.out = 5)` | `np.linspace(0, 1, 5)` |
| Repeat a constant | `rep(0, 3)` | `np.zeros(3)` |
| Scale element-wise | `x * 2` | `arr * 2` |
| Vectorised maths | `sqrt(x)` | `np.sqrt(arr)` |
| Mean / sum | `mean(x)` / `sum(x)` | `arr.mean()` / `arr.sum()` |
| Standard deviation | `sd(x)` | `arr.std(ddof=1)` |
| Matrix | `matrix(...)` | `np.array([[...], [...]])` |
| Dimensions | `dim(m)` | `m.shape` |
| Element (1- vs 0-indexed) | `m[1, 2]` | `m[0, 1]` |
| Matrix multiply | `m %*% m` | `m @ m` |
| Logical filter | `x[x > 2]` | `arr[arr > 2]` |
| Vectorised if-else | `ifelse(c, a, b)` | `np.where(c, a, b)` |

You now have the core translation: a NumPy array is R's vector, arithmetic is element-wise, summaries are methods, and indexing starts at 0. With that mental model, most NumPy code reads like the R you already know.
