---
title: "pandas: DataFrames for Data Wrangling"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
    mathjax: true
engine: knitr
---

If you have spent any time with R, you have spent time with `data.frame`s and tibbles, and almost certainly with **dplyr** and **tidyr** for wrangling them. Python's answer to all of this is a single library: **pandas**. The idea of this page is simple — almost everything you already know how to do in the tidyverse has a direct pandas counterpart. We will lean on that knowledge throughout, so think of this as a translation exercise rather than learning something from scratch.

One small but important habit to flag up front: Python uses **snake_case** for names (`mean_speed`), whereas in this course we tend to write R in **camelCase** (`meanSpeed`). Keep an eye on that as you read.

## The DataFrame: pandas' tibble

::: {.callout-note title="In a nutshell: the DataFrame is pandas' data.frame, and pandas is dplyr + tidyr"}
A pandas `DataFrame` is a rectangular table of columns, exactly like an R `data.frame` or tibble. The **pandas** library provides the verbs to wrangle it — filtering, selecting, mutating, arranging, grouping and summarising — playing the combined role of **dplyr** and **tidyr**. By convention you import it as `pd`.
:::

Let us build a small DataFrame inline. We will use a handful of cars, with a numeric `speed` column and a categorical `brand` column to group by.

```{python}
import pandas as pd
import numpy as np

cars = pd.DataFrame({
    "name": ["Mini", "Golf", "Polo", "Civic", "Jazz", "Corolla"],
    "brand": ["BMW", "VW", "VW", "Honda", "Honda", "Toyota"],
    "speed": [110, 130, 105, 125, 100, 120],
    "doors": [3, 5, 5, 5, 5, 4],
})

cars
```

For contrast, here is the same table built in base R. It should feel very familiar:

```{r}
cars <- data.frame(
  name  = c("Mini", "Golf", "Polo", "Civic", "Jazz", "Corolla"),
  brand = c("BMW", "VW", "VW", "Honda", "Honda", "Toyota"),
  speed = c(110, 130, 105, 125, 100, 120),
  doors = c(3, 5, 5, 5, 5, 4)
)

cars
```

The two are conceptually identical: named columns of equal length, each with its own type.

## Inspecting a DataFrame

Before wrangling, you usually want to look at the shape and structure of your data. In R you reach for `head()`, `dim()`, `names()`, `str()` and `summary()`. pandas has a near one-to-one set of tools.

```{python}
cars.head()        # first few rows, like head() in R
```

```{python}
cars.shape         # (rows, columns) — like dim()
```

```{python}
cars.columns       # column names — like names()
```

```{python}
cars.dtypes        # type of each column — like sapply(df, class)
```

```{python}
cars.describe()    # numeric summary — like summary() on numeric columns
```

Note that `describe()` by default only summarises the numeric columns (`speed` and `doors`), which is usually what you want.

## Selecting columns and rows

This is the first place where Python's conventions differ a little, so it is worth slowing down.

**Columns.** Selecting a single column with `df['col']` returns a **Series** (a one-dimensional labelled vector, like a single R column). Selecting a list of columns with `df[['a', 'b']]` returns a **DataFrame**.

```{python}
cars["speed"]              # a Series (one column)
```

```{python}
cars[["name", "speed"]]    # a DataFrame (note the double brackets)
```

The double brackets `[[ ]]` are easy to remember if you think of the inner `[ ]` as building a Python *list* of column names.

**Rows.** Here is the big one: pandas indexing is **0-based**, whereas R is 1-based. So the first row is at position `0`, not `1`. Use `.iloc[...]` to select by integer **position** and `.loc[...]` to select by **label** (the row's index).

```{python}
cars.iloc[0]               # FIRST row, by position (0-based!)
```

```{python}
cars.loc[0, "name"]        # cell at label 0, column "name"
```

::: {.callout-note title="In a nutshell: iloc is positional, loc is by label, and counting starts at 0"}
`df.iloc[...]` is integer-position based (think "i for integer"); `df.loc[...]` is label based. Because our DataFrame has the default integer index, the labels happen to be `0, 1, 2, ...`, so the two can look similar here — but they are different tools. And remember: the first row is position `0`, not `1`.
:::

## Filtering rows (dplyr `filter`)

To keep rows matching a condition, you build a boolean mask and index with it — `df[df['x'] > 3]`. This is the pandas version of dplyr's `filter()`.

```{python}
cars[cars["speed"] > 115]
```

There is also a tidier, more readable alternative, `query()`, which lets you write the condition as a string and refer to columns by bare name — much like `filter()` does with non-standard evaluation:

```{python}
cars.query("speed > 115")
```

The dplyr equivalent for reference:

```{r}
# library(dplyr)
# cars |> filter(speed > 115)
subset(cars, speed > 115)   # base R version, no packages needed
```

## Creating new columns (dplyr `mutate`)

Adding a column is as simple as assigning to it, `df['new'] = ...`. This mutates the DataFrame in place.

```{python}
cars["kph_to_mph"] = cars["speed"] * 0.621
cars
```

If you prefer a non-mutating, pipe-friendly style (closer to dplyr's `mutate()`, which returns a new data frame), use `assign()`, which returns a **new** DataFrame:

```{python}
cars.assign(fast = cars["speed"] > 115)
```

In dplyr you would write `cars |> mutate(fast = speed > 115)`.

## Sorting (dplyr `arrange`)

Use `sort_values()` to order rows by one or more columns. The `ascending=False` argument is the analogue of wrapping a column in `desc()` in dplyr.

```{python}
cars.sort_values("speed", ascending=False)
```

The dplyr equivalent would be `cars |> arrange(desc(speed))`.

## Grouping and summarising (dplyr `group_by |> summarise`)

This is the heart of data wrangling, and pandas maps onto the dplyr workflow neatly. `df.groupby('group')` is `group_by()`, and you then apply an aggregation.

```{python}
cars.groupby("brand")["speed"].mean()
```

That gives the mean speed per brand — the equivalent of:

```{r}
# cars |> group_by(brand) |> summarise(mean_speed = mean(speed))
aggregate(speed ~ brand, data = cars, FUN = mean)
```

For several summaries at once, use `.agg(...)`, just as you would list several expressions inside `summarise()`:

```{python}
cars.groupby("brand").agg(
    mean_speed = ("speed", "mean"),
    n_cars     = ("speed", "size"),
)
```

The `("speed", "mean")` syntax names the output column `mean_speed` and says "take the `speed` column and apply `mean`" — closely mirroring `summarise(mean_speed = mean(speed), n_cars = n())`.

## dplyr → pandas translation table

Keep this handy. Once these patterns click, most wrangling tasks are a quick mental lookup.

| Task                       | dplyr                                   | pandas                                         |
|----------------------------|-----------------------------------------|------------------------------------------------|
| Filter rows                | `filter(speed > 115)`                   | `df[df["speed"] > 115]` or `df.query("speed > 115")` |
| Select columns             | `select(name, speed)`                   | `df[["name", "speed"]]`                        |
| Create / modify a column   | `mutate(fast = speed > 115)`            | `df.assign(fast = df["speed"] > 115)` or `df["fast"] = ...` |
| Arrange rows               | `arrange(desc(speed))`                  | `df.sort_values("speed", ascending=False)`     |
| Group and summarise        | `group_by(brand) |> summarise(m = mean(speed))` | `df.groupby("brand").agg(m=("speed", "mean"))` |
| The pipe                   | `df |> ... |> ...`                      | method chaining: `df.method().method()...`     |

The last row is the key conceptual bridge: where R threads a data frame through a sequence of verbs with the pipe `|>`, pandas threads a DataFrame through a sequence of **methods** with the dot `.`.

## Going further

::: {.callout-tip title="Going further 🚀"}
**Method chaining is the pipe.** Because each pandas verb returns a DataFrame, you can chain them just like a dplyr pipeline. The `reset_index()` at the end turns the grouping labels back into an ordinary column (otherwise `brand` would live in the index):

```{python}
(
    cars
    .groupby("brand")
    .agg(mean_speed=("speed", "mean"))
    .reset_index()
    .sort_values("mean_speed", ascending=False)
)
```

Reading top to bottom, this is `cars |> group_by(brand) |> summarise(...) |> arrange(desc(mean_speed))`.

**Real data.** Instead of building tables by hand, you would normally load them with `pd.read_csv("file.csv")` — the analogue of `read.csv()` or `readr::read_csv()`.

**Joins.** pandas has `df.merge(other, on="key", how="left")` (and `df.join(...)`), which correspond to dplyr's `left_join()`, `inner_join()` and friends.

**reticulate interop.** This very document mixes R and Python via **reticulate**, and you can pass data between them. In an R chunk, `py$cars` reads the Python `cars` DataFrame as an R data frame; in a Python chunk, `r.some_r_dataframe` reads an R data frame as a pandas DataFrame. For example, an R chunk could run `head(py$cars)` to inspect the Python object we built above, and a Python chunk could run `r.cars` to pull in the base-R `cars` data frame. This lets you wrangle in whichever language fits the task best.
:::

You now have the core of pandas: it is the same mental model as the tidyverse, expressed with brackets and dots instead of verbs and pipes. When you get stuck, ask yourself "what would I write in dplyr?" and then translate — the table above will usually get you there.
