---
title: "Data types"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

Before we start coding in R, we need to be familiar with how R stores data - i.e. we need to review the data types in R. The following table states the 6 data types and what they store.

| Data type name  | Stores  |
|--------|--------|
| Logical or Boolean  | TRUE or FALSE    |
| Numeric   | a real number, including decimal values    |
| Integer | an integer number (i.e. no decimal values)    |
| Complex | a complex number with both a real and an imaginary part, e.g. 3+2i |
| Character | character or string values, for example a letter 'A', typically surrounded by `''` or `""`  |
| Raw | raw bytes, stored in hexadecimal |

: Data types {.hover}

:::{.callout-note}
Numeric and integer are different data types! When inputting data, any integer value will be stored as a numeric data type unless we add an `L` to the end. For example, `5` will be stored as a numeric data type but `5L` will be stored as an integer. 
:::

# What data type is this variable?

We can check what data type a variable is using `class()`. Alternatively, we can confirm if a variable is an integer, logical or a character using `is.integer()`, `is.logical()` and `is.character()` respectively.

```{r}
###| Character
character <- "a banana"

class(character)
is.character(character)

###| Numeric v integer
number <- 5
integer <- as.integer(5)

class(number)
is.integer(number)

class(integer)
is.integer(integer)

###| Decimal value
decimal <- 0.5

class(decimal)
is.integer(decimal)

###| Logical value
is.logical(TRUE)
is.logical(FALSE)
```

# Coercion (changing data type)

We can force a variable to switch data type using `as.numeric()`, `as.logical()`, `as.character()` and `as.integer()`.

```{r}
###| Coercing a boolean variable
as.numeric(TRUE)
as.numeric(FALSE)
as.character(TRUE)
as.character(FALSE)
```

```{r}
###| Coercing a character variable
as.numeric("a") # this will produce NA with a warning, not an error!
```

```{r}
###| Coercing a character variable
as.character("a")

###| Coercing a numeric variable
as.numeric(2) 
as.character(2)

###| Coercing to a logic variable
as.logical(1)
as.logical(0)

as.logical("TRUE")
as.logical("FALSE")
```

# Check your understanding

Identify the following variable's data type with and without using R: 

:::{} 
* a = 2025 
* b = "3.14" 
* c = pi 
* d = TRUE 
* e = c(1,0) 
* f = 2022L 

:::

```{r}
#| code-fold: true

###| Answers
class(2025)
class("3.14")
class(pi)
class(TRUE)
class(c(1,0))
class(2022L)
```
