---
title: "Custom functions"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

A function is a piece of code written to carry out a specific task. It can take in inputs and may return one or more values.

The 3 main components are:

<div>

-   The formals(): the list of arguments/inputs of the function
-   The body(): the code in the function
-   The environment(): the map of the location of the function's variable

</div>

A function follows the following general structure:

<div>

-   function_name \<- function(arguments) {function body}

</div>

Custom functions are used to incorporate sets of instructions that you want to use repeatedly, or that are easier to store so they can be called when needed.

Functions are created using function() and are stored as R objects. They can also be passed to other functions as arguments (as input) and can be nested (functions can be inside other functions).

```{r}
###| Function example 1
add <- function(x,y){
  x + y
}

add(2,4)

###| Function example 2
square <- function(x){
  x^2
}

square(7)
```

::: {.callout-tip title="Going further 🚀"}
Functions can do a lot more than take fixed inputs. Arguments can have **defaults**, a function can return **several values** at once by bundling them in a list, `...` lets you **pass extra arguments straight through** to another function, and `\(x)` is a shorthand for a quick **anonymous function**.

```{r}
###| Default argument: base defaults to 10 unless given
logBase <- function(x, base = 10) log(x, base = base)
logBase(1000)          # uses base 10
logBase(8, base = 2)   # override it

###| Return multiple values via a named list
stats <- function(x) list(mean = mean(x), sd = sd(x))
stats(c(1, 2, 3, 4))

###| ... passes extra arguments through to mean()
trimmedMean <- function(x, ...) mean(x, ...)
trimmedMean(c(1, 2, 3, 100), trim = 0.25)

###| Anonymous function: handy with sapply()
sapply(1:4, \(x) x^2)
```
:::

# Check your understanding

Write functions that do the following things:

<div>

-   Create a function which returns integer division of x by y and the remainder
-   Create a function which finds the area and circumference of a circle given a radius r.
-   Create a function that calculates the factorial of some number n.
    -   Hint: make sure that n is an integer!

</div>

```{r}
#| code-fold: true
###| Answer

modulus <- function(x,y) {
  remainder <- x %% y
  multiple <- x %/% y
  print(paste("x modulo y is", multiple, "with remainder", remainder))
}

modulus(6,3)
```

```{r}
#| code-fold: true
###| Answer

circle <- function(r){
  area <- r^2 * pi
  circumference <- 2*r*pi
  print(paste("Area is",area, "and the circumference is",circumference))
}

circle(5)
```

```{r}
#| code-fold: true
###| Answer

factorial <- function(x) {
  
  if (x >= 0 && x%%1 == 0) {

    factorial <- 1

    while (x >= 1) {
      factorial <- factorial * x
      x <- x - 1
    }

    print(factorial)
  } else {
    print(paste(x," is not a non-negative integer"))
  }
}

factorial(2) 
```

