---
title: "Operators"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

We will begin by familiarising ourselves with the operations in R, where

> "an operator is a character, or characters, that determine what action is to be performed or considered",

according to the [BBC Bitesize website](https://www.bbc.co.uk/bitesize/guides/znh6pbk/revision/9). These will form the base building blocks of everything we will do in the future. 

We will start with arithmetic operators ($+$,$-$,$\times$,$\div$), then go to comparison operators ($==$,$\ne$,$\geq$,$\leq$) and finally logical operators (and, or).

# Arithmetic operators

The 4 base arithmetic operators are addition (`+`), subtraction (`-`), multiplication (`*`) and division (`/`). The following are some base examples of how these can be used.

```{r}
###| Addition
3+10

###| Subtraction
6-2

###| Multiplication
5*15

###| Division
23/7
```

:::{.callout-note}
It is important to note that R follows the traditional [BODMAS](https://thirdspacelearning.com/blog/what-is-bodmas/#:~:text=BODMAS%20is%20an%20acronym%20to,%2DAddition%2C%20S%2DSubtraction.) order of operations. As such, we can use brackets to control the order in which operations are performed. 
:::

```{r}
###| BODMAS
(1+7)/2
1 + 7/2
3*5+2/3
3*(5+2)/3
```

We can also do exponents (to the power of) using `^` or `**`, and we can find the modulo using `%%` and `%/%`

```{r}
###| Exponents 
5^2
5**2

4^(1/2)
4**0.5
```

```{r}
###| Modulo
10 %% 2 # remainder of 10 divided by 2
10 %% 4

10 %/% 2 # how many times 2 goes into 10
10 %/% 4
```

# Comparison operators

Comparison operators compare 2 values to one another and will return TRUE or FALSE.

The 2 operators that can be used for any data type are equal to `==` and not equal to `!=`. 

We can also compare two numbers using less than `<`, less than or equal to `<=`, greater than `>`, and greater than or equal to `>=`.

```{r}
###| Equal to
"a banana" == "a banana"
"a banana" == "a fruit"

7 == 5
5 == 5

###| Not equal to
"a banana" != "a banana"
"a banana" != "a fruit"

7 != 5
5 != 5
```

```{r}
###| Less than
5 < 7 
7 < 5
5 < 5 # note that this will return FALSE!

###| Greater than
5 > 7 
7 > 5
5 > 5 # note that this will return FALSE!

###| Less than or equal to
5 <= 7 
7 <= 5
5 <= 5 # note that this will return TRUE!

###| Greater than or equal to
5 >= 7 
7 >= 5
5 >= 5 # note that this will return TRUE!
```

# Logical operators

Logical operators allow us to combine things or negate something.

If we want to find the *negative*, we use the `!` operator. For example, we may want to say something is not true. we do this with `!TRUE`. Or we may want to check if 5 is not less than 7, `!(5<7)`. 

```{r}
###| Negation of TRUE FALSE
TRUE
!(TRUE)

FALSE
!(TRUE)
```

```{r}
###| Example of negation on relational operators
5<7
!(5<7)

5==7
!(5==7)
```

We can also combine operators using AND, `&`. For example, we may want to check that `5<7` and `10<12`. Alternatively, we may want to allow A or B to be true. We do this with `|`. 

```{r}
###| And on TRUE FALSE
TRUE & TRUE
FALSE & FALSE
TRUE & FALSE

###| And on relational operators example
(7>5) & (10<12)
(7<5) & (10>12)
(7<5) & (10<12)
```

```{r}
###| OR on TRUE FALSE
TRUE | TRUE
FALSE | FALSE
TRUE | FALSE

###| OR on relational operators example
(7>5) | (10<12)
(7<5) | (10>12)
(7<5) | (10<12)
```

# Check your understanding

The following are some exercises to test your understanding about operations.

:::{}
*   Use R to determine the area of a circle with diameter of 20cm.
    +   Hint: `pi` is recognised in R.
*   Calculate the cube root of 14 $\times$ 0.51.
    +   Hint: think carefully about the order of operations.
* Test the following statement in R: Not (4 squared is greater than or equal to 15, or (the sum of 7 and 10 is less than 16)).
    +   Hint: the output should be FALSE.
:::

(The code is hidden to give you a chance to practise. However, if you toggle the `>`, example code will appear.)

```{r}
#| code-fold: true

###| Answers
(10^2)*pi
(14 * 0.51)^(1/3)
!( 4^2>=15 | 7+10 < 16)
```
