---
title: "Random numbers"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

Inbuilt into R, there are many functions focused on statistical distributions. These are all stored in the 'stats' package, which is loaded in by default in R. However, for practise, (and to ensure everything will work), we will load in the package directly.

```{r}
###| Loading in stats package
library(stats)
```

We will begin by looking at generating a uniform number before doing on to look at continuous and discrete random variables.

::: {.callout-note title="In a nutshell: the r / d / p / q prefixes"}
Every distribution in R comes with four functions, named by a prefix plus the distribution's short name (`unif`, `norm`, `binom`, `pois`, …). You only need to recognise the prefix:

| Prefix | Example | Plain English |
|---|---|---|
| `r` | `runif(5)` | **r**andom — draw random values from the distribution |
| `d` | `dnorm(0)` | **d**ensity — the height of the curve at a point (how "likely" that value is) |
| `p` | `pnorm(1.96)` | **p**robability — the area to the *left* (the cumulative probability, $P(X \le x)$) |
| `q` | `qnorm(0.975)` | **q**uantile — the reverse of `p`: give it a probability, it returns the cut-off value |

So `runif` *makes* numbers, and `punif` *tells you the probability* of getting one that small. If you remember just this table, the rest of the page is filling in the details.
:::

# Random uniform numbers

To generate a random number, we are going to use the uniform distribution. Let $X\sim \text{Uniform}[a,b]$. Then, the probability density function of X is

$$
f(x) = \begin{cases}
\frac{1}{b-a} & \text{for } a\leq x\leq b, \\
0 & \text{for } x<a \text{ or } x>b
\end{cases}
$$

We are going to use the inbuilt R functions to generate a random number, x, in the interval $[a,b]$ and find the probability we generated x.

For simplicity, we will demonstrate this with 2 uniform random variables: $X\sim \text{Uniform}[0,1]$ and $Y\sim \text{Uniform}[1,5]$.

```{r}
###| Generate a random number, x
x <- runif(n = 1, min = 0, max = 1)
x

###| Cumulative probability P(X <= x)
probX <- punif(x, min = 0, max = 1)
probX

###| Generate a random number, y
y <- runif(n = 1, min = 1, max = 5)
y

###| Cumulative probability P(X <= x)
probY <- punif(y, min = 1, max = 5)
probY
```

We can also find the value, $x$, that corresponds to the probability $P(X\leq x) = a$.

```{r}
###| a = 0.5, X ~ Uniform[0,1]
qunif(0.5, min = 0, max = 1)

###| a = 0.5, X ~ Uniform[0,5]
qunif(0.5, min = 0, max = 5)

###| a = 0.75, X ~ Uniform[0,5]
qunif(0.75, min = 0, max = 5)
```

We can also find the density of values $x$ and $y$.

```{r}
###| Density of x, X ~ Uniform[0,1]
x <- runif(n = 1, min = 0, max = 1)
dunif(x, min = 0,max = 1)

###| Density of y, Y ~ Uniform[1,5]
y <- runif(n = 1, min = 1, max = 5)
dunif(y, min = 1,max = 5)
```

Up until now, we have only generated 1 number at a time. However, we can generate several numbers at once by changing the \$n = \$ argument. This works for all of R's random-generation functions (rnorm, rbinom, rpois, runif, ...): given n > 1, each generates n values.

```{r}
###| Multiple values at once
runif(n = 5, min = 0, max = 1)
runif(n = 3, min = -1, max = 1)
```

# Binomial distribution

Now, suppose we want to sample from a binomial distribution, i.e. we want to generate a random number so that the probability $X=x$ is the same as the probability $Y=x$, where $Y \sim \text{Binomial}(m,p)$. We do this using 'rbinom()', 'dbinom()', 'pbinom()' and 'qbinom()'. Note that $n$ refers to the number of samples while size refers to the number of trials!

```{r}
###| Generate a random value from X ~ Binom(m = 5, p = 0.2)
x <- rbinom(n=1, size = 5, prob = 0.2)
x

###| Cumulative probability P(X <= x)
probX <- pbinom(x, size = 5, prob = 0.2)
probX

###| f(x), probability mass, P(X = x)
densityX <- dbinom(x, size = 5, prob = 0.2)
densityX

###| Quantile of 0.75
qbinom(p = 0.75, size = 5, prob = 0.2)
```

# Check your understanding - Normal and Poisson distribution

In the above sections, we have gone over how to sample from Uniform and Binomial distribution. To check your understanding, try the following:

<div>

-   Sample 5 values from a Poisson distribution with rate $\lambda = 5$ and find the probability and density of each value. Return this as a dataframe.
    -   Hint: type ?rpois into the command line or into the help section.
-   Sample 5 values from a Normal distribution with mean 0 and variance 4. Find the density of each value, to 3 significant figures.
    -   Hint: type ?rnorm and ?round into the command line or into the help section.

</div>

```{r}
#| code-fold: true
###| Generate and store 5 values from Poisson
sample <- rpois(n = 5, lambda = 5)
sample

###| Find probability, density and quantile
df <- data.frame("sample" = sample,
                 "prob" = rep(NA,length(sample)),
                 "density" = rep(NA,length(sample)))


for (i in 1:length(sample)) {
  df$prob[i] <- ppois(df$sample[i], lambda = 5)
  df$density[i] <- dpois(df$sample[i], lambda = 5)
}

df
```

```{r}
#| code-fold: true
###| Generate Normal distribution
norm <- rnorm(n = 5, mean = 0, sd = 2)
norm

###| Find density of each variable
for (i in 1:length(norm)){
  print(paste(round(norm[i],3)," has density ", round(dnorm(norm[i],mean = 0, sd = 2),3)))
}

# Alternatively, we can do this in one step without the loop as dnorm() 
# allows vector input
dnorm(norm, mean=0,sd=2) 
```
