---
title: "Variables"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---


Thus far, we have covered some very important content. However, it is not in a practical, useable form. Instead of having to rewrite values each time, we want to be able to have the computer remember them (store them) so that we can recall them. We do this using variables.

We can assign a value to a named variable using `<-` (the most common operator), `=`, or `->`. (Note that we use `=` for assigning variables and `==` for check if equal!) We can check what our variable is by outputting it.

```{r}
###| Assigning a variable
x <- 5
y = TRUE

###| Outputting a variable
x
y
```

We can use all our previous operators on variables. The following are just some examples of how we can do this.

```{r}
###| Assigning variables
a = 1
b = 2
c = 3
d = 4
e = 5

###| Arithmetic operators
d %% b
c + d
a * e - d

###| Relational operators
d < b
c > d
c == d

###| Logical operators 
(d < b) & (c > d)
(d < b) | (c > d)
!(d < b)
```

# Check your understanding

The following are some exercises to test your understanding about variables.

<div>

-   Define variables called `velocity`, `time` and `acceleration` with 5, 10 and 9.8 respectively. Calculate the displacement.
    -   Hint: $s = ut+0.5at^2$.
-   Define variables `height` and `width` of a rectangle with initial values 5 and 10. Calculate the area of the rectangle.
-   Extension: given $f(x)=\frac{1}{\sqrt{2\pi}}\exp{(-\frac{1}{2}x^2)}$, find $f(5)$.
    -   Hint: look at sqrt() and exp().

</div>

::: {.callout-note title="In a nutshell: build the formula one step at a time"}
A long formula is much easier if you translate it piece by piece into named variables. For $f(x)=\frac{1}{\sqrt{2\pi}}\exp{(-\frac{1}{2}x^2)}$:

1. Compute the **exponent** first: `expPart <- exp(-0.5 * x^2)`.
2. Then the **constant** out front: `constant <- 1/sqrt(2*pi)`.
3. Finally **multiply** them together: `fX <- constant * expPart`.

Breaking it into steps means each part is easy to check before you combine them.
:::

```{r}
#| code-fold: true

###| Answers
velocity <- 5
time <- 10
acceleration <- 9.8
displacement <- velocity * time + 0.5 * acceleration * time^2
displacement

height <- 5
width <- 10
area <- height * width
area

###| Extension question
x <- 5
fX <- 1/sqrt(2*pi)*exp(-0.5*x^2)
fX
```
