---
title: "Control Structures & Code Abstraction"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

## Abstraction
Abstraction is the process of hiding unnecessary detail so we can focus on the essential structure of a problem. In programming, this means solving a smaller, simpler problem first and then using that solution as a building block.

We use high-level languages because they let us write logic without thinking about hardware-level details. R and Python are far from the machine, so we can spend our energy on the algorithm rather than memory addresses.

Programming languages exist on a spectrum of abstraction. The level refers to how close the syntax is to human reasoning versus how close it is to the computer hardware.

|Attribute| Low Level (Binary, Assembly)| Higher Level (R, Python)|
|---------|-----------------------------|-------------------------|
|Abstraction| Very low; interacts directly with memory/CPU | High; hides hardware complexities|
|Execution Time|Extremely fast; minimal translation required|Slower; requires interpretation or heavy runtime environments|
|Memory Management|Manual (allocating/freeing memory addresses)|Automatic|
|Development Time|Long; requires writing many lines of verbose code|Short; concise libraries solve problems instantly|

From this it is evident that there is a trade off between efficiency and ease of use. Languages like C or C# are lower level than Python or R but are much more pedantic about small issues like semi-colons and declaring variable types.

For this course and for general understanding we will use very high level code syntax as a baseline and explore more efficient versions when needed (The lowest level versions I will leave up your interpretation). For this course examples will mostly be in R or Python since most uses of pseudo logic that people use is very similar to these programming languages.

## The Three Pillars of Logic

Every algorithm ever written relies on exactly three fundamental control structures to direct the flow of execution: **Sequence**, **Selection**, and **Iteration**. 

## Sequence

**Sequence** means execution happens sequentially, one line after another, from top to bottom. The computer must completely finish the current statement before moving to the next one.

$$\text{Step 1: } x \leftarrow 5 \longrightarrow \text{Step 2: } y \leftarrow x + 10 \longrightarrow \text{Step 3: } \text{print}(y)$$

### Example: Linear Execution in R
```{r}
# Sequential operations
radius <- 7
area <- pi * radius^2
print(area)
```

---

## Selection

Selection (conditional branching) allows a program to choose different paths of execution based on whether a condition evaluates to TRUE or FALSE. There are multiple ways to use this to your advantage, including:
1. Adding extra steps in certain cases
2. Ignoring steps in certain cases
3. Mutually exclusive steps

Examples of each are shown below

### Adding New Steps
```{r}
# Assume condition is some arbitrary TRUE/FALSE statement.
# Using this we can run a block of code only when the condition holds.
condition <- FALSE
if (condition) {
  print("Condition was TRUE")
} else {
  print("Condition was FALSE")
}
```

### Skipping Steps
```{r}
x <- 5
if(x<=10){
  x <- 3*x+1
  x <- x**0.5
}
# This is a very simple example but conditioning can be used to ignore very complex code chunks where they are not applicable
print(x+17)
```

### Mutually Exclusive Steps
```{r}
x <- 7
if (x <= 6) {
  print("Less than 6")
} else if (x <= 10) {
  print("Between 7 and 10")
} else {
  print("11 or greater")
}
```

Selection is not about the exact syntax; it is about the idea that the code can branch into different paths.

---

## Iteration
This is the process of designing the algorithm to repeat sections of code. There are 2 main ways of doing this:
1. Count-controlled loops - loops that repeat a fixed number of times.
2. Condition-controlled loops - loops that repeat until a condition changes.

::: {.callout-important}
## Important Tip
There is a very clear reason why we iterate which is to control the sequence, but more so **to prevent writing the same code** which can compound the number of mistakes you make when writing. When you have all of the code in 1 place it is really easy to diagnose where things go wrong and you only have to edit the code in one place.
:::

### Count-controlled loops
This shows an example in both C# and R. Both do the same thing: start with a counter, test whether the loop should continue, execute the loop body, and then update the counter.
```{csharp}
//| eval: false
using System;

class Program {
    static void Main() {
        for (int i = 0; i < 5; i++) {
            Console.WriteLine(i);
        }
    }
}
```
```{r}
for (i in 1:5) {
  print(i)
}
```

### Condition Controlling
Unlike a count-controlled loop which knows exactly how many times it will run before it starts, a condition-controlled loop (`while` loop) continues indefinitely until a specific logical condition becomes `FALSE`. This is incredibly useful for statistical simulations or algorithms like Newton-Raphson where you do not know how many steps it will take to converge.

::: {.callout-warning}
 If the condition never evaluates to `FALSE`, the loop will run forever, crashing your program!
:::
```{r}
count <- 1
while (count < 10) {
  print(count)
  count <- count + 3
}
```

This form is useful when the loop should continue until some condition changes.

### Modular programming
Modular programming means breaking code into reusable pieces, such as functions. This is the best way to keep control logic manageable.

```{r}
zScore <- function(value, mean, sd) {
  z <- (value - mean) / sd
  return(z)
}

print(zScore(85, 70, 10))
print(zScore(60, 70, 10))
```

A function encapsulates a sequence of steps and makes the overall program easier to read, test, and maintain.

---

## Visualising Control Structures
It can be difficult to see how these three pillars interact. The flowchart below shows how Sequence, Selection, and Iteration form the core workflow of any algorithm.
```{mermaid}
%%| echo: false
flowchart TD
    subgraph Sequence
    direction TB
    A[Step 1] --> B[Step 2] --> C[Step 3]
    end

    subgraph Selection
    direction TB
    D{Condition?}
    D -- True --> E[Path A]
    D -- False --> F[Path B]
    end

    subgraph Iteration
    direction TB
    G[Start Loop] --> H{Condition Met?}
    H -- No --> I[Execute Code] --> H
    H -- Yes --> J[Exit Loop]
    end
```

---

## Putting It All Together

In practice, algorithms rarely use just one of these structures. They are nested. Here is a classic statistical example: simulating a coin flip to count how many heads we get in 10 flips.

* **Iteration:** We need to flip the coin 10 times.
* **Sequence:** In each flip, we generate a random number, check it, and update our count.
* **Selection:** We only add to our count *if* the result is a Head.

```r
# SEQUENCE: Initialise our starting variables
nFlips <- 10
heads <- 0

# ITERATION: Run the process 10 times
for (i in 1:nFlips) {

  # SEQUENCE: Generate a random number (0 or 1)
  flip <- sample(0:1, 1)

  # SELECTION: Check if it is a 1 (Head)
  if (flip == 1) {
    heads <- heads + 1
  }
}

# SEQUENCE: Print final result
print(paste("Total Heads:", heads))
```
---

## 💻 Try It Yourself: The FizzBuzz Challenge

This is a classic programming interview question. Write an R script that loops through the numbers 1 to 15. 
* If the number is divisible by 3, print "Fizz".
* If it is divisible by 5, print "Buzz".
* If it is divisible by both 3 and 5, print "FizzBuzz".
* Otherwise, just print the number.

**Hint:** You will need a `for` loop, an `if...else if` chain, and the modulo operator (`%%`) to check for remainders!

**Answer**
```{r}
#| eval: false
#| code-fold: true
for (i in 1:15) {
  if (i %% 3 == 0 && i %% 5 == 0) {
    print("FizzBuzz")
  } else if (i %% 3 == 0) {
    print("Fizz")
  } else if (i %% 5 == 0) {
    print("Buzz")
  } else {
    print(i)
  }
}
```

::: {.callout-tip title="Going further 🚀"}
The loop checks one number at a time. In R we can often replace such a loop with a **vectorised** expression that acts on the whole vector `1:15` at once — usually shorter and faster. `dplyr::case_when()` reads like the `if...else if` chain, testing each condition in order:

```{r}
#| eval: false
library(dplyr)
n <- 1:15
case_when(
  n %% 15 == 0 ~ "FizzBuzz",   # divisible by both 3 and 5
  n %% 3  == 0 ~ "Fizz",
  n %% 5  == 0 ~ "Buzz",
  TRUE         ~ as.character(n)
)
```

Pure base R can do the same with nested `ifelse()`, which also works element-by-element across the vector:

```{r}
n <- 1:15
ifelse(n %% 15 == 0, "FizzBuzz",
ifelse(n %% 3 == 0, "Fizz",
ifelse(n %% 5 == 0, "Buzz", as.character(n))))
```

Same logic, no explicit loop — the iteration is handled for you inside the vectorised call.
:::