---
title: "Data Structures and Algorithms"
format: 
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

## Why data structures matter

A data structure is a way of organising and storing information so a computer can use it efficiently. In statistics and data science, choosing the right structure makes your code easier to write, easier to read, and often much faster.

For example, a vector is perfect for a list of numbers, while a data frame is better for a table with many variables. The right structure also helps you choose the right algorithm for the task.

---

## Common data structures for statistics

### Vectors and arrays

A vector is an ordered collection of values of the same type. In R, vectors are the simplest data structure; in Python, the closest equivalent is a `numpy` array.

- Use a vector for a single numeric or character sequence.
- Use vector operations when the same calculation applies to every element.

Example in R:
```{r}
x <- c(2, 4, 6, 8)
mean(x)
```

Example in Python:
```python
#| eval: false
import numpy as np
x = np.array([2, 4, 6, 8])
print(x.mean())
```

### Lists

A list can hold values of different types and structures. In R, lists are flexible containers for numbers, strings, vectors, or even other lists. In Python, dictionaries and lists serve a similar purpose.

Use a list when the items are not all the same shape or type.

### Data frames and tables

A data frame is a table where each column is a variable and each row is an observation. This is the standard structure for statistical datasets.

In R, `data.frame` and `tibble` are common. In Python, `pandas.DataFrame` is the usual choice.

A data frame is ideal for:

- storing raw data,
- joining tables,
- grouping and summarising by variable,
- fitting models.

### Dictionaries and maps

A dictionary (Python) or named list (R) stores values by key. This is useful when you need to look up values by name rather than by position.

Example in Python:
```python
person = {'name': 'Alice', 'age': 24, 'score': 82}
print(person['score'])
```

Example in R:
```{r}
person <- list(name = 'Alice', age = 24, score = 82)
print(person$score)
```

---

## Algorithms in statistics

An algorithm is a step-by-step procedure for solving a problem. In statistics, algorithms manage tasks such as:

- searching and filtering data,
- summarising observations,
- sorting and ranking values,
- sampling and simulation,
- fitting models.

A good algorithm is:

- correct,
- simple enough to understand,
- efficient enough for the dataset you are using.

---

## A simple algorithm example

Suppose you want to count how many values in a dataset are greater than 100. The algorithm is:

1. start a counter at zero,
2. look at each value in the dataset,
3. if the value is greater than 100, add one to the counter,
4. return the counter.

In R:
```{r}
values <- c(70, 105, 98, 121, 115)
countOver100 <- 0
for (val in values) {
  if (val > 100) {
    countOver100 <- countOver100 + 1
  }
}
print(countOver100)
```

In Python:
```python
#| eval: false
values = [70, 105, 98, 121, 115]
count_over_100 = sum(1 for val in values if val > 100)
print(count_over_100)
```

Both versions follow the same algorithm, but the Python example uses a more compact expression.

---

## Algorithm cost: Big O at a glance

Big O notation describes how the runtime grows as the dataset gets larger. A few important ideas:

- **O(1)**: constant time. The cost does not depend on the dataset size.
- **O(n)**: linear time. The cost grows directly with the number of items.
- **O(n^2)**: quadratic time. The cost grows with the square of the number of items.

### Common operations

- Access one element in a vector: **O(1)**
- Scan all rows in a table: **O(n)**
- Nested loops over two lists: **O(n^2)**
- Sorting a list: typically **O(n log n)**

In statistics, many useful operations are O(n), such as computing means, sums, and simple counts. Nested loops can be expensive for large datasets.

---

## A statistical example: duplicates

Finding duplicates is a common operation. A simple nested loop is easy to write, but it can be slow for large datasets.

In R:
```{r}
findDuplicates <- function(x) {
  n <- length(x)
  res <- logical(n)
  for (i in seq_len(n - 1)) {
    for (j in (i + 1):n) {
      if (x[i] == x[j]) {
        res[i] <- TRUE
        res[j] <- TRUE
      }
    }
  }
  res
}

findDuplicates(c(1, 3, 3, 2, 1))
```

This approach is roughly **O(n^2)** because it compares every pair.

A more efficient strategy uses a hash-like lookup:

::: {.callout-note title="In a nutshell: a named list as a poor-man's hash map"}
There is no separate "dictionary" here — `seen` is just a named list being used as a lookup table. The **names** act as keys (the value we have already seen) and the stored entry is where we first saw it. The check `x[i] %in% names(seen)` is the lookup: "have I met this value before?" Because looking a name up is far cheaper than re-scanning the whole vector, we replace the nested loop with a single pass.
:::

```{r}
findDuplicatesFast <- function(x) {
  seen <- list()
  dups <- rep(FALSE, length(x))
  for (i in seq_along(x)) {
    if (x[i] %in% names(seen)) {
      dups[i] <- TRUE
      dups[seen[[as.character(x[i])]]] <- TRUE
    } else {
      seen[[as.character(x[i])]] <- i
    }
  }
  dups
}
findDuplicatesFast(c(1, 3, 3, 2, 1))
```

This version is closer to **O(n)** because it avoids repeated comparisons.

::: {.callout-tip title="Going further 🚀"}
A named list is a *convincing* hash map but not a true one — name lookups in a growing list are not guaranteed O(1). For real hashed keys, R gives you environments, which are backed by an actual hash table:

```{r}
#| eval: false
seen <- new.env(hash = TRUE)   # genuine hashed key-value store
assign("3", 1L, envir = seen)
exists("3", envir = seen)      # fast O(1) lookup
```

The `fastmap` package offers an even faster, purpose-built map (`fastmap::fastmap()`) for heavy use. But the real lesson is: base R already solves duplicate detection in compiled C — prefer the built-in.

```{r}
duplicated(c(1, 3, 3, 2, 1))   # vectorised, C-level, O(n) on average
```
:::

---

## Why this matters for modelling and machine learning

Data structures and algorithms are the foundation of statistical computing. When you fit a model, the software is using algorithms to:

- prepare features,
- compute coefficients,
- search for the best parameters,
- validate predictions.

Choosing the right structure makes every step faster and more reliable.

For example, a data frame is usually the best structure for a dataset, but a vector or array can be faster for numeric calculations once the dataset is prepared.

---

## Try it yourself

1. Load a vector of numbers in R.
2. Write a function that returns the number of values greater than the mean.
3. Change the function to return the number of values between the 25th and 75th percentiles.

This helps you practise the two key skills in this section: choosing a structure and writing a clear algorithm.
