---
title: "Efficiency in R"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

## Vectorisation in R

Vectorisation means applying operations to whole vectors or arrays at once, rather than iterating element by element. In R, many built-in functions are vectorised.

Vectorised code is often faster because it lets the underlying implementation work in compiled code and avoid the overhead of many separate loop iterations.

**In R we can measure time efficiency using the `system.time()` function.**

### Example

Using a loop:

```{r}
n <- 100000000
x <- rnorm(n)
y <- numeric(n)
for (i in seq_len(n)) {
  y[i] <- x[i] ^ 2
}
system.time({
  n <- 100000000
  x <- rnorm(n)
  y <- numeric(n)
  for (i in seq_len(n)) {
    y[i] <- x[i] ^ 2
  }
})
```

Using vectorisation:

```{r}
y <- x ^ 2
system.time({
  y <- x ^ 2
})
```

The vectorised version is shorter, easier to read, and typically much faster **for large data**. Remember that for small vectors, the difference may not be noticeable, but as the size of the data grows, the performance benefits of vectorisation become more pronounced. In this case we save approx. 8 seconds but with even larger datasets, the improvement becomes more significant.

## When to use vectorisation

Vectorisation is a good first choice when:

- the operation applies independently to each element,
- the same transformation is used across a whole vector,
- readability and speed are both important.

However, not every problem can be expressed cleanly in vectorised form. Complex algorithms or stateful loops may still require explicit iteration. The main benefit is that vectorised code is often more concise and can leverage optimized C or Fortran code under the hood, which can lead to significant performance improvements since it is compiled.

## Pre-allocate, don't grow

When a loop *is* genuinely needed, the single most common performance mistake in R is **growing an object one element at a time**. Each time you write `y <- c(y, value)`, R cannot extend the vector in place — it allocates a brand new, slightly larger vector and copies everything across. Doing that `n` times copies roughly `1 + 2 + ... + n` elements, which is **O(n^2)** work hidden inside an innocent-looking loop.

The fix is to **pre-allocate** the result to its final size once, then fill it by index:

```{r}
n <- 20000

# Slow: the vector is regrown on every iteration (O(n^2))
grow_vector <- function(n) {
  y <- numeric(0)
  for (i in seq_len(n)) {
    y <- c(y, i^2)
  }
  y
}

# Fast: allocate once, then assign into existing slots (O(n))
prealloc_vector <- function(n) {
  y <- numeric(n)
  for (i in seq_len(n)) {
    y[i] <- i^2
  }
  y
}

system.time(grow_vector(n))
system.time(prealloc_vector(n))
```

Even at only 20,000 elements the difference is dramatic, and the gap widens quadratically as `n` grows. The same warning applies to `rbind()`-ing rows onto a data frame inside a loop — build a list and combine once at the end instead.

::: {.callout-tip}
The hierarchy to remember: **vectorise if you can → if you must loop, pre-allocate → never grow inside a loop.**
:::

## The apply family

R provides a family of functions that apply an operation across the elements of a vector, list, or the rows/columns of a matrix without an explicit `for` loop. They are not magically faster than a *well-written* loop, but they are concise, avoid the grow-in-a-loop trap automatically, and make intent clear.

```{r}
# sapply: apply a function to each element and simplify to a vector
sapply(1:5, function(i) i^2)

# vapply: like sapply but you declare the return type — safer and slightly faster
vapply(1:5, function(i) i^2, numeric(1))

# lapply: returns a list (use when results are not all the same shape)
lapply(1:3, function(i) rnorm(i))
```

Use `vapply()` in production-style code: by stating the expected output type you catch errors early and avoid the surprises `sapply()` can produce when it cannot simplify the result.

## Prefer specialised built-ins

For common aggregations, R ships **dedicated compiled functions** that beat both loops and the apply family. If one exists for your task, use it:

```{r}
m <- matrix(rnorm(1e6), nrow = 1000)

# Manual apply across rows
system.time(apply(m, 1, sum))

# Specialised, fully compiled equivalent
system.time(rowSums(m))
```

`rowSums()`, `colMeans()`, `cumsum()`, `tabulate()`, and `which.max()` are all examples of purpose-built functions that run almost entirely in C. Reaching for the specialised tool is usually both faster and clearer than assembling the same result yourself.

## Benchmarking properly

`system.time()` is fine for operations that take a noticeable fraction of a second, but for fast operations it is too coarse and noisy. The `microbenchmark` package runs each expression many times and reports a distribution, which is far more reliable for comparing close alternatives:

```{r}
#| eval: false
library(microbenchmark)

x <- rnorm(1e5)

microbenchmark(
  loop      = { y <- numeric(length(x)); for (i in seq_along(x)) y[i] <- x[i]^2; y },
  vectorised = x^2,
  times = 100
)
```

::: {.callout-note}
Always benchmark on data sizes close to your real problem. Vectorisation and compiled code win decisively on large inputs, but for a handful of elements the overhead of setting them up can make a plain loop just as fast — efficiency only matters where the data is big enough to feel it.
:::

## R and compiled code

R is a high-level language, but much of its performance comes from compiled code under the hood. Many base functions and packages delegate work to C or C++.

This means the command you type in R may still run fast because it calls a compiled implementation.


## C++ code injection in R

When a calculation is very expensive, it can make sense to move the core work into compiled code.

Common approaches include:

- the `compiler` package for bytecode compilation,
- `Rcpp` for writing C++ code that interfaces with R,
- `sourceCpp()` to compile and load C++ functions directly.

These techniques are usually reserved for performance-critical kernels rather than everyday data manipulation.

### Example using Rcpp

```{r}
library(Rcpp)

cppFunction('NumericVector square_cpp(NumericVector x) {
  int n = x.size();
  NumericVector out(n);
  for (int i = 0; i < n; ++i) {
    out[i] = x[i] * x[i];
  }
  return out;
}')

n <- 100000000
x <- rnorm(n)
y <- square_cpp(x)

system.time({
  y <- square_cpp(x)
})
```

::: {.callout-warning}
This example requires a working C++ toolchain and the `Rcpp` package. If you do not have these installed, it will not work.
:::

::: {.callout-note}
Look at the fact that using Rcpp and vecorisation lead to almost the same speed. This is because Rcpp is compiled code, and vectorisation in R uses underlying C++ code since it is a very common operation but just hidden from you.
:::

This example compiles the loop into native code, which can reduce overhead compared with an interpreted R loop.

### Why this can help

- The loop runs in compiled code rather than in R's interpreter.
- Memory management is still handled by R, but the inner computation is faster.
- For large inputs or tight loops, compiled code can make a noticeable difference.

### When to use it

Use C/C++ integration when you have:

- a performance bottleneck in a tight loop,
- a calculation that cannot be expressed cleanly with vectorised R functions,
- a need for lower-level control over the computation.

If the operation can be expressed with existing vectorised R functions, that is usually the simpler and safer choice.

## Practical guidance

- Start with readable R code.
- Use vectorisation to avoid slow explicit loops.
- Use compiled code only when you have identified a real bottleneck.
- Prefer existing packages before writing your own C/C++ extension.

## Summary

Vectorisation and compiled code are two ways to make R faster. Vectorisation is the first tool to try, and C/C++ integration is a more advanced option for the heaviest computations.
