---
title: "Time and Space Efficiency"
format:
  html:
    fig-width: 12
    fig-height: 6
    code-fold: show
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    toc: true
    code-copy: true
    number_sections: true
    echo: fenced
---

## Why efficiency matters

Efficiency is about solving problems with a sensible amount of time and memory. In data science, performance matters because datasets can be large and analysis pipelines can be repeated many times.

A clear understanding of efficiency helps you choose the right approach before the code becomes too slow or too expensive in memory to use.

## Algorithmic complexity

Big O notation describes how an algorithm's running time grows as the size of the input increases. It captures the dominant cost for large inputs.

Common complexity classes:

- **O(1)** — constant time: the cost does not change when input size grows.
- **O(n)** — linear time: the cost grows in proportion to the number of items.
- **O(n log n)** — linearithmic time: common for efficient sorting.
- **O(n^2)** — quadratic time: often caused by nested loops.
- **O(2^n)** — exponential time: grows extremely fast with input size.

::: {.callout-note title="In a nutshell: what these classes feel like"}
The labels make more sense with a tangible task on `n` items:

| Class | A tangible example |
|-------|--------------------|
| O(1) | Asking one student their name. |
| O(n) | Reading every student's name once down the register. |
| O(n log n) | Sorting the register into alphabetical order. |
| O(n^2) | Comparing every student against every other student (every pair). |
| O(2^n) | Listing every possible group (subset) the students could form. |

The further down the table, the faster the cost explodes as the class grows.
:::

## Space efficiency

Space efficiency describes how much memory an algorithm or operation needs as the input grows.

Common space categories:

- **O(1)** — constant space: uses a fixed amount of memory regardless of input size.
- **O(n)** — linear space: memory grows in proportion to the input.
- **O(n^2)** — quadratic space: memory grows with the square of the input size.

Many data analysis operations are linear in space because they create vectors or tables that are proportional to the dataset.

## Example: time and space in R

### Loop vs vectorised squaring

A direct loop computes squared values one element at a time:

```{r}
n <- 1000000
x <- rnorm(n)
y <- numeric(n)
for (i in seq_len(n)) {
  y[i] <- x[i] ^ 2
}
```

This has:

- time complexity roughly **O(n)**,
- space complexity **O(n)** because the result vector `y` grows with the input size.

The vectorised version does the same work more compactly:

```{r}
y <- x ^ 2
```

It still uses **O(n)** time and **O(n)** space, but the work is performed inside a compiled routine and avoids the overhead of the explicit R loop.

### Time vs memory trade-off

If you only need the squared values and not the original vector, you can overwrite `x`:

```{r}
x <- x ^ 2
```

This still uses **O(n)** time, but it avoids keeping both `x` and `y` in memory at the same time. If you need both, then memory usage is higher.

## Example: quadratic time in R

A naive way to find duplicate pairs uses nested loops:

```{r}
findDuplicates <- function(x) {
  n <- length(x)
  res <- logical(n)
  for (i in seq_len(n - 1)) {
    for (j in (i + 1):n) {
      if (x[i] == x[j]) {
        res[i] <- TRUE
        res[j] <- TRUE
      }
    }
  }
  return(res)
}
```

This is roughly **O(n^2)** in time and also uses **O(n)** space for the output vector.

A good design balances time and space for the problem at hand.

## Recursion and the call stack

So far our examples have used **iteration** (loops) to repeat work. A second way to repeat work is **recursion**, where a function calls *itself* on a smaller version of the same problem. Every recursive function needs two parts:

- a **base case** that stops the recursion, and
- a **recursive case** that moves the problem closer to the base case.

A classic example is the factorial $n! = n \times (n-1) \times \dots \times 1$:

```{r}
factRec <- function(n) {
  if (n <= 1) {
    return(1)          # base case
  }
  n * factRec(n - 1)  # recursive case
}

factRec(5)
```

### The hidden space cost

Recursion has a space cost that loops do not. Each time a function calls itself, R must remember where it was so it can resume once the inner call returns. It does this by adding a **stack frame** to the *call stack*. The frames are only cleared once the base case is reached and the calls unwind.

This means a recursion that goes `n` levels deep uses **O(n) space on the call stack**, even if it returns a single number. The equivalent loop uses **O(1)** space:

```{r}
factIter <- function(n) {
  res <- 1
  for (i in seq_len(n)) {
    res <- res * i
  }
  res
}

factIter(5)
```

Both are **O(n)** in time, but the loop keeps memory flat while the recursion grows the stack.

::: {.callout-warning}
R does **not** optimise tail-recursion, and it caps recursion depth (see `options(expressions = ...)`). Very deep recursion will throw a *"evaluation nested too deeply"* / stack overflow error. For deep problems in R, prefer iteration or vectorisation.
:::

### When recursion is expensive in time, too

Recursion can also hide a severe *time* cost if the same sub-problems are recomputed. The naive Fibonacci sequence makes two recursive calls each step, so the work doubles at every level — **O(2^n)** time, the worst class from our table above:

```{r}
fibRec <- function(n) {
  if (n < 2) return(n)        # base cases: fib(0)=0, fib(1)=1
  fibRec(n - 1) + fibRec(n - 2)
}

system.time(fibRec(27))
```

The same sequence computed iteratively is **O(n)** time and **O(1)** space — astronomically faster:

```{r}
fibIter <- function(n) {
  a <- 0; b <- 1
  for (i in seq_len(n)) {
    new <- a + b
    a <- b
    b <- new
  }
  a
}

system.time(fibIter(27))
```

::: {.callout-note}
Recursion is not "bad" — it often expresses an algorithm far more clearly than a loop (the Euclidean algorithm and Newton-Raphson method in the *Mathematics to Code* section are natural recursions). The lesson is to be aware of its **two costs**: stack space proportional to depth, and repeated time if sub-problems overlap. When overlap happens, *memoisation* (caching results) or an iterative rewrite restores efficiency.
:::

::: {.callout-tip title="Going further 🚀"}
Memoisation caches each result the first time it is computed, so every sub-problem is solved once — collapsing `fibRec`'s **O(2^n)** time down to **O(n)** without rewriting the recursion. The `memoise` package wraps any function for you:

```{r}
#| eval: false
library(memoise)
fibMemo <- memoise(fibRec)   # cache lives in the wrapper
system.time(fibMemo(40))     # near-instant; naive fibRec(40) would hang
```

Under the hood it is just a lookup table. You can build a tiny one by hand with a base-R environment as the cache:

```{r}
cache <- new.env()
fibCached <- function(n) {
  key <- as.character(n)
  if (!is.null(cache[[key]])) return(cache[[key]])   # already known?
  val <- if (n < 2) n else fibCached(n - 1) + fibCached(n - 2)
  cache[[key]] <- val                                # store for next time
  val
}
fibCached(40)
```

:::

## Practical optimisation strategy

A sensible workflow is:

1. write correct code first,
2. check that it produces the right results,
3. measure performance,
4. optimise the real bottlenecks.

Optimising too early can waste effort on parts of the code that are not important.

## Why this matters in statistics

Statistical analysis often involves repeated tasks such as:

- scanning large datasets,
- computing summaries,
- fitting models multiple times,
- running simulations.

In these cases, choosing an efficient algorithm can make the difference between a script that finishes quickly and one that is too slow to use.

## Summary

Efficiency is a combination of algorithmic thinking and practical measurement. Big O notation helps you compare approaches, but real performance also depends on the specific language, data size, and implementation.